---
title: "Exploratory_Analysis_Project3"
output: html_document
---


## Section 1: Create Analytic Dataset
```{r}
library(tidyverse)

#load and inspect data
data <- read.csv("C:/Users/kokeshem/Spring 2026/Advanced Statistical Methods/Project 3/frmgham2.csv")

str(data)
table(Period = data$PERIOD)
n_distinct(data$RANDID)

#restrict to period 1, then check missingness on variables of interest
baseline <- data %>% filter(PERIOD == 1)

baseline %>%
  select(AGE, SYSBP, TOTCHOL, BMI, DIABETES, BPMEDS, CURSMOKE,
         PREVCHD, PREVSTRK, STROKE, TIMESTRK, DEATH, TIMEDTH) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N_Missing") %>%
  mutate(Pct_Missing = round(N_Missing / nrow(baseline) * 100, 2)) %>%
  arrange(desc(N_Missing))

#exclude prevalent stroke at baseline
analytic <- baseline %>% filter(PREVSTRK == 0)
#Number excluded due to prevalent stroke: 
nrow(baseline) - nrow(analytic)
#Rows remaining after prevalent stroke exclusion:
nrow(analytic)


#apply 10-year censoring (365.25 x 10 = 3652.5 days)
days_10yr <- 365.25 * 10

analytic <- analytic %>%
  mutate(
    STROKE_10yr   = ifelse(STROKE == 1 & TIMESTRK <= days_10yr, 1, 0),
    TIMESTRK_10yr = ifelse(TIMESTRK > days_10yr, days_10yr, TIMESTRK)
  )

#number of strokes after 10 years that were re-coded as censored
sum(analytic$STROKE == 1 & analytic$TIMESTRK > days_10yr)

#flag strokes at time = 0- check there are no prevalent stroke cases remaining after exclusion
strk_t0 <- analytic %>% filter(STROKE_10yr == 1 & TIMESTRK_10yr == 0)
nrow(strk_t0)
```

## Section 2: Outlier, Data Quality, and missingness Checks
```{r}
#summary statistics for continuous variables
analytic %>%
  select(AGE, SYSBP, TOTCHOL, BMI) %>%
  summary()

#outlier flags looped into one table- review for investigator to confirm if any values are biologically implausible and should be excluded
cont_checks <- list(
  AGE     = list(lo = 32,  hi = 81),
  SYSBP   = list(lo = 80,  hi = 250),
  TOTCHOL = list(lo = 100, hi = 600),
  BMI     = list(lo = 15,  hi = 80)
)

map_dfr(names(cont_checks), function(v) {
  tibble(
    Variable   = v,
    Low_bound  = cont_checks[[v]]$lo,
    N_below    = sum(analytic[[v]] < cont_checks[[v]]$lo, na.rm = TRUE),
    High_bound = cont_checks[[v]]$hi,
    N_above    = sum(analytic[[v]] > cont_checks[[v]]$hi, na.rm = TRUE)
  )
})

#binary variable checks — confirm only 0/1 values exist
binary_vars <- c("DIABETES", "BPMEDS", "CURSMOKE", "PREVCHD")

map_dfr(binary_vars, function(v) {
  tibble(Variable     = v,
         Unique_values = paste(sort(unique(analytic[[v]])), collapse = ", "))
})

#check missingness on analytic sample
analytic %>%
  select(AGE, SYSBP, TOTCHOL, BMI, DIABETES, BPMEDS, CURSMOKE,
         PREVCHD, STROKE_10yr, TIMESTRK_10yr, DEATH, TIMEDTH) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N_Missing") %>%
  mutate(Pct_Missing = round(N_Missing / nrow(analytic) * 100, 2)) %>%
  arrange(desc(N_Missing))

#check if missingness in BPMEDS and TOTCHOL is related to stroke outcome
#stroke rates compared between missing and non-missing groups
analytic %>%
  mutate(miss_BPMEDS = is.na(BPMEDS)) %>%
  group_by(miss_BPMEDS) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#fisher's exact test - BPMEDS missingness vs stroke outcome (small cell sizes)
fisher.test(table(is.na(analytic$BPMEDS), analytic$STROKE_10yr))

analytic %>%
  mutate(miss_TOTCHOL = is.na(TOTCHOL)) %>%
  group_by(miss_TOTCHOL) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#fisher's exact test - TOTCHOL missingness vs stroke outcome
fisher.test(table(is.na(analytic$TOTCHOL), analytic$STROKE_10yr))

analytic %>%
  mutate(miss_BMI = is.na(BMI)) %>%
  group_by(miss_BMI) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#summarize possible exclusions- biologically impossible values, data errors, and core missing
#reviewed with investigator: flagged extreme SYSBP and TOTCHOL values will be retained
cat("Flagged (SYSBP > 250):", sum(analytic$SYSBP > 250, na.rm = TRUE), "\n")
cat("Flagged (TOTCHOL > 600):", sum(analytic$TOTCHOL > 600, na.rm = TRUE), "\n")
cat("Remaining:", nrow(analytic), "\n")

```


## Section 3: Final Analytic Sample

```{r}

#core covariates (AGE, SYSBP, DIABETES) are complete — no further exclusions needed yet

#Exclusion summary
data.frame(
  Step = c(
    "Period 1 baseline",
    "Exclude prevalent stroke (PREVSTRK = 1)",
    "Final analytic sample"
  ),
  N_remaining = c(
    nrow(baseline),
    nrow(baseline %>% filter(PREVSTRK == 0)),
    nrow(analytic)
  ),
  N_excluded = c(NA, nrow(baseline) - nrow(baseline %>% filter(PREVSTRK == 0)), NA)
)

#final sample size and stroke counts by sex
analytic %>%
  mutate(Sex = ifelse(SEX == 1, "Men", "Women")) %>%
  group_by(Sex) %>%
  summarise(
    N            = n(),
    N_stroke     = sum(STROKE_10yr),
    Stroke_pct   = round(mean(STROKE_10yr) * 100, 2),
    Avg_followup = round(mean(TIMESTRK_10yr), 1)
  )

#cohort overview table for interim presentation
table_a <- tibble(
  item = c(
    "period 1 baseline observations",
    "excluded prevalent stroke",
    "flagged sysbp > 250 (retained)",
    "flagged totchol > 600 (retained)",
    "exploratory analysis sample",
    "men: n (10-year strokes, %)",
    "women: n (10-year strokes, %)"
  ),
  value = c(
    nrow(baseline),
    nrow(baseline) - nrow(baseline %>% filter(PREVSTRK == 0)),
    sum((baseline %>% filter(PREVSTRK == 0))$SYSBP > 250, na.rm = TRUE),
    sum((baseline %>% filter(PREVSTRK == 0))$TOTCHOL > 600, na.rm = TRUE),
    nrow(analytic),
    analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = paste0(n(), " (", sum(STROKE_10yr), ", ",
                             round(mean(STROKE_10yr) * 100, 2), "%)")) %>%
      pull(txt),
    analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = paste0(n(), " (", sum(STROKE_10yr), ", ",
                             round(mean(STROKE_10yr) * 100, 2), "%)")) %>%
      pull(txt)
  )
)

table_a

knitr::kable(table_a, col.names = c("Cohort Summary", "Value"))

#descriptive table by sex for presentation

cont_slide_table <- bind_rows(
  tibble(
    variable = "age",
    overall = analytic %>%
      summarise(txt = sprintf("%.1f (%.1f), %.0f-%.0f",
                              mean(AGE, na.rm = TRUE),
                              sd(AGE, na.rm = TRUE),
                              min(AGE, na.rm = TRUE),
                              max(AGE, na.rm = TRUE))) %>%
      pull(txt),
    men = analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = sprintf("%.1f (%.1f), %.0f-%.0f",
                              mean(AGE, na.rm = TRUE),
                              sd(AGE, na.rm = TRUE),
                              min(AGE, na.rm = TRUE),
                              max(AGE, na.rm = TRUE))) %>%
      pull(txt),
    women = analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = sprintf("%.1f (%.1f), %.0f-%.0f",
                              mean(AGE, na.rm = TRUE),
                              sd(AGE, na.rm = TRUE),
                              min(AGE, na.rm = TRUE),
                              max(AGE, na.rm = TRUE))) %>%
      pull(txt),
    missing = sum(is.na(analytic$AGE))
  ),
  tibble(
    variable = "systolic bp",
    overall = analytic %>%
      summarise(txt = sprintf("%.1f (%.1f), %.0f-%.0f",
                              mean(SYSBP, na.rm = TRUE),
                              sd(SYSBP, na.rm = TRUE),
                              min(SYSBP, na.rm = TRUE),
                              max(SYSBP, na.rm = TRUE))) %>%
      pull(txt),
    men = analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = sprintf("%.1f (%.1f), %.0f-%.0f",
                              mean(SYSBP, na.rm = TRUE),
                              sd(SYSBP, na.rm = TRUE),
                              min(SYSBP, na.rm = TRUE),
                              max(SYSBP, na.rm = TRUE))) %>%
      pull(txt),
    women = analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = sprintf("%.1f (%.1f), %.0f-%.0f",
                              mean(SYSBP, na.rm = TRUE),
                              sd(SYSBP, na.rm = TRUE),
                              min(SYSBP, na.rm = TRUE),
                              max(SYSBP, na.rm = TRUE))) %>%
      pull(txt),
    missing = sum(is.na(analytic$SYSBP))
  ),
  tibble(
    variable = "total cholesterol",
    overall = analytic %>%
      summarise(txt = sprintf("%.1f (%.1f), %.0f-%.0f",
                              mean(TOTCHOL, na.rm = TRUE),
                              sd(TOTCHOL, na.rm = TRUE),
                              min(TOTCHOL, na.rm = TRUE),
                              max(TOTCHOL, na.rm = TRUE))) %>%
      pull(txt),
    men = analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = sprintf("%.1f (%.1f), %.0f-%.0f",
                              mean(TOTCHOL, na.rm = TRUE),
                              sd(TOTCHOL, na.rm = TRUE),
                              min(TOTCHOL, na.rm = TRUE),
                              max(TOTCHOL, na.rm = TRUE))) %>%
      pull(txt),
    women = analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = sprintf("%.1f (%.1f), %.0f-%.0f",
                              mean(TOTCHOL, na.rm = TRUE),
                              sd(TOTCHOL, na.rm = TRUE),
                              min(TOTCHOL, na.rm = TRUE),
                              max(TOTCHOL, na.rm = TRUE))) %>%
      pull(txt),
    missing = sum(is.na(analytic$TOTCHOL))
  ),
  tibble(
    variable = "bmi",
    overall = analytic %>%
      summarise(txt = sprintf("%.1f (%.1f), %.1f-%.1f",
                              mean(BMI, na.rm = TRUE),
                              sd(BMI, na.rm = TRUE),
                              min(BMI, na.rm = TRUE),
                              max(BMI, na.rm = TRUE))) %>%
      pull(txt),
    men = analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = sprintf("%.1f (%.1f), %.1f-%.1f",
                              mean(BMI, na.rm = TRUE),
                              sd(BMI, na.rm = TRUE),
                              min(BMI, na.rm = TRUE),
                              max(BMI, na.rm = TRUE))) %>%
      pull(txt),
    women = analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = sprintf("%.1f (%.1f), %.1f-%.1f",
                              mean(BMI, na.rm = TRUE),
                              sd(BMI, na.rm = TRUE),
                              min(BMI, na.rm = TRUE),
                              max(BMI, na.rm = TRUE))) %>%
      pull(txt),
    missing = sum(is.na(analytic$BMI))
  )
)

bin_slide_table <- bind_rows(
  tibble(
    variable = "diabetes, yes n (%)",
    overall = analytic %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(DIABETES == 1, na.rm = TRUE),
                              mean(DIABETES == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    men = analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(DIABETES == 1, na.rm = TRUE),
                              mean(DIABETES == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    women = analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(DIABETES == 1, na.rm = TRUE),
                              mean(DIABETES == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    missing = sum(is.na(analytic$DIABETES))
  ),
  tibble(
    variable = "current smoking, yes n (%)",
    overall = analytic %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(CURSMOKE == 1, na.rm = TRUE),
                              mean(CURSMOKE == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    men = analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(CURSMOKE == 1, na.rm = TRUE),
                              mean(CURSMOKE == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    women = analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(CURSMOKE == 1, na.rm = TRUE),
                              mean(CURSMOKE == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    missing = sum(is.na(analytic$CURSMOKE))
  ),
  tibble(
    variable = "prevalent chd, yes n (%)",
    overall = analytic %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(PREVCHD == 1, na.rm = TRUE),
                              mean(PREVCHD == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    men = analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(PREVCHD == 1, na.rm = TRUE),
                              mean(PREVCHD == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    women = analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(PREVCHD == 1, na.rm = TRUE),
                              mean(PREVCHD == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    missing = sum(is.na(analytic$PREVCHD))
  ),
  tibble(
    variable = "bp meds, yes n (%)",
    overall = analytic %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(BPMEDS == 1, na.rm = TRUE),
                              mean(BPMEDS == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    men = analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(BPMEDS == 1, na.rm = TRUE),
                              mean(BPMEDS == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    women = analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = sprintf("%d (%.1f%%)",
                              sum(BPMEDS == 1, na.rm = TRUE),
                              mean(BPMEDS == 1, na.rm = TRUE) * 100)) %>%
      pull(txt),
    missing = sum(is.na(analytic$BPMEDS))
  )
)

slide_table <- bind_rows(cont_slide_table, bin_slide_table)

knitr::kable(
  slide_table,
  col.names = c("Variable", "Overall", "Men", "Women", "Missing")
)


# make table 1 style descriptive summary by sex

library(gtsummary)
library(dplyr)

table1_data <- analytic %>%
  mutate(
    sex = factor(SEX, levels = c(1, 2), labels = c("Men", "Women")),
    diabetes = factor(DIABETES, levels = c(0, 1), labels = c("No", "Yes")),
    current_smoker = factor(CURSMOKE, levels = c(0, 1), labels = c("No", "Yes")),
    antihyp_med = factor(BPMEDS, levels = c(0, 1), labels = c("No", "Yes")),
    prev_chd = factor(PREVCHD, levels = c(0, 1), labels = c("No", "Yes")),
    stroke_10yr = factor(STROKE_10yr, levels = c(0, 1), labels = c("No", "Yes"))
  ) %>%
  select(
    sex,
    AGE,
    SYSBP,
    TOTCHOL,
    BMI,
    diabetes,
    current_smoker,
    antihyp_med,
    prev_chd,
    stroke_10yr
  )

table1 <- table1_data %>%
  tbl_summary(
    by = sex,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_dichotomous() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 1,
      all_dichotomous() ~ c(0, 1)
    ),
    type = list(
      diabetes ~ "dichotomous",
      current_smoker ~ "dichotomous",
      antihyp_med ~ "dichotomous",
      prev_chd ~ "dichotomous",
      stroke_10yr ~ "dichotomous"
    ),
    value = list(
      diabetes ~ "Yes",
      current_smoker ~ "Yes",
      antihyp_med ~ "Yes",
      prev_chd ~ "Yes",
      stroke_10yr ~ "Yes"
    ),
    label = list(
      AGE ~ "Age (years)",
      SYSBP ~ "Systolic BP (mmHg)",
      TOTCHOL ~ "Total cholesterol (mg/dL)",
      BMI ~ "BMI (kg/m^2)",
      diabetes ~ "Diabetes (yes)",
      current_smoker ~ "Current smoker",
      antihyp_med ~ "Blood Pressure Mdication (yes)",
      prev_chd ~ "Prevalent CHD (yes)",
      stroke_10yr ~ "10-year stroke event"
    ),
    missing = "no"
  ) %>%
  add_overall(last = FALSE) %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

table1



```

# NOTE: TOTCHOL (52 missing), BMI (17 missing), and BPMEDS (60 missing) are
# retained in the analytic dataset at this stage. If any of these variables are
# included in the final model, additional participants will be excluded and
# the exclusion table above will be updated accordingly.

## Section 4: #Change Over Time in SYSBP and DIABETES (Descriptive Only), correlation checks Univariate and Bivariate Distributions


```{r}
library(ggplot2)

#summarize change over time in sysbp and diabetes

change_over_time <- data %>%
  mutate(
    sex = ifelse(SEX == 1, "men", "women"),
    period = factor(PERIOD, levels = c(1, 2, 3))
  ) %>%
  group_by(sex, period) %>%
  summarise(
    n_obs = n(),
    sysbp_mean = round(mean(SYSBP, na.rm = TRUE), 1),
    sysbp_sd = round(sd(SYSBP, na.rm = TRUE), 1),
    diabetes_n = sum(DIABETES == 1, na.rm = TRUE),
    diabetes_pct = round(mean(DIABETES == 1, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

change_over_time

change_over_time %>%
  ggplot(aes(x = period, y = sysbp_mean, group = sex, linetype = sex, shape = sex)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "mean systolic blood pressure across examination periods",
    x = "examination period",
    y = "mean sysbp (mmhg)"
  ) +
  theme_minimal()

change_over_time %>%
  ggplot(aes(x = period, y = diabetes_pct, group = sex, linetype = sex, shape = sex)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "diabetes prevalence across examination periods",
    x = "examination period",
    y = "diabetes prevalence (%)"
  ) +
  theme_minimal()

#define variable sets used below

cont_vars <- c("AGE", "SYSBP", "TOTCHOL", "BMI")
bin_vars <- c("CURSMOKE", "DIABETES", "BPMEDS", "PREVCHD")

#check sysbp and bpmeds relationship

sysbp_bpmeds_summary <- analytic %>%
  mutate(
    sex = ifelse(SEX == 1, "men", "women"),
    bpmeds_cat = case_when(
      BPMEDS == 0 ~ "no",
      BPMEDS == 1 ~ "yes",
      TRUE ~ "missing"
    ),
    bpmeds_cat = factor(bpmeds_cat, levels = c("no", "yes", "missing"))
  ) %>%
  group_by(sex, bpmeds_cat) %>%
  summarise(
    n = n(),
    sysbp_mean = round(mean(SYSBP, na.rm = TRUE), 1),
    sysbp_sd = round(sd(SYSBP, na.rm = TRUE), 1),
    .groups = "drop"
  )

sysbp_bpmeds_summary

sysbp_bpmeds_corr <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex) %>%
  summarise(
    corr_sysbp_bpmeds = round(cor(SYSBP, BPMEDS, use = "complete.obs"), 3),
    .groups = "drop"
  )

sysbp_bpmeds_corr

sysbp_bpmeds_test <- analytic %>%
  filter(!is.na(BPMEDS)) %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex) %>%
  summarise(
    p_value = wilcox.test(SYSBP ~ BPMEDS)$p.value,
    .groups = "drop"
  )

sysbp_bpmeds_test

analytic %>%
  mutate(
    sex = ifelse(SEX == 1, "men", "women"),
    bpmeds_cat = case_when(
      BPMEDS == 0 ~ "no",
      BPMEDS == 1 ~ "yes",
      TRUE ~ "missing"
    ),
    bpmeds_cat = factor(bpmeds_cat, levels = c("no", "yes", "missing"))
  ) %>%
  ggplot(aes(x = bpmeds_cat, y = SYSBP, fill = sex)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~ sex) +
  labs(
    title = "systolic blood pressure by bp medication status",
    x = "on bp medication",
    y = "sysbp (mmhg)",
    fill = "sex"
  ) +
  theme_minimal()

#check correlations among continuous predictors

cont_corr <- analytic %>%
  select(all_of(cont_vars)) %>%
  cor(use = "pairwise.complete.obs")

round(cont_corr, 3)

#check associations among binary candidate predictors

binary_assoc <- expand_grid(var1 = bin_vars, var2 = bin_vars) %>%
  filter(var1 < var2) %>%
  rowwise() %>%
  mutate(
    n_complete = sum(!is.na(analytic[[var1]]) & !is.na(analytic[[var2]])),
    p_value = {
      keep <- !is.na(analytic[[var1]]) & !is.na(analytic[[var2]])
      tbl <- table(analytic[[var1]][keep], analytic[[var2]][keep])
      if (any(tbl < 5)) fisher.test(tbl)$p.value else chisq.test(tbl)$p.value
    },
    test_used = {
      keep <- !is.na(analytic[[var1]]) & !is.na(analytic[[var2]])
      tbl <- table(analytic[[var1]][keep], analytic[[var2]][keep])
      if (any(tbl < 5)) "fisher" else "chi-square"
    }
  ) %>%
  ungroup()

binary_assoc

#look at univariate distributions

analytic %>%
  select(all_of(cont_vars), STROKE_10yr) %>%
  pivot_longer(all_of(cont_vars), names_to = "variable", values_to = "value") %>%
  mutate(stroke = ifelse(STROKE_10yr == 1, "stroke", "no stroke")) %>%
  ggplot(aes(x = value, fill = stroke)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  facet_wrap(~ variable, scales = "free") +
  labs(
    title = "distribution of continuous variables by stroke status",
    x = NULL,
    y = "count",
    fill = "stroke status"
  ) +
  theme_minimal()

analytic %>%
  select(all_of(cont_vars), STROKE_10yr, SEX) %>%
  pivot_longer(all_of(cont_vars), names_to = "variable", values_to = "value") %>%
  mutate(
    stroke = ifelse(STROKE_10yr == 1, "stroke", "no stroke"),
    sex = ifelse(SEX == 1, "men", "women")
  ) %>%
  ggplot(aes(x = stroke, y = value, fill = sex)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "boxplots of continuous variables by stroke status and sex",
    x = NULL,
    y = NULL,
    fill = "sex"
  ) +
  theme_minimal()

analytic %>%
  select(all_of(bin_vars), STROKE_10yr, SEX) %>%
  pivot_longer(all_of(bin_vars), names_to = "variable", values_to = "value") %>%
  mutate(
    stroke = ifelse(STROKE_10yr == 1, "stroke", "no stroke"),
    sex = ifelse(SEX == 1, "men", "women"),
    value = case_when(
      value == 1 ~ "yes",
      value == 0 ~ "no",
      TRUE ~ "missing"
    )
  ) %>%
  ggplot(aes(x = value, fill = stroke)) +
  geom_bar(position = "dodge", alpha = 0.8) +
  facet_grid(sex ~ variable) +
  labs(
    title = "binary variables by stroke status and sex",
    x = NULL,
    y = "count",
    fill = "stroke status"
  ) +
  theme_minimal()

#do bivariate analysis by sex for continuous variables

cont_results <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex) %>%
  reframe(
    tibble(
      variable = cont_vars,
      stroke_mean = c(
        round(mean(AGE[STROKE_10yr == 1], na.rm = TRUE), 1),
        round(mean(SYSBP[STROKE_10yr == 1], na.rm = TRUE), 1),
        round(mean(TOTCHOL[STROKE_10yr == 1], na.rm = TRUE), 1),
        round(mean(BMI[STROKE_10yr == 1], na.rm = TRUE), 1)
      ),
      no_stroke_mean = c(
        round(mean(AGE[STROKE_10yr == 0], na.rm = TRUE), 1),
        round(mean(SYSBP[STROKE_10yr == 0], na.rm = TRUE), 1),
        round(mean(TOTCHOL[STROKE_10yr == 0], na.rm = TRUE), 1),
        round(mean(BMI[STROKE_10yr == 0], na.rm = TRUE), 1)
      ),
      p_value = c(
        round(wilcox.test(AGE ~ STROKE_10yr)$p.value, 4),
        round(wilcox.test(SYSBP ~ STROKE_10yr)$p.value, 4),
        round(wilcox.test(TOTCHOL ~ STROKE_10yr)$p.value, 4),
        round(wilcox.test(BMI ~ STROKE_10yr)$p.value, 4)
      )
    )
  )

cont_results

#do bivariate analysis by sex for binary variables
bin_results <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_split(sex) %>%
  map_dfr(function(df_sex) {
    
    sex_label <- unique(df_sex$sex)
    
    map_dfr(bin_vars, function(v) {
      df_use <- df_sex[!is.na(df_sex[[v]]), ]
      tbl <- table(df_use[[v]], df_use$STROKE_10yr)
      
      test <- if (any(tbl < 5)) {
        fisher.test(tbl)
      } else {
        suppressWarnings(chisq.test(tbl))
      }
      
      stroke_pct_no <- df_use %>%
        filter(.data[[v]] == 0) %>%
        summarise(stroke_pct = round(mean(STROKE_10yr) * 100, 1)) %>%
        pull(stroke_pct)
      
      stroke_pct_yes <- df_use %>%
        filter(.data[[v]] == 1) %>%
        summarise(stroke_pct = round(mean(STROKE_10yr) * 100, 1)) %>%
        pull(stroke_pct)
      
      tibble(
        sex = sex_label,
        variable = v,
        stroke_pct_no = stroke_pct_no,
        stroke_pct_yes = stroke_pct_yes,
        p_value = round(test$p.value, 4),
        test_used = if (any(tbl < 5)) "fisher" else "chi-square"
      )
    })
  })

bin_results
```
