---
title: "Exploratory_Analysis_Project3"
output: html_document
---


## Section 1: Create Analytic Dataset
```{r}
library(tidyverse)

#load and inspect data
data <- read.csv("C:/Users/kokeshem/Spring 2026/Advanced Statistical Methods/Project 3/frmgham2.csv")

str(data)
table(Period = data$PERIOD)
n_distinct(data$RANDID)

#restrict to period 1, then check missingness on variables of interest
baseline <- data %>% filter(PERIOD == 1)

baseline %>%
  select(AGE, SYSBP, TOTCHOL, BMI, DIABETES, BPMEDS, CURSMOKE,
         PREVCHD, PREVSTRK, STROKE, TIMESTRK, DEATH, TIMEDTH) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N_Missing") %>%
  mutate(Pct_Missing = round(N_Missing / nrow(baseline) * 100, 2)) %>%
  arrange(desc(N_Missing))

#exclude prevalent stroke at baseline
analytic <- baseline %>% filter(PREVSTRK == 0)
#Number excluded due to prevalent stroke: 
nrow(baseline) - nrow(analytic)
#Rows remaining after prevalent stroke exclusion:
nrow(analytic)


#apply 10-year censoring (365.25 x 10 = 3652.5 days)
days_10yr <- 365.25 * 10

analytic <- analytic %>%
  mutate(
    STROKE_10yr   = ifelse(STROKE == 1 & TIMESTRK <= days_10yr, 1, 0),
    TIMESTRK_10yr = ifelse(TIMESTRK > days_10yr, days_10yr, TIMESTRK)
  )

#number of strokes after 10 years that were re-coded as censored
sum(analytic$STROKE == 1 & analytic$TIMESTRK > days_10yr)

#flag strokes at time = 0- check there are no prevalent stroke cases remaining after exclusion
strk_t0 <- analytic %>% filter(STROKE_10yr == 1 & TIMESTRK_10yr == 0)
nrow(strk_t0)
```

## Section 2: Outlier, Data Quality, and missingness Checks
```{r}
#summary statistics for continuous variables
analytic %>%
  select(AGE, SYSBP, TOTCHOL, BMI) %>%
  summary()

#outlier flags looped into one table- review for investigator to confirm if any values are biologically implausible and should be excluded
cont_checks <- list(
  AGE     = list(lo = 32,  hi = 81),
  SYSBP   = list(lo = 80,  hi = 250),
  TOTCHOL = list(lo = 100, hi = 600),
  BMI     = list(lo = 15,  hi = 80)
)

map_dfr(names(cont_checks), function(v) {
  tibble(
    Variable   = v,
    Low_bound  = cont_checks[[v]]$lo,
    N_below    = sum(analytic[[v]] < cont_checks[[v]]$lo, na.rm = TRUE),
    High_bound = cont_checks[[v]]$hi,
    N_above    = sum(analytic[[v]] > cont_checks[[v]]$hi, na.rm = TRUE)
  )
})

#binary variable checks — confirm only 0/1 values exist
binary_vars <- c("DIABETES", "BPMEDS", "CURSMOKE", "PREVCHD")

map_dfr(binary_vars, function(v) {
  tibble(Variable     = v,
         Unique_values = paste(sort(unique(analytic[[v]])), collapse = ", "))
})

#check missingness on analytic sample
analytic %>%
  select(AGE, SYSBP, TOTCHOL, BMI, DIABETES, BPMEDS, CURSMOKE,
         PREVCHD, STROKE_10yr, TIMESTRK_10yr, DEATH, TIMEDTH) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N_Missing") %>%
  mutate(Pct_Missing = round(N_Missing / nrow(analytic) * 100, 2)) %>%
  arrange(desc(N_Missing))

#check if missingness in BPMEDS and TOTCHOL is related to stroke outcome
#stroke rates compared between missing and non-missing groups
analytic %>%
  mutate(miss_BPMEDS = is.na(BPMEDS)) %>%
  group_by(miss_BPMEDS) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#fisher's exact test - BPMEDS missingness vs stroke outcome (small cell sizes)
fisher.test(table(is.na(analytic$BPMEDS), analytic$STROKE_10yr))

analytic %>%
  mutate(miss_TOTCHOL = is.na(TOTCHOL)) %>%
  group_by(miss_TOTCHOL) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#fisher's exact test - TOTCHOL missingness vs stroke outcome
fisher.test(table(is.na(analytic$TOTCHOL), analytic$STROKE_10yr))

analytic %>%
  mutate(miss_BMI = is.na(BMI)) %>%
  group_by(miss_BMI) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#summarize possible exclusions- biologically impossible values, data errors, and core missing
n_before <- nrow(analytic)

analytic <- analytic %>%
  filter(SYSBP <= 250)

#Exclude SYSBP > 250?
cat("Excluded (SYSBP > 250):", n_before - nrow(analytic), "\n")
cat("Remaining:", nrow(analytic), "\n")

n_before <- nrow(analytic)

analytic <- analytic %>%
  filter(is.na(TOTCHOL) | TOTCHOL <= 600)

#Exclude TOTCHOL > 600?
cat("Excluded (TOTCHOL > 600):", n_before - nrow(analytic), "\n")
cat("Remaining:", nrow(analytic), "\n")


```


## Section 3: Final Analytic Sample

```{r}

#core covariates (AGE, SYSBP, DIABETES) are complete — no further exclusions needed yet

#Exclusion summary
data.frame(
  Step = c(
    "Period 1 baseline",
    "Exclude prevalent stroke (PREVSTRK = 1)",
    "Exclude implausible SYSBP (> 250)",
    "Exclude data entry error TOTCHOL (> 600)",
    "Final analytic sample"
  ),
  N_remaining = c(
    nrow(baseline),
    nrow(baseline %>% filter(PREVSTRK == 0)),
    nrow(baseline %>% filter(PREVSTRK == 0)) - 1,
    nrow(baseline %>% filter(PREVSTRK == 0)) - 2,
    nrow(analytic)
  ),
  N_excluded = c(NA, 32, 1, 1, NA)
)

#final sample size and stroke counts by sex
analytic %>%
  mutate(Sex = ifelse(SEX == 1, "Men", "Women")) %>%
  group_by(Sex) %>%
  summarise(
    N            = n(),
    N_stroke     = sum(STROKE_10yr),
    Stroke_pct   = round(mean(STROKE_10yr) * 100, 2),
    Avg_followup = round(mean(TIMESTRK_10yr), 1)
  )


```

# NOTE: TOTCHOL (52 missing), BMI (17 missing), and BPMEDS (60 missing) are
# retained in the analytic dataset at this stage. If any of these variables are
# included in the final model, additional participants will be excluded and
# the exclusion table above will be updated accordingly.

## Section 4: Univariate Distributions

```{r}
library(ggplot2)


#continuous variables- histograms and boxplots stratified by stroke status and sex

cont_vars <- c("AGE", "SYSBP", "TOTCHOL", "BMI")

#histograms by stroke status
analytic %>%
  select(all_of(cont_vars), STROKE_10yr) %>%
  pivot_longer(all_of(cont_vars), names_to = "Variable", values_to = "Value") %>%
  mutate(Stroke = ifelse(STROKE_10yr == 1, "Stroke", "No Stroke")) %>%
  ggplot(aes(x = Value, fill = Stroke)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  facet_wrap(~ Variable, scales = "free") +
  labs(title = "Distribution of Continuous Variables by Stroke Status",
       x = NULL, y = "Count", fill = "Stroke Status") +
  theme_minimal()

#boxplots by sex and stroke status
analytic %>%
  select(all_of(cont_vars), STROKE_10yr, SEX) %>%
  pivot_longer(all_of(cont_vars), names_to = "Variable", values_to = "Value") %>%
  mutate(Stroke = ifelse(STROKE_10yr == 1, "Stroke", "No Stroke"),
         Sex    = ifelse(SEX == 1, "Men", "Women")) %>%
  ggplot(aes(x = Stroke, y = Value, fill = Sex)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~ Variable, scales = "free_y") +
  labs(title = "Boxplots of Continuous Variables by Stroke Status and Sex",
       x = NULL, y = NULL, fill = "Sex") +
  theme_minimal()

#binary variables: bar charts by sex and stroke status

bin_vars <- c("CURSMOKE", "DIABETES", "BPMEDS", "PREVCHD")

analytic %>%
  select(all_of(bin_vars), STROKE_10yr, SEX) %>%
  pivot_longer(all_of(bin_vars), names_to = "Variable", values_to = "Value") %>%
  mutate(Stroke = ifelse(STROKE_10yr == 1, "Stroke", "No Stroke"),
         Sex    = ifelse(SEX == 1, "Men", "Women"),
         Value  = ifelse(Value == 1, "Yes", "No")) %>%
  ggplot(aes(x = Value, fill = Stroke)) +
  geom_bar(position = "dodge", alpha = 0.8) +
  facet_grid(Sex ~ Variable) +
  labs(title = "Binary Variables by Stroke Status and Sex",
       x = NULL, y = "Count", fill = "Stroke Status") +
  theme_minimal()
```



