---
title: "Exploratory_Analysis_Project3"
output: html_document
---


## Section 1: Baseline Cohort Restriction and 10-Year Stroke Outcome Definition
```{r}
library(tidyverse)

#load and inspect data
data <- read.csv("C:/Users/kokeshem/Spring 2026/Advanced Statistical Methods/Project 3/frmgham2.csv")

str(data)
table(Period = data$PERIOD)
n_distinct(data$RANDID)

#restrict to period 1, then check missingness on variables of interest
baseline <- data %>% filter(PERIOD == 1)

baseline %>%
  select(AGE, SYSBP, TOTCHOL, BMI, DIABETES, BPMEDS, CURSMOKE,
         PREVCHD, PREVSTRK, STROKE, TIMESTRK, DEATH, TIMEDTH) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N_Missing") %>%
  mutate(Pct_Missing = round(N_Missing / nrow(baseline) * 100, 2)) %>%
  arrange(desc(N_Missing))

#exclude prevalent stroke at baseline
analytic <- baseline %>% filter(PREVSTRK == 0)
#Number excluded due to prevalent stroke: 
nrow(baseline) - nrow(analytic)
#Rows remaining after prevalent stroke exclusion:
nrow(analytic)

#apply 10-year censoring (365.25 x 10 = 3652.5 days)
days_10yr <- 365.25 * 10

analytic <- analytic %>%
  mutate(
    #event indicator for stroke occurring within 10 years
    STROKE_10yr = ifelse(STROKE == 1 & TIMESTRK <= days_10yr, 1, 0),

    #observed followup time for 10-year survival analysis:
    #use TIMESTRK directly for the stroke endpoint and administratively censor at 10 years
    FU_10yr = pmin(TIMESTRK, days_10yr),

    #retained for descriptive summaries of stroke time truncated at 10 years
    TIMESTRK_10yr = ifelse(TIMESTRK > days_10yr, days_10yr, TIMESTRK)
  )

#number of strokes after 10 years that were re-coded as censored
sum(analytic$STROKE == 1 & analytic$TIMESTRK > days_10yr)

#check that no strokes should be recorded at day 0 after excluding PREVSTRK == 1
#use TIMESTRK (the original event time) rather than FU_10yr so the check is independent of 10-year truncation
strk_t0 <- analytic %>% filter(STROKE == 1 & TIMESTRK == 0)
nrow(strk_t0)
```

#Section 2: Data Quality and Missingness Checks
```{r}
#summary statistics for continuous variables
analytic %>%
  select(AGE, SYSBP, TOTCHOL, BMI) %>%
  summary()

#outlier flags looped into one table- review for investigator to confirm if any values are biologically implausible and should be excluded
cont_checks <- list(
  AGE     = list(lo = 32,  hi = 81),
  SYSBP   = list(lo = 80,  hi = 250),
  TOTCHOL = list(lo = 100, hi = 600),
  BMI     = list(lo = 15,  hi = 80)
)

map_dfr(names(cont_checks), function(v) {
  tibble(
    Variable   = v,
    Low_bound  = cont_checks[[v]]$lo,
    N_below    = sum(analytic[[v]] < cont_checks[[v]]$lo, na.rm = TRUE),
    High_bound = cont_checks[[v]]$hi,
    N_above    = sum(analytic[[v]] > cont_checks[[v]]$hi, na.rm = TRUE)
  )
})

#binary variable checks — confirm only 0/1 values exist
binary_vars <- c("DIABETES", "BPMEDS", "CURSMOKE", "PREVCHD")

map_dfr(binary_vars, function(v) {
  tibble(
    Variable = v,
    Unique_values = paste(sort(unique(analytic[[v]])), collapse = ", ")
  )
})

#check missingness on analytic sample
analytic %>%
  select(AGE, SYSBP, TOTCHOL, BMI, DIABETES, BPMEDS, CURSMOKE,
         PREVCHD, STROKE_10yr, TIMESTRK_10yr, DEATH, TIMEDTH) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N_Missing") %>%
  mutate(Pct_Missing = round(N_Missing / nrow(analytic) * 100, 2)) %>%
  arrange(desc(N_Missing))

#check if missingness in BPMEDS and TOTCHOL is related to stroke outcome
#stroke rates compared between missing and non-missing groups
analytic %>%
  mutate(miss_BPMEDS = is.na(BPMEDS)) %>%
  group_by(miss_BPMEDS) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#fisher's exact test - BPMEDS missingness vs stroke outcome (small cell sizes)
fisher.test(table(is.na(analytic$BPMEDS), analytic$STROKE_10yr))

analytic %>%
  mutate(miss_TOTCHOL = is.na(TOTCHOL)) %>%
  group_by(miss_TOTCHOL) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#fisher's exact test - TOTCHOL missingness vs stroke outcome
fisher.test(table(is.na(analytic$TOTCHOL), analytic$STROKE_10yr))

analytic %>%
  mutate(miss_BMI = is.na(BMI)) %>%
  group_by(miss_BMI) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#summarize possible exclusions- biologically impossible values, data errors, and core missing
#reviewed with investigator: flagged extreme SYSBP and TOTCHOL values will be retained
cat("Flagged (SYSBP > 250):", sum(analytic$SYSBP > 250, na.rm = TRUE), "\n")
cat("Flagged (TOTCHOL > 600):", sum(analytic$TOTCHOL > 600, na.rm = TRUE), "\n")
cat("Remaining:", nrow(analytic), "\n")
```

#Section 3: Cohort Summary and Descriptive Characteristics
```{r}
#core covariates (AGE, SYSBP, DIABETES) are complete — no further exclusions needed yet

#Exclusion summary
data.frame(
  Step = c(
    "Period 1 baseline",
    "Exclude prevalent stroke (PREVSTRK = 1)",
    "Analytic cohort for exploratory and survival analyses"
  ),
  N_remaining = c(
    nrow(baseline),
    nrow(baseline %>% filter(PREVSTRK == 0)),
    nrow(analytic)
  ),
  N_excluded = c(
    NA,
    nrow(baseline) - nrow(baseline %>% filter(PREVSTRK == 0)),
    NA
  )
)

#cohort overview table for interim presentation
table_a <- tibble(
  item = c(
    "period 1 baseline observations",
    "excluded prevalent stroke",
    "flagged sysbp > 250 (retained)",
    "flagged totchol > 600 (retained)",
    "exploratory analysis sample",
    "men: n (10-year strokes, %)",
    "women: n (10-year strokes, %)"
  ),
  value = c(
    nrow(baseline),
    nrow(baseline) - nrow(baseline %>% filter(PREVSTRK == 0)),
    sum((baseline %>% filter(PREVSTRK == 0))$SYSBP > 250, na.rm = TRUE),
    sum((baseline %>% filter(PREVSTRK == 0))$TOTCHOL > 600, na.rm = TRUE),
    nrow(analytic),
    analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = paste0(n(), " (", sum(STROKE_10yr), ", ",
                             round(mean(STROKE_10yr) * 100, 2), "%)")) %>%
      pull(txt),
    analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = paste0(n(), " (", sum(STROKE_10yr), ", ",
                             round(mean(STROKE_10yr) * 100, 2), "%)")) %>%
      pull(txt)
  )
)

table_a
knitr::kable(table_a, col.names = c("Cohort Summary", "Value"))


#cohort flow diagram

library(DiagrammeR)
library(glue)

n_period1 <- nrow(baseline)
n_prevstroke_excluded <- sum(baseline$PREVSTRK == 1, na.rm = TRUE)
n_analytic <- nrow(analytic)

n_men <- nrow(analytic_men)
n_women <- nrow(analytic_women)

n_lasso_men <- nrow(cox_men_data)
n_lasso_women <- nrow(cox_women_data)

n_refit_men <- nrow(refit_men_data)
n_refit_women <- nrow(refit_women_data)

grViz(glue("
digraph cohort_flow {{

  graph [layout = dot, rankdir = TB]
  node [shape = box, style = rounded, fontname = Helvetica, fontsize = 11]
  edge [fontname = Helvetica, fontsize = 10]

  A [label = 'Period 1 baseline sample\\nN = {n_period1}']
  B [label = 'Excluded: prevalent stroke\\nN = {n_prevstroke_excluded}']
  C [label = 'Baseline-eligible analytic cohort\\nN = {n_analytic}']

  D [label = 'Men\\nAnalytic sample: N = {n_men}\\nLASSO sample: N = {n_lasso_men}\\nFinal refit sample: N = {n_refit_men}']
  E [label = 'Women\\nAnalytic sample: N = {n_women}\\nLASSO sample: N = {n_lasso_women}\\nFinal refit sample: N = {n_refit_women}']

  A -> C
  A -> B [style = dashed]
  C -> D
  C -> E
}}
"))

# make table 1 style descriptive summary by sex

library(gtsummary)
library(dplyr)

table1_data <- analytic %>%
  mutate(
    sex = factor(SEX, levels = c(1, 2), labels = c("Men", "Women")),
    diabetes = factor(DIABETES, levels = c(0, 1), labels = c("No", "Yes")),
    current_smoker = factor(CURSMOKE, levels = c(0, 1), labels = c("No", "Yes")),
    antihyp_med = factor(BPMEDS, levels = c(0, 1), labels = c("No", "Yes")),
    prev_chd = factor(PREVCHD, levels = c(0, 1), labels = c("No", "Yes")),
    stroke_10yr = factor(STROKE_10yr, levels = c(0, 1), labels = c("No", "Yes"))
  ) %>%
  select(
    sex,
    AGE,
    SYSBP,
    TOTCHOL,
    BMI,
    diabetes,
    current_smoker,
    antihyp_med,
    prev_chd,
    stroke_10yr
  )

table1 <- table1_data %>%
  tbl_summary(
    by = sex,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_dichotomous() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 1,
      all_dichotomous() ~ c(0, 1)
    ),
    type = list(
      diabetes ~ "dichotomous",
      current_smoker ~ "dichotomous",
      antihyp_med ~ "dichotomous",
      prev_chd ~ "dichotomous",
      stroke_10yr ~ "dichotomous"
    ),
    value = list(
      diabetes ~ "Yes",
      current_smoker ~ "Yes",
      antihyp_med ~ "Yes",
      prev_chd ~ "Yes",
      stroke_10yr ~ "Yes"
    ),
    label = list(
      AGE ~ "Age (years)",
      SYSBP ~ "Systolic BP (mmHg)",
      TOTCHOL ~ "Total cholesterol (mg/dL)",
      BMI ~ "BMI (kg/m^2)",
      diabetes ~ "Diabetes (yes)",
      current_smoker ~ "Current smoker",
      antihyp_med ~ "Blood Pressure Medication (yes)",
      prev_chd ~ "Prevalent CHD (yes)",
      stroke_10yr ~ "10-year stroke event"
    ),
    missing = "no"
  ) %>%
  add_overall(last = FALSE) %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

table1

#added for survival ready follow-up time and sex stratified datasets

#follow-up and event summary by sex
analytic %>%
  mutate(Sex = ifelse(SEX == 1, "Men", "Women")) %>%
  group_by(Sex) %>%
  summarise(
    N = n(),
    N_stroke_10yr = sum(STROKE_10yr, na.rm = TRUE),
    N_censored_before_10yr = sum(STROKE_10yr == 0 & FU_10yr < days_10yr, na.rm = TRUE),
    N_censored_at_10yr = sum(STROKE_10yr == 0 & FU_10yr >= days_10yr, na.rm = TRUE),
    Mean_followup_days = round(mean(FU_10yr, na.rm = TRUE), 1),
    Median_followup_days = round(median(FU_10yr, na.rm = TRUE), 1)
  )

#create sex stratified datasets for later Kaplan-Meier and Cox models
analytic_men <- analytic %>% filter(SEX == 1)
analytic_women <- analytic %>% filter(SEX == 2)

#check event counts in sex stratified datasets
nrow(analytic_men)
sum(analytic_men$STROKE_10yr, na.rm = TRUE)

nrow(analytic_women)
sum(analytic_women$STROKE_10yr, na.rm = TRUE)




library(dplyr)
library(tidyr)
library(gt)
library(survival)

#New Table 1 with missingness + descriptive follow-up summaries

# helper formatting functions
fmt_mean_sd <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return("NA")
  sprintf(paste0("%.", digits, "f (%.", digits, "f)"), mean(x), sd(x))
}

fmt_n_pct <- function(x, digits = 1) {
  x_nonmiss <- x[!is.na(x)]
  if (length(x_nonmiss) == 0) return("NA")
  n <- sum(x_nonmiss == 1)
  pct <- 100 * n / length(x_nonmiss)
  sprintf("%s (%.1f%%)", format(n, big.mark = ","), pct)
}

fmt_missing_n <- function(x) {
  n_miss <- sum(is.na(x))
  if (n_miss == 0) return(NA_character_)
  format(n_miss, big.mark = ",")
}

fmt_time_mean_sd <- function(time, indicator, value = 1, digits = 1) {
  x <- time[indicator == value & !is.na(indicator) & !is.na(time)]
  if (length(x) == 0) return("NA")
  sprintf(paste0("%.", digits, "f (%.", digits, "f)"), mean(x), sd(x))
}

fmt_km_surv <- function(time, event, t_star = days_10yr, digits = 1) {
  keep <- !is.na(time) & !is.na(event)
  if (sum(keep) == 0) return("NA")
  
  fit <- survfit(Surv(time[keep], event[keep]) ~ 1)
  s <- summary(fit, times = t_star, extend = TRUE)$surv
  
  if (length(s) == 0 || is.na(s)) return("NA")
  sprintf(paste0("%.", digits, "f%%"), 100 * s)
}

#build display dataset
table1_data <- analytic %>%
  mutate(
    sex = factor(SEX, levels = c(1, 2), labels = c("Men", "Women")),
    diabetes = DIABETES,
    current_smoker = CURSMOKE,
    antihyp_med = BPMEDS,
    prev_chd = PREVCHD,
    stroke_10yr = STROKE_10yr
  ) %>%
  select(
    sex,
    AGE,
    SYSBP,
    TOTCHOL,
    BMI,
    diabetes,
    current_smoker,
    antihyp_med,
    prev_chd,
    stroke_10yr,
    FU_10yr
  )

#column Ns
n_overall <- nrow(table1_data)
n_men <- sum(table1_data$sex == "Men")
n_women <- sum(table1_data$sex == "Women")

#helper functions for one row across overall / men / women
get_cont_row <- function(var, label) {
  tibble(
    Characteristic = label,
    Overall = fmt_mean_sd(table1_data[[var]]),
    Men = fmt_mean_sd(table1_data %>% filter(sex == "Men") %>% pull(all_of(var))),
    Women = fmt_mean_sd(table1_data %>% filter(sex == "Women") %>% pull(all_of(var)))
  )
}

get_missing_row <- function(var) {
  miss_overall <- fmt_missing_n(table1_data[[var]])
  if (is.na(miss_overall)) return(NULL)
  
  tibble(
    Characteristic = "Missing",
    Overall = miss_overall,
    Men = fmt_missing_n(table1_data %>% filter(sex == "Men") %>% pull(all_of(var))),
    Women = fmt_missing_n(table1_data %>% filter(sex == "Women") %>% pull(all_of(var)))
  )
}

get_bin_row <- function(var, label) {
  tibble(
    Characteristic = label,
    Overall = fmt_n_pct(table1_data[[var]]),
    Men = fmt_n_pct(table1_data %>% filter(sex == "Men") %>% pull(all_of(var))),
    Women = fmt_n_pct(table1_data %>% filter(sex == "Women") %>% pull(all_of(var)))
  )
}

#baseline rows
table1_rows <- bind_rows(
  get_cont_row("AGE", "Age (years)"),
  get_cont_row("SYSBP", "Systolic BP (mmHg)"),
  get_cont_row("TOTCHOL", "Total cholesterol (mg/dL)"),
  get_missing_row("TOTCHOL"),
  get_cont_row("BMI", "BMI (kg/m^2)"),
  get_missing_row("BMI"),
  get_bin_row("diabetes", "Diabetes mellitus"),
  get_bin_row("current_smoker", "Current smoker"),
  get_bin_row("antihyp_med", "Antihypertensive medication"),
  get_missing_row("antihyp_med"),
  get_bin_row("prev_chd", "Prevalent CHD"),
  get_bin_row("stroke_10yr", "10-year stroke event")
)

#follow-up / survival descriptive rows
followup_rows <- bind_rows(
  tibble(
    Characteristic = "Avg time to event, mean (SD), days",
    Overall = fmt_time_mean_sd(table1_data$FU_10yr, table1_data$stroke_10yr, value = 1),
    Men = fmt_time_mean_sd(
      table1_data %>% filter(sex == "Men") %>% pull(FU_10yr),
      table1_data %>% filter(sex == "Men") %>% pull(stroke_10yr),
      value = 1
    ),
    Women = fmt_time_mean_sd(
      table1_data %>% filter(sex == "Women") %>% pull(FU_10yr),
      table1_data %>% filter(sex == "Women") %>% pull(stroke_10yr),
      value = 1
    )
  ),
  tibble(
    Characteristic = "Avg time to censoring, mean (SD), days",
    Overall = fmt_time_mean_sd(table1_data$FU_10yr, table1_data$stroke_10yr, value = 0),
    Men = fmt_time_mean_sd(
      table1_data %>% filter(sex == "Men") %>% pull(FU_10yr),
      table1_data %>% filter(sex == "Men") %>% pull(stroke_10yr),
      value = 0
    ),
    Women = fmt_time_mean_sd(
      table1_data %>% filter(sex == "Women") %>% pull(FU_10yr),
      table1_data %>% filter(sex == "Women") %>% pull(stroke_10yr),
      value = 0
    )
  ),
  tibble(
    Characteristic = "10-year stroke-free survival from KM",
    Overall = fmt_km_surv(table1_data$FU_10yr, table1_data$stroke_10yr),
    Men = fmt_km_surv(
      table1_data %>% filter(sex == "Men") %>% pull(FU_10yr),
      table1_data %>% filter(sex == "Men") %>% pull(stroke_10yr)
    ),
    Women = fmt_km_surv(
      table1_data %>% filter(sex == "Women") %>% pull(FU_10yr),
      table1_data %>% filter(sex == "Women") %>% pull(stroke_10yr)
    )
  )
)

table1_final <- bind_rows(table1_rows, followup_rows)

#table 1
table1_gt <- table1_final %>%
  gt() %>%
  cols_label(
    Characteristic = "Characteristic",
    Overall = md(paste0("**Overall N = ", n_overall, "**")),
    Men = md(paste0("**Men N = ", n_men, "**")),
    Women = md(paste0("**Women N = ", n_women, "**"))
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      rows = Characteristic %in% c(
        "Age (years)",
        "Systolic BP (mmHg)",
        "Total cholesterol (mg/dL)",
        "BMI (kg/m^2)",
        "Diabetes mellitus",
        "Current smoker",
        "Antihypertensive medication",
        "Prevalent CHD",
        "10-year stroke event",
        "Avg time to event, mean (SD), days",
        "Avg time to censoring, mean (SD), days",
        "10-year stroke-free survival from KM"
      )
    )
  ) %>%
  tab_source_note(
    source_note = md(
      "Mean (SD); n (%). Missing rows report number missing. Percentages for binary variables use the non-missing denominator. The KM row reports estimated stroke-free survival at 10 years."
    )
  )

table1_gt
```

# NOTE: TOTCHOL (52 missing), BMI (17 missing), and BPMEDS (60 missing) are
# retained in the analytic dataset at this stage. Flagged extreme SYSBP and
# TOTCHOL values are also retained per investigator guidance. If any of these
# variables are included in the final model, additional participants will be
# excluded only for missingness and the exclusion table above will be updated
# accordingly.

## Section 4: Descriptive Variable Assessment and Model Preparation

```{r}
library(ggplot2)
library(knitr)
library(gt)
library(dplyr)
library(tibble)

library(ggplot2)
library(knitr)
library(gt)
library(dplyr)
library(tibble)

#Table 2: Distribution of risk factors across examination periods
#using the same baseline-eligible cohort as Table 1

analytic_ids <- analytic %>%
  distinct(RANDID)

table2_data <- data %>%
  semi_join(analytic_ids, by = "RANDID")

#helper functions
cell_cont_mean_sd <- function(x) {
  if (sum(!is.na(x)) == 0) return("—")
  sprintf("%.1f (%.1f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
}

cell_bin_n_pct <- function(x, level) {
  n_valid <- sum(!is.na(x))
  if (n_valid == 0) return("—")
  n_level <- sum(x == level, na.rm = TRUE)
  sprintf("%d (%.1f%%)", n_level, 100 * n_level / n_valid)
}

n_nonmissing <- function(x) {
  sum(!is.na(x))
}

n_level <- function(x, level) {
  sum(x == level, na.rm = TRUE)
}

#helper to build one continuous row
build_cont_row_t2 <- function(var_name, display_label) {
  get_n <- function(period) {
    x <- table2_data[[var_name]][table2_data$PERIOD == period]
    as.character(n_nonmissing(x))
  }
  
  get_val <- function(period) {
    x <- table2_data[[var_name]][table2_data$PERIOD == period]
    cell_cont_mean_sd(x)
  }
  
  tibble(
    Variable = display_label,
    p1_n = get_n(1),
    p1_value = get_val(1),
    p2_n = get_n(2),
    p2_value = get_val(2),
    p3_n = get_n(3),
    p3_value = get_val(3)
  )
}

#helper to build one binary section with No and Yes rows
build_bin_rows_t2 <- function(var_name, display_label) {
  get_x <- function(period) {
    table2_data[[var_name]][table2_data$PERIOD == period]
  }
  
  bind_rows(
    tibble(
      Variable = display_label,
      p1_n = "",
      p1_value = "",
      p2_n = "",
      p2_value = "",
      p3_n = "",
      p3_value = ""
    ),
    tibble(
      Variable = "  No",
      p1_n = as.character(n_level(get_x(1), 0)),
      p1_value = cell_bin_n_pct(get_x(1), 0),
      p2_n = as.character(n_level(get_x(2), 0)),
      p2_value = cell_bin_n_pct(get_x(2), 0),
      p3_n = as.character(n_level(get_x(3), 0)),
      p3_value = cell_bin_n_pct(get_x(3), 0)
    ),
    tibble(
      Variable = "  Yes",
      p1_n = as.character(n_level(get_x(1), 1)),
      p1_value = cell_bin_n_pct(get_x(1), 1),
      p2_n = as.character(n_level(get_x(2), 1)),
      p2_value = cell_bin_n_pct(get_x(2), 1),
      p3_n = as.character(n_level(get_x(3), 1)),
      p3_value = cell_bin_n_pct(get_x(3), 1)
    )
  )
}

#table body
table2_final <- bind_rows(
  build_cont_row_t2("AGE",     "Age (years)"),
  build_cont_row_t2("SYSBP",   "Systolic BP (mmHg)"),
  build_cont_row_t2("TOTCHOL", "Total cholesterol (mg/dL)"),
  build_cont_row_t2("BMI",     "BMI (kg/m^2)"),
  build_bin_rows_t2("DIABETES", "Diabetes"),
  build_bin_rows_t2("CURSMOKE", "Current smoker"),
  build_bin_rows_t2("BPMEDS",   "BP medication"),
  build_bin_rows_t2("PREVCHD",  "Prevalent CHD")
)

#period-specific total N from the baseline-eligible cohort
n_p1 <- sum(table2_data$PERIOD == 1, na.rm = TRUE)
n_p2 <- sum(table2_data$PERIOD == 2, na.rm = TRUE)
n_p3 <- sum(table2_data$PERIOD == 3, na.rm = TRUE)

#display table
table2_gt <- table2_final %>%
  gt() %>%
  tab_spanner(
    label = md(paste0("**Period 1**<br>N = ", n_p1)),
    columns = c(p1_n, p1_value)
  ) %>%
  tab_spanner(
    label = md(paste0("**Period 2**<br>N = ", n_p2)),
    columns = c(p2_n, p2_value)
  ) %>%
  tab_spanner(
    label = md(paste0("**Period 3**<br>N = ", n_p3)),
    columns = c(p3_n, p3_value)
  ) %>%
  cols_label(
    Variable = "Variable",
    p1_n = "n",
    p1_value = "Value",
    p2_n = "n",
    p2_value = "Value",
    p3_n = "n",
    p3_value = "Value"
  ) %>%
  cols_align(
    align = "center",
    columns = c(p1_n, p1_value, p2_n, p2_value, p3_n, p3_value)
  ) %>%
  cols_align(
    align = "left",
    columns = Variable
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      rows = Variable %in% c("Diabetes", "Current smoker", "BP medication", "Prevalent CHD")
    )
  ) %>%
  tab_source_note(
    source_note = md(
      "This table uses the same baseline-eligible cohort as Table 1. For each period, N is the number of available observations from that cohort. Continuous variables are shown as mean (SD). For binary variables, the No and Yes rows show category-specific n (%), where percentages use the non-missing denominator within period."
    )
  )

table2_gt

#define variable sets used below

cont_vars <- c("AGE", "SYSBP", "TOTCHOL", "BMI")
bin_vars <- c("CURSMOKE", "DIABETES", "BPMEDS", "PREVCHD")

#added for categorized systolic blood pressure and model completeness checks

#categorize systolic blood pressure for Kaplan-Meier curves
#split aligns with profile definition (SYSBP=160 used as "high" threshold in risk profiles)
analytic <- analytic %>%
  mutate(
    SYSBP_CAT = case_when(
      SYSBP < 160 ~ "<160",
      SYSBP >= 160 ~ ">=160",
      TRUE ~ NA_character_
    ),
    SYSBP_CAT = factor(SYSBP_CAT, levels = c("<160", ">=160"))
  )

#check distribution of categorized systolic blood pressure by sex
analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex, SYSBP_CAT) %>%
  summarise(n = n(), .groups = "drop")

#check completeness for core and candidate model variables
core_vars <- c("AGE", "SYSBP", "DIABETES")
candidate_vars <- c("CURSMOKE", "BPMEDS", "PREVCHD", "TOTCHOL", "BMI")
full_model_vars <- c(core_vars, candidate_vars)

model_completeness <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex) %>%
  summarise(
    N_total = n(),
    N_core_complete = sum(complete.cases(across(all_of(core_vars)))),
    N_full_candidate_complete = sum(complete.cases(across(all_of(full_model_vars)))),
    N_add_cursmoke = sum(complete.cases(across(all_of(c(core_vars, "CURSMOKE"))))),
    N_add_bpmeds = sum(complete.cases(across(all_of(c(core_vars, "BPMEDS"))))),
    N_add_prevchd = sum(complete.cases(across(all_of(c(core_vars, "PREVCHD"))))),
    N_add_totchol = sum(complete.cases(across(all_of(c(core_vars, "TOTCHOL"))))),
    N_add_bmi = sum(complete.cases(across(all_of(c(core_vars, "BMI"))))),
    .groups = "drop"
  )

model_completeness

#more detailed summaries of continuous model variables by sex
cont_distribution_summary <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  select(sex, all_of(cont_vars)) %>%
  pivot_longer(cols = all_of(cont_vars), names_to = "variable", values_to = "value") %>%
  group_by(sex, variable) %>%
  summarise(
    n_nonmissing = sum(!is.na(value)),
    mean = round(mean(value, na.rm = TRUE), 1),
    sd = round(sd(value, na.rm = TRUE), 1),
    median = round(median(value, na.rm = TRUE), 1),
    q1 = round(quantile(value, 0.25, na.rm = TRUE), 1),
    q3 = round(quantile(value, 0.75, na.rm = TRUE), 1),
    min = round(min(value, na.rm = TRUE), 1),
    max = round(max(value, na.rm = TRUE), 1),
    .groups = "drop"
  )

cont_distribution_summary

#look at univariate distributions by sex for continuous variables going into the models
#this is the main visual check for whether the continuous variables appear skewed

analytic %>%
  select(all_of(cont_vars), SEX) %>%
  pivot_longer(all_of(cont_vars), names_to = "variable", values_to = "value") %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  ggplot(aes(x = value, fill = sex)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  facet_wrap(~ variable, scales = "free") +
  labs(
    title = "distribution of continuous variables by sex",
    x = NULL,
    y = "count",
    fill = "sex"
  ) +
  theme_minimal()

#check sysbp and bpmeds relationship

sysbp_bpmeds_summary <- analytic %>%
  mutate(
    sex = ifelse(SEX == 1, "men", "women"),
    bpmeds_cat = case_when(
      BPMEDS == 0 ~ "no",
      BPMEDS == 1 ~ "yes",
      TRUE ~ "missing"
    ),
    bpmeds_cat = factor(bpmeds_cat, levels = c("no", "yes", "missing"))
  ) %>%
  group_by(sex, bpmeds_cat) %>%
  summarise(
    n = n(),
    sysbp_mean = round(mean(SYSBP, na.rm = TRUE), 1),
    sysbp_sd = round(sd(SYSBP, na.rm = TRUE), 1),
    .groups = "drop"
  )

sysbp_bpmeds_summary

sysbp_bpmeds_corr <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex) %>%
  summarise(
    corr_sysbp_bpmeds = round(cor(SYSBP, BPMEDS, use = "complete.obs"), 3),
    .groups = "drop"
  )

sysbp_bpmeds_corr

sysbp_bpmeds_test <- analytic %>%
  filter(!is.na(BPMEDS)) %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex) %>%
  summarise(
    p_value = wilcox.test(SYSBP ~ BPMEDS)$p.value,
    .groups = "drop"
  )

sysbp_bpmeds_test

analytic %>%
  mutate(
    sex = ifelse(SEX == 1, "men", "women"),
    bpmeds_cat = case_when(
      BPMEDS == 0 ~ "no",
      BPMEDS == 1 ~ "yes",
      TRUE ~ "missing"
    ),
    bpmeds_cat = factor(bpmeds_cat, levels = c("no", "yes", "missing"))
  ) %>%
  ggplot(aes(x = bpmeds_cat, y = SYSBP, fill = sex)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~ sex) +
  labs(
    title = "systolic blood pressure by blood pressure medication status",
    x = "blood pressure medication status",
    y = "systolic blood pressure (mmHg)",
    fill = "sex"
  ) +
  theme_minimal()

#check correlations among continuous candidate predictors

cont_corr <- analytic %>%
  select(all_of(cont_vars)) %>%
  cor(use = "pairwise.complete.obs")

round(cont_corr, 3)

#check associations between continuous and binary candidate predictors

cont_bin_assoc <- expand_grid(cont_var = cont_vars, bin_var = bin_vars) %>%
  rowwise() %>%
  mutate(
    n_complete = sum(!is.na(analytic[[cont_var]]) & !is.na(analytic[[bin_var]])),
    corr = round(cor(analytic[[cont_var]], analytic[[bin_var]], use = "complete.obs"), 3),
    p_value = {
      df_use <- analytic %>%
        select(all_of(c(cont_var, bin_var))) %>%
        drop_na()
      wilcox.test(df_use[[cont_var]] ~ df_use[[bin_var]])$p.value
    },
    mean_if_no = {
      df_use <- analytic %>%
        select(all_of(c(cont_var, bin_var))) %>%
        drop_na()
      round(mean(df_use[[cont_var]][df_use[[bin_var]] == 0], na.rm = TRUE), 1)
    },
    mean_if_yes = {
      df_use <- analytic %>%
        select(all_of(c(cont_var, bin_var))) %>%
        drop_na()
      round(mean(df_use[[cont_var]][df_use[[bin_var]] == 1], na.rm = TRUE), 1)
    }
  ) %>%
  ungroup()

cont_bin_assoc

#check associations among binary candidate predictors

binary_assoc <- expand_grid(var1 = bin_vars, var2 = bin_vars) %>%
  filter(var1 < var2) %>%
  rowwise() %>%
  mutate(
    n_complete = sum(!is.na(analytic[[var1]]) & !is.na(analytic[[var2]])),
    p_value = {
      keep <- !is.na(analytic[[var1]]) & !is.na(analytic[[var2]])
      tbl <- table(analytic[[var1]][keep], analytic[[var2]][keep])
      chisq_obj <- suppressWarnings(chisq.test(tbl, correct = FALSE))
      
      if (any(chisq_obj$expected < 5)) {
        fisher.test(tbl)$p.value
      } else {
        chisq_obj$p.value
      }
    },
    test_used = {
      keep <- !is.na(analytic[[var1]]) & !is.na(analytic[[var2]])
      tbl <- table(analytic[[var1]][keep], analytic[[var2]][keep])
      chisq_obj <- suppressWarnings(chisq.test(tbl, correct = FALSE))
      
      if (any(chisq_obj$expected < 5)) {
        "fisher"
      } else {
        "chi-square"
      }
    }
  ) %>%
  ungroup()

binary_assoc

#do bivariate analysis by sex for continuous variables

#added for more detailed bivariate summaries of continuous variables by sex

cont_results_detailed <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_split(sex) %>%
  map_dfr(function(df_sex) {
    
    sex_label <- unique(df_sex$sex)
    
    map_dfr(cont_vars, function(v) {
      df_use <- df_sex %>% filter(!is.na(.data[[v]]))
      
      tibble(
        sex = sex_label,
        variable = v,
        stroke_mean_sd = df_use %>%
          filter(STROKE_10yr == 1) %>%
          summarise(txt = sprintf("%.1f (%.1f)",
                                  mean(.data[[v]], na.rm = TRUE),
                                  sd(.data[[v]], na.rm = TRUE))) %>%
          pull(txt),
        no_stroke_mean_sd = df_use %>%
          filter(STROKE_10yr == 0) %>%
          summarise(txt = sprintf("%.1f (%.1f)",
                                  mean(.data[[v]], na.rm = TRUE),
                                  sd(.data[[v]], na.rm = TRUE))) %>%
          pull(txt),
        stroke_median_iqr = df_use %>%
          filter(STROKE_10yr == 1) %>%
          summarise(txt = sprintf("%.1f (%.1f, %.1f)",
                                  median(.data[[v]], na.rm = TRUE),
                                  quantile(.data[[v]], 0.25, na.rm = TRUE),
                                  quantile(.data[[v]], 0.75, na.rm = TRUE))) %>%
          pull(txt),
        no_stroke_median_iqr = df_use %>%
          filter(STROKE_10yr == 0) %>%
          summarise(txt = sprintf("%.1f (%.1f, %.1f)",
                                  median(.data[[v]], na.rm = TRUE),
                                  quantile(.data[[v]], 0.25, na.rm = TRUE),
                                  quantile(.data[[v]], 0.75, na.rm = TRUE))) %>%
          pull(txt),
        p_value = round(wilcox.test(df_use[[v]] ~ df_use$STROKE_10yr)$p.value, 4)
      )
    })
  })

cont_results_detailed

#do bivariate analysis by sex for binary variables
bin_results <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_split(sex) %>%
  map_dfr(function(df_sex) {
    
    sex_label <- unique(df_sex$sex)
    
    map_dfr(bin_vars, function(v) {
      df_use <- df_sex[!is.na(df_sex[[v]]), ]
      tbl <- table(df_use[[v]], df_use$STROKE_10yr)
      
      test <- if (any(tbl < 5)) {
        fisher.test(tbl)
      } else {
        suppressWarnings(chisq.test(tbl))
      }
      
      stroke_pct_no <- df_use %>%
        filter(.data[[v]] == 0) %>%
        summarise(stroke_pct = round(mean(STROKE_10yr) * 100, 1)) %>%
        pull(stroke_pct)
      
      stroke_pct_yes <- df_use %>%
        filter(.data[[v]] == 1) %>%
        summarise(stroke_pct = round(mean(STROKE_10yr) * 100, 1)) %>%
        pull(stroke_pct)
      
      tibble(
        sex = sex_label,
        variable = v,
        stroke_pct_no = stroke_pct_no,
        stroke_pct_yes = stroke_pct_yes,
        p_value = round(test$p.value, 4),
        test_used = if (any(tbl < 5)) "fisher" else "chi-square"
      )
    })
  })

bin_results
```




## Section 5: Kaplan-Meier Survival Analysis

```{r}
library(survival)
library(survminer)

#build labels, then subset into male and female

#since stroke events are rare, curves will be plotted as cumulative incidence (1 - S(t)) via fun = "event" 

#build all labeled factor variables once the analytic dataset
analytic <- analytic %>%
  mutate(
    sex_label      = factor(SEX,      levels = c(1, 2), labels = c("Men", "Women")),
    diabetes_label = factor(DIABETES, levels = c(0, 1), labels = c("No diabetes", "Diabetes")),
    smoke_label    = factor(CURSMOKE, levels = c(0, 1), labels = c("Not current", "Current")),
    prevchd_label  = factor(PREVCHD,  levels = c(0, 1), labels = c("No CHD", "CHD")),
    bpmeds_label   = factor(BPMEDS,   levels = c(0, 1), labels = c("No BP meds", "BP meds"))
  )

#sex-stratified datasets
analytic_men   <- analytic %>% filter(SEX == 1)
analytic_women <- analytic %>% filter(SEX == 2)

#BPMEDS has missing values- create complete-case subsets for BPMEDS-stratified
#log-rank tests so the reported N reflects the comparison actually being made
analytic_men_bpmeds   <- analytic_men   %>% filter(!is.na(BPMEDS))
analytic_women_bpmeds <- analytic_women %>% filter(!is.na(BPMEDS))

#follow-up summary by sex

followup_summary <- analytic %>%
  group_by(sex_label) %>%
  summarise(
    n                    = n(),
    n_events             = sum(STROKE_10yr, na.rm = TRUE),
    median_followup_days = round(median(FU_10yr, na.rm = TRUE), 1),
    q1_followup_days     = round(quantile(FU_10yr, 0.25, na.rm = TRUE), 1),
    q3_followup_days     = round(quantile(FU_10yr, 0.75, na.rm = TRUE), 1),
    .groups = "drop"
  )

followup_summary

#Helper functions

#KM plots formatting
plot_km_event <- function(fit, data, title) {
  ggsurvplot(
    fit,
    data               = data,
    fun                = "event",
    conf.int           = FALSE,
    censor             = FALSE,
    risk.table         = TRUE,
    risk.table.height  = 0.22,
    risk.table.fontsize = 3.2,
    break.time.by      = 730,
    xlim               = c(0, days_10yr),
    xlab               = "Follow-up time (days)",
    ylab               = "Cumulative stroke probability",
    title              = title,
    ggtheme            = theme_minimal(base_size = 12),
    tables.theme       = theme_cleantable(base_size = 10)
  )
}

#log-rank test summary for each KM plot
logrank_summary <- function(formula, data, group_var) {
  fit <- survdiff(formula, data = data)
  tibble(
    variable = group_var,
    chisq    = round(fit$chisq, 3),
    df       = length(fit$n) - 1,
    p_value  = round(1 - pchisq(fit$chisq, df = length(fit$n) - 1), 4),
    n_total  = sum(fit$n)
  )
}

#overall KM by sex

km_sex <- survfit(Surv(FU_10yr, STROKE_10yr) ~ sex_label, data = analytic)
plot_km_event(km_sex, analytic, "Cumulative stroke probability by sex")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ sex_label, analytic, "Sex")

#KM plots within men

km_men_diabetes <- survfit(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, data = analytic_men)
plot_km_event(km_men_diabetes, analytic_men, "Men: cumulative stroke probability by diabetes")

km_men_smoke <- survfit(Surv(FU_10yr, STROKE_10yr) ~ smoke_label, data = analytic_men)
plot_km_event(km_men_smoke, analytic_men, "Men: cumulative stroke probability by current smoking")

km_men_bpmeds <- survfit(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label, data = analytic_men_bpmeds)
plot_km_event(km_men_bpmeds, analytic_men_bpmeds, "Men: cumulative stroke probability by BP medication use")

km_men_prevchd <- survfit(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label, data = analytic_men)
plot_km_event(km_men_prevchd, analytic_men, "Men: cumulative stroke probability by prevalent CHD")

km_men_sysbp <- survfit(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT, data = analytic_men)
plot_km_event(km_men_sysbp, analytic_men, "Men: cumulative stroke probability by systolic BP category")

#KM plots within women

km_women_diabetes <- survfit(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, data = analytic_women)
plot_km_event(km_women_diabetes, analytic_women, "Women: cumulative stroke probability by diabetes")

km_women_smoke <- survfit(Surv(FU_10yr, STROKE_10yr) ~ smoke_label, data = analytic_women)
plot_km_event(km_women_smoke, analytic_women, "Women: cumulative stroke probability by current smoking")

km_women_bpmeds <- survfit(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label, data = analytic_women_bpmeds)
plot_km_event(km_women_bpmeds, analytic_women_bpmeds, "Women: cumulative stroke probability by BP medication use")

km_women_prevchd <- survfit(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label, data = analytic_women)
plot_km_event(km_women_prevchd, analytic_women, "Women: cumulative stroke probability by prevalent CHD")

km_women_sysbp <- survfit(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT, data = analytic_women)
plot_km_event(km_women_sysbp, analytic_women, "Women: cumulative stroke probability by systolic BP category")

#Combined log-rank summary table

km_logrank_results <- bind_rows(
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ sex_label,      analytic,              "Overall: sex"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_men,          "Men: diabetes"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label,    analytic_men,          "Men: current smoking"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label,   analytic_men_bpmeds,   "Men: BP medication"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label,  analytic_men,          "Men: prevalent CHD"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT,      analytic_men,          "Men: systolic BP category"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_women,        "Women: diabetes"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label,    analytic_women,        "Women: current smoking"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label,   analytic_women_bpmeds, "Women: BP medication"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label,  analytic_women,        "Women: prevalent CHD"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT,      analytic_women,        "Women: systolic BP category")
)

knitr::kable(km_logrank_results, digits = 4)
```

## Section 6: Cox Proportional Hazards Models with LASSO Variable Selection

```{r}
library(survival)
library(glmnet)
library(broom)
library(tidyr)

#added for sex-stratified Cox PH models with LASSO-based variable selection

#define core and candidate predictors for Cox modeling
core_vars <- c("AGE", "SYSBP", "DIABETES")
candidate_vars <- c("CURSMOKE", "BPMEDS", "PREVCHD", "TOTCHOL", "BMI")
all_model_vars <- c(core_vars, candidate_vars)

#restrict to complete cases for variables considered in the LASSO step
cox_men_data <- analytic_men %>%
  select(FU_10yr, STROKE_10yr, all_of(all_model_vars)) %>%
  drop_na()

cox_women_data <- analytic_women %>%
  select(FU_10yr, STROKE_10yr, all_of(all_model_vars)) %>%
  drop_na()

#check sample sizes after complete-case restriction for LASSO modeling
lasso_sample_summary <- tibble(
  sex = c("men", "women"),
  n_analytic = c(nrow(analytic_men), nrow(analytic_women)),
  n_lasso_complete_case = c(nrow(cox_men_data), nrow(cox_women_data)),
  n_dropped = c(nrow(analytic_men) - nrow(cox_men_data),
                nrow(analytic_women) - nrow(cox_women_data)),
  n_events = c(sum(cox_men_data$STROKE_10yr), sum(cox_women_data$STROKE_10yr))
)

lasso_sample_summary

#create survival responses for glmnet
y_men <- with(cox_men_data, Surv(FU_10yr, STROKE_10yr))
y_women <- with(cox_women_data, Surv(FU_10yr, STROKE_10yr))

#create design matrices for glmnet (no intercept column)
x_men <- model.matrix(
  ~ AGE + SYSBP + DIABETES + CURSMOKE + BPMEDS + PREVCHD + TOTCHOL + BMI,
  data = cox_men_data
)[, -1]

x_women <- model.matrix(
  ~ AGE + SYSBP + DIABETES + CURSMOKE + BPMEDS + PREVCHD + TOTCHOL + BMI,
  data = cox_women_data
)[, -1]

#penalty factors:
#force AGE, SYSBP, and DIABETES into the model by setting penalty = 0
penalty_men <- ifelse(colnames(x_men) %in% core_vars, 0, 1)
penalty_women <- ifelse(colnames(x_women) %in% core_vars, 0, 1)

#cross-validated LASSO Cox models by sex
#set seed once for reproducibility
set.seed(6624)

cv_lasso_men <- cv.glmnet(
  x = x_men,
  y = y_men,
  family = "cox",
  alpha = 1,
  penalty.factor = penalty_men,
  nfolds = 5,
  standardize = TRUE
)

cv_lasso_women <- cv.glmnet(
  x = x_women,
  y = y_women,
  family = "cox",
  alpha = 1,
  penalty.factor = penalty_women,
  nfolds = 5,
  standardize = TRUE
)

#plot cross-validation curves
plot(cv_lasso_men)
plot(cv_lasso_women)

#extract selected coefficients at lambda.min
coef_men_min <- coef(cv_lasso_men, s = "lambda.min")
coef_women_min <- coef(cv_lasso_women, s = "lambda.min")

selected_men_min <- rownames(coef_men_min)[as.vector(coef_men_min != 0)]
selected_women_min <- rownames(coef_women_min)[as.vector(coef_women_min != 0)]

#extract selected coefficients at lambda.1se
coef_men_1se <- coef(cv_lasso_men, s = "lambda.1se")
coef_women_1se <- coef(cv_lasso_women, s = "lambda.1se")

selected_men_1se <- rownames(coef_men_1se)[as.vector(coef_men_1se != 0)]
selected_women_1se <- rownames(coef_women_1se)[as.vector(coef_women_1se != 0)]

selected_men_min
selected_women_min
selected_men_1se
selected_women_1se

#summarize selected predictors at lambda.min and lambda.1se
lasso_selected_summary <- tibble(
  sex = c("men", "women"),
  lambda_min = c(cv_lasso_men$lambda.min, cv_lasso_women$lambda.min),
  lambda_1se = c(cv_lasso_men$lambda.1se, cv_lasso_women$lambda.1se),
  selected_predictors_lambda_min = c(
    paste(selected_men_min, collapse = ", "),
    paste(selected_women_min, collapse = ", ")
  ),
  selected_predictors_lambda_1se = c(
    paste(selected_men_1se, collapse = ", "),
    paste(selected_women_1se, collapse = ", ")
  )
)

lasso_selected_summary
knitr::kable(lasso_selected_summary, digits = 4)

#primary analysis uses lambda.min
#force matching variable sets across sexes by using the union of selections
final_vars_union <- sort(unique(c(selected_men_min, selected_women_min)))

final_vars_union

final_vars_men <- final_vars_union
final_vars_women <- final_vars_union

#refit datasets: complete-case on only the variables actually in the final model
#this can recover observations dropped during LASSO if unused candidates had missingness
refit_men_data <- analytic_men %>%
  select(FU_10yr, STROKE_10yr, all_of(final_vars_men)) %>%
  drop_na()

refit_women_data <- analytic_women %>%
  select(FU_10yr, STROKE_10yr, all_of(final_vars_women)) %>%
  drop_na()

#report N loss for the refit step
refit_sample_summary <- tibble(
  sex = c("men", "women"),
  n_lasso = c(nrow(cox_men_data), nrow(cox_women_data)),
  n_refit = c(nrow(refit_men_data), nrow(refit_women_data)),
  n_recovered = c(nrow(refit_men_data) - nrow(cox_men_data),
                  nrow(refit_women_data) - nrow(cox_women_data)),
  n_events_refit = c(sum(refit_men_data$STROKE_10yr), sum(refit_women_data$STROKE_10yr))
)

refit_sample_summary

#refit standard coxph models with predictors selected by LASSO
formula_men <- as.formula(
  paste("Surv(FU_10yr, STROKE_10yr) ~", paste(final_vars_men, collapse = " + "))
)

formula_women <- as.formula(
  paste("Surv(FU_10yr, STROKE_10yr) ~", paste(final_vars_women, collapse = " + "))
)

cox_final_men <- coxph(formula_men, data = refit_men_data)
cox_final_women <- coxph(formula_women, data = refit_women_data)

summary(cox_final_men)
summary(cox_final_women)

#tidy final model results with hazard ratios and 95% confidence intervals
#rescale SYSBP and TOTCHOL for interpretability
rescale_hr <- function(tidy_df) {
  tidy_df %>%
    mutate(
      scale_factor = case_when(
        term == "SYSBP" ~ 10,
        term == "TOTCHOL" ~ 10,
        TRUE ~ 1
      ),
      estimate = exp(estimate * scale_factor),
      conf.low = exp(conf.low * scale_factor),
      conf.high = exp(conf.high * scale_factor),
      term = case_when(
        term == "SYSBP" ~ "SYSBP (per 10 mmHg)",
        term == "TOTCHOL" ~ "TOTCHOL (per 10 mg/dL)",
        term == "AGE" ~ "AGE (per 1 year)",
        term == "BMI" ~ "BMI (per 1 kg/m^2)",
        TRUE ~ term
      )
    )
}

cox_final_men_tidy <- tidy(cox_final_men, conf.int = TRUE) %>%
  rescale_hr() %>%
  mutate(sex = "men")

cox_final_women_tidy <- tidy(cox_final_women, conf.int = TRUE) %>%
  rescale_hr() %>%
  mutate(sex = "women")

cox_final_results <- bind_rows(cox_final_men_tidy, cox_final_women_tidy) %>%
  select(sex, term, estimate, conf.low, conf.high, p.value)

cox_final_results

#Table 3: Adjusted hazard ratios from final Cox PH models by sex

format_hr_ci <- function(hr, lo, hi) {
  sprintf("%.2f (%.2f, %.2f)", hr, lo, hi)
}

format_p <- function(p) {
  ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))
}

table3 <- cox_final_results %>%
  mutate(
    hr_ci = format_hr_ci(estimate, conf.low, conf.high),
    p_fmt = format_p(p.value)
  ) %>%
  select(sex, term, hr_ci, p_fmt) %>%
  pivot_wider(
    names_from = sex,
    values_from = c(hr_ci, p_fmt),
    names_glue = "{sex}_{.value}"
  ) %>%
  select(term, men_hr_ci, men_p_fmt, women_hr_ci, women_p_fmt)

kableExtra::kbl(
  table3,
  format = "html",
  col.names = c(
    "Variable",
    "HR (95% CI)", "p-value",
    "HR (95% CI)", "p-value"
  ),
  align = c("l", "c", "c", "c", "c"),
  booktabs = TRUE,
  escape = FALSE,
  caption = paste(
    "**Table 3.** Adjusted hazard ratios for 10-year incident stroke from the",
    "sex-specific Cox proportional hazards models. The final predictor set was",
    "based on the union of variables selected across the male and female LASSO",
    "models so that both Cox models used the same covariates. Age, systolic blood",
    "pressure, and diabetes were forced into the model. Hazard ratios for systolic",
    "blood pressure are reported per 10 mmHg increase; age is reported per 1-year increase."
  ),
  table.attr = 'style="width:auto; caption-side: bottom;"'
) %>%
  kableExtra::add_header_above(c(" " = 1, "Men" = 2, "Women" = 2)) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 11
  ) %>%
  kableExtra::column_spec(1, width = "2.4in") %>%
  kableExtra::column_spec(2, width = "1.25in") %>%
  kableExtra::column_spec(3, width = "0.8in") %>%
  kableExtra::column_spec(4, width = "1.25in") %>%
  kableExtra::column_spec(5, width = "0.8in")

#save selected variable lists for later prediction section
final_vars_men
final_vars_women
```


## Section 7: Proportional Hazards Diagnostics for Final Cox Models

```{r}
#Section 7.1 schoenfeld residual tests
library(survival)
library(survminer)
library(dplyr)
library(tibble)


ph_men <- cox.zph(cox_final_men)
ph_women <- cox.zph(cox_final_women)

ph_men
ph_women

#combine Schoenfeld test results into one table
ph_men_table <- as.data.frame(ph_men$table) %>%
  rownames_to_column("term") %>%
  mutate(sex = "men")

ph_women_table <- as.data.frame(ph_women$table) %>%
  rownames_to_column("term") %>%
  mutate(sex = "women")

ph_results <- bind_rows(ph_men_table, ph_women_table) %>%
  select(sex, term, chisq, df, p) %>%
  rename(p_value = p)

ph_results
knitr::kable(
  ph_results,
  digits = 4,
  col.names = c("Sex", "Term", "Chi-square", "DF", "P-value"),
  caption = "Schoenfeld residual tests of proportional hazards by sex"
)
```

```{r}
#Section 7.2 Schoenfeld residual plots
plot(
  ph_men, var = "AGE",
  main = "Men: Schoenfeld residual plot for age",
  xlab = "Follow-up time (days)",
  ylab = "Scaled Schoenfeld residuals"
)

plot(
  ph_men, var = "SYSBP",
  main = "Men: Schoenfeld residual plot for systolic blood pressure",
  xlab = "Follow-up time (days)",
  ylab = "Scaled Schoenfeld residuals"
)

plot(
  ph_men, var = "DIABETES",
  main = "Men: Schoenfeld residual plot for diabetes",
  xlab = "Follow-up time (days)",
  ylab = "Scaled Schoenfeld residuals"
)

plot(
  ph_men, var = "CURSMOKE",
  main = "Men: Schoenfeld residual plot for current smoking",
  xlab = "Follow-up time (days)",
  ylab = "Scaled Schoenfeld residuals"
)

plot(
  ph_men, var = "BPMEDS",
  main = "Men: Schoenfeld residual plot for blood pressure medication",
  xlab = "Follow-up time (days)",
  ylab = "Scaled Schoenfeld residuals"
)

plot(
  ph_women, var = "AGE",
  main = "Women: Schoenfeld residual plot for age",
  xlab = "Follow-up time (days)",
  ylab = "Scaled Schoenfeld residuals"
)

plot(
  ph_women, var = "SYSBP",
  main = "Women: Schoenfeld residual plot for systolic blood pressure",
  xlab = "Follow-up time (days)",
  ylab = "Scaled Schoenfeld residuals"
)

plot(
  ph_women, var = "DIABETES",
  main = "Women: Schoenfeld residual plot for diabetes",
  xlab = "Follow-up time (days)",
  ylab = "Scaled Schoenfeld residuals"
)

plot(
  ph_women, var = "CURSMOKE",
  main = "Women: Schoenfeld residual plot for current smoking",
  xlab = "Follow-up time (days)",
  ylab = "Scaled Schoenfeld residuals"
)

plot(
  ph_women, var = "BPMEDS",
  main = "Women: Schoenfeld residual plot for blood pressure medication",
  xlab = "Follow-up time (days)",
  ylab = "Scaled Schoenfeld residuals"
)
```

```{r}
#Section 7.3 Log-minus-log survival plots for binary predictors

#for binary predictors in the final model (DIABETES, CURSMOKE, BPMEDS):
#curves that are roughly parallel over time support PH
#men: diabetes
surv_men_diabetes_ll <- survfit(Surv(FU_10yr, STROKE_10yr) ~ DIABETES, data = refit_men_data)
ggsurvplot(
  surv_men_diabetes_ll,
  data = refit_men_data,
  fun = "cloglog",
  conf.int = FALSE,
  censor = FALSE,
  legend.labs = c("No diabetes", "Diabetes"),
  legend.title = "Diabetes",
  xlab = "Follow-up time (days)",
  ylab = "log(-log(Survival))",
  title = "Men: log-minus-log survival for diabetes",
  ggtheme = theme_minimal(base_size = 12)
)

#men: current smoking
surv_men_smoke_ll <- survfit(Surv(FU_10yr, STROKE_10yr) ~ CURSMOKE, data = refit_men_data)
ggsurvplot(
  surv_men_smoke_ll,
  data = refit_men_data,
  fun = "cloglog",
  conf.int = FALSE,
  censor = FALSE,
  legend.labs = c("Not current", "Current"),
  legend.title = "Current smoker",
  xlab = "Follow-up time (days)",
  ylab = "log(-log(Survival))",
  title = "Men: log-minus-log survival for current smoking",
  ggtheme = theme_minimal(base_size = 12)
)

#men: blood pressure medication
surv_men_bpmeds_ll <- survfit(Surv(FU_10yr, STROKE_10yr) ~ BPMEDS, data = refit_men_data)
ggsurvplot(
  surv_men_bpmeds_ll,
  data = refit_men_data,
  fun = "cloglog",
  conf.int = FALSE,
  censor = FALSE,
  legend.labs = c("No BP meds", "BP meds"),
  legend.title = "BP medication",
  xlab = "Follow-up time (days)",
  ylab = "log(-log(Survival))",
  title = "Men: log-minus-log survival for blood pressure medication",
  ggtheme = theme_minimal(base_size = 12)
)

#women: diabetes
surv_women_diabetes_ll <- survfit(Surv(FU_10yr, STROKE_10yr) ~ DIABETES, data = refit_women_data)
ggsurvplot(
  surv_women_diabetes_ll,
  data = refit_women_data,
  fun = "cloglog",
  conf.int = FALSE,
  censor = FALSE,
  legend.labs = c("No diabetes", "Diabetes"),
  legend.title = "Diabetes",
  xlab = "Follow-up time (days)",
  ylab = "log(-log(Survival))",
  title = "Women: log-minus-log survival for diabetes",
  ggtheme = theme_minimal(base_size = 12)
)

#women: current smoking
surv_women_smoke_ll <- survfit(Surv(FU_10yr, STROKE_10yr) ~ CURSMOKE, data = refit_women_data)
ggsurvplot(
  surv_women_smoke_ll,
  data = refit_women_data,
  fun = "cloglog",
  conf.int = FALSE,
  censor = FALSE,
  legend.labs = c("Not current", "Current"),
  legend.title = "Current smoker",
  xlab = "Follow-up time (days)",
  ylab = "log(-log(Survival))",
  title = "Women: log-minus-log survival for current smoking",
  ggtheme = theme_minimal(base_size = 12)
)

#women: blood pressure medication
surv_women_bpmeds_ll <- survfit(Surv(FU_10yr, STROKE_10yr) ~ BPMEDS, data = refit_women_data)
ggsurvplot(
  surv_women_bpmeds_ll,
  data = refit_women_data,
  fun = "cloglog",
  conf.int = FALSE,
  censor = FALSE,
  legend.labs = c("No BP meds", "BP meds"),
  legend.title = "BP medication",
  xlab = "Follow-up time (days)",
  ylab = "log(-log(Survival))",
  title = "Women: log-minus-log survival for blood pressure medication",
  ggtheme = theme_minimal(base_size = 12)
)
```


```{r}
## Section 8: Profile-Specific 10-Year Stroke Probability Predictions

library(survival)
library(dplyr)
library(tidyr)
library(tibble)
library(kableExtra)

# Sex-specific normal SYSBP for low-risk profiles
# normal SYSBP = mean among participants with no diabetes and not current smoker
# computed separately for each sex from the refit sample

sysbp_normal_men <- refit_men_data %>%
  filter(DIABETES == 0, CURSMOKE == 0) %>%
  summarise(mean_sysbp = mean(SYSBP)) %>%
  pull(mean_sysbp)

sysbp_normal_women <- refit_women_data %>%
  filter(DIABETES == 0, CURSMOKE == 0) %>%
  summarise(mean_sysbp = mean(SYSBP)) %>%
  pull(mean_sysbp)

tibble(
  sex = c("men", "women"),
  n_lowrisk_subjects = c(
    sum(refit_men_data$DIABETES == 0 & refit_men_data$CURSMOKE == 0),
    sum(refit_women_data$DIABETES == 0 & refit_women_data$CURSMOKE == 0)
  ),
  normal_sysbp_mean = c(sysbp_normal_men, sysbp_normal_women)
)

# profile construction
# five prespecified profiles, each evaluated at ages 40, 50, 60
# BPMEDS set to 0 (no treatment) in all profiles for interpretability
# high SYSBP profiles represent untreated severe hypertension

build_profiles <- function(sysbp_normal) {
  ages <- c(40, 50, 60)
  
  profile_defs <- tribble(
    ~profile, ~profile_label,                    ~SYSBP,        ~DIABETES, ~CURSMOKE,
    "P1",     "No major risk factors",           sysbp_normal,  0,         0,
    "P2",     "High SYSBP only",                 160,           0,         0,
    "P3",     "Diabetes only",                   sysbp_normal,  1,         0,
    "P4",     "High SYSBP + diabetes",           160,           1,         0,
    "P5",     "High SYSBP + diabetes + smoking", 160,           1,         1
  )
  
  expand_grid(AGE = ages, profile_defs) %>%
    mutate(BPMEDS = 0) %>%
    select(profile, profile_label, AGE, SYSBP, DIABETES, CURSMOKE, BPMEDS)
}

profiles_men   <- build_profiles(sysbp_normal_men)
profiles_women <- build_profiles(sysbp_normal_women)

# display the full profile specifications for transparency
profiles_men
profiles_women

# predict 10-year survival and confidence intervals
# helper to predict 10-year stroke probability with 95% CI for a profile data frame

predict_10yr_stroke <- function(cox_fit, profile_df) {
  sf <- survfit(cox_fit, newdata = profile_df)
  sf_summary <- summary(sf, times = days_10yr)
  
  tibble(
    surv        = as.numeric(sf_summary$surv),
    surv_lower  = as.numeric(sf_summary$lower),
    surv_upper  = as.numeric(sf_summary$upper)
  ) %>%
    mutate(
      stroke_prob       = 1 - surv,
      stroke_prob_lower = 1 - surv_upper,
      stroke_prob_upper = 1 - surv_lower
    ) %>%
    bind_cols(profile_df, .)
}

men_predictions   <- predict_10yr_stroke(cox_final_men, profiles_men)
women_predictions <- predict_10yr_stroke(cox_final_women, profiles_women)

# keep a long-format prediction dataset for later plotting / reporting
all_predictions <- bind_rows(
  men_predictions   %>% mutate(sex = "men"),
  women_predictions %>% mutate(sex = "women")
) %>%
  mutate(
    stroke_prob_pct       = round(stroke_prob * 100, 2),
    stroke_prob_lower_pct = round(stroke_prob_lower * 100, 2),
    stroke_prob_upper_pct = round(stroke_prob_upper * 100, 2)
  ) %>%
  select(
    sex, profile, profile_label, AGE, SYSBP, DIABETES, CURSMOKE, BPMEDS,
    stroke_prob_pct, stroke_prob_lower_pct, stroke_prob_upper_pct
  )

# ---- Table 4: Predicted 10-year stroke probability by sex, age, and risk profile ----

format_pct_ci <- function(est, lo, hi) {
  sprintf("%.1f%% (%.1f, %.1f)", est, lo, hi)
}

# profile definition info, keeping sex-specific SBP values in separate columns
profile_info_men <- profiles_men %>%
  distinct(profile, profile_label, SYSBP, DIABETES, CURSMOKE, BPMEDS) %>%
  rename(Men_SBP = SYSBP)

profile_info_women <- profiles_women %>%
  distinct(profile, SYSBP) %>%
  rename(Women_SBP = SYSBP)

profile_info <- profile_info_men %>%
  left_join(profile_info_women, by = "profile") %>%
  mutate(
    Men_SBP = sprintf("%.1f", Men_SBP),
    Women_SBP = sprintf("%.1f", Women_SBP),
    Diabetes = ifelse(DIABETES == 1, "Yes", "No"),
    Smoking = ifelse(CURSMOKE == 1, "Yes", "No"),
    `BP medication` = ifelse(BPMEDS == 1, "Yes", "No")
  ) %>%
  select(
    profile, profile_label, Diabetes, Smoking, `BP medication`,
    Men_SBP, Women_SBP
  )

# format predictions for wide display
predictions_wide <- all_predictions %>%
  mutate(
    sex = ifelse(sex == "men", "Men", "Women"),
    risk_10yr = format_pct_ci(
      stroke_prob_pct,
      stroke_prob_lower_pct,
      stroke_prob_upper_pct
    ),
    sex_age = paste0(sex, "_Age_", AGE)
  ) %>%
  select(profile, sex_age, risk_10yr) %>%
  pivot_wider(names_from = sex_age, values_from = risk_10yr)

# final table
table4 <- profile_info %>%
  left_join(predictions_wide, by = c("profile")) %>%
  rename(
    Profile = profile,
    Description = profile_label,
    `Men SBP` = Men_SBP,
    `Women SBP` = Women_SBP
  ) %>%
  select(
    Profile, Description, Diabetes, Smoking, `BP medication`,
    `Men SBP`, Men_Age_40, Men_Age_50, Men_Age_60,
    `Women SBP`, Women_Age_40, Women_Age_50, Women_Age_60
  )

kableExtra::kbl(
  table4,
  format = "html",
  col.names = c(
    "Profile", "Description", "Diabetes", "Smoking", "BP medication",
    "SBP", "Age 40", "Age 50", "Age 60",
    "SBP", "Age 40", "Age 50", "Age 60"
  ),
  align = c("l", "l", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c"),
  booktabs = TRUE,
  escape = FALSE,
  caption = paste(
    "**Table 4.** Predicted 10-year stroke probability by sex, age, and prespecified risk profile.",
    "Predicted risks are shown as percentages with 95% confidence intervals from the final sex-specific Cox proportional hazards models."
  ),
  table.attr = 'style="width:auto; caption-side: bottom;"'
) %>%
  kableExtra::add_header_above(c(" " = 5, "Men" = 4, "Women" = 4)) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 10
  ) %>%
  kableExtra::column_spec(1, width = "0.7in") %>%
  kableExtra::column_spec(2, width = "2.0in") %>%
  kableExtra::column_spec(3, width = "0.8in") %>%
  kableExtra::column_spec(4, width = "0.8in") %>%
  kableExtra::column_spec(5, width = "1.0in") %>%
  kableExtra::column_spec(6, width = "0.8in") %>%
  kableExtra::column_spec(7, width = "1.2in") %>%
  kableExtra::column_spec(8, width = "1.2in") %>%
  kableExtra::column_spec(9, width = "1.2in") %>%
  kableExtra::column_spec(10, width = "0.8in") %>%
  kableExtra::column_spec(11, width = "1.2in") %>%
  kableExtra::column_spec(12, width = "1.2in") %>%
  kableExtra::column_spec(13, width = "1.2in")


#figure 1: Predicted 10-year stroke probability by risk profile

library(ggplot2)
library(dplyr)

figure1_data <- all_predictions %>%
  mutate(
    sex = ifelse(sex == "men", "Men", "Women"),
    sex = factor(sex, levels = c("Men", "Women")),
    profile = factor(
      profile,
      levels = c("P1", "P2", "P3", "P4", "P5"),
      labels = c(
        "P1: Low risk",
        "P2: High SBP",
        "P3: Diabetes",
        "P4: SBP + DM",
        "P5: SBP + DM + Smoke"
      )
    )
  )

ggplot(
  figure1_data,
  aes(
    x = AGE,
    y = stroke_prob_pct,
    group = profile,
    color = profile
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = stroke_prob_lower_pct,
      ymax = stroke_prob_upper_pct
    ),
    width = 1.2,
    alpha = 0.8
  ) +
  facet_wrap(~ sex) +
  scale_x_continuous(breaks = c(40, 50, 60)) +
  labs(
    title = "Predicted 10-year stroke probability by age, sex, and risk profile",
    x = "Age",
    y = "Predicted 10-year stroke probability (%)",
    color = "Profile"
  ) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    axis.title.x = element_text(margin = margin(t = 10)),
    plot.margin = margin(10, 10, 20, 10)
  )
```




