---
title: "Exploratory_Analysis_Project3"
output: html_document
---


## Section 1: Baseline Cohort Restriction and 10-Year Stroke Outcome Definition
```{r}
library(tidyverse)

#load and inspect data
data <- read.csv("C:/Users/kokeshem/Spring 2026/Advanced Statistical Methods/Project 3/frmgham2.csv")

str(data)
table(Period = data$PERIOD)
n_distinct(data$RANDID)

#restrict to period 1, then check missingness on variables of interest
baseline <- data %>% filter(PERIOD == 1)

baseline %>%
  select(AGE, SYSBP, TOTCHOL, BMI, DIABETES, BPMEDS, CURSMOKE,
         PREVCHD, PREVSTRK, STROKE, TIMESTRK, DEATH, TIMEDTH) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N_Missing") %>%
  mutate(Pct_Missing = round(N_Missing / nrow(baseline) * 100, 2)) %>%
  arrange(desc(N_Missing))

#exclude prevalent stroke at baseline
analytic <- baseline %>% filter(PREVSTRK == 0)
#Number excluded due to prevalent stroke: 
nrow(baseline) - nrow(analytic)
#Rows remaining after prevalent stroke exclusion:
nrow(analytic)

#apply 10-year censoring (365.25 x 10 = 3652.5 days)
days_10yr <- 365.25 * 10

analytic <- analytic %>%
  mutate(
    #event indicator for stroke occurring within 10 years
    STROKE_10yr = ifelse(STROKE == 1 & TIMESTRK <= days_10yr, 1, 0),

    #observed followup time for 10-year survival analysis:
    #use TIMESTRK directly for the stroke endpoint and administratively censor at 10 years
    FU_10yr = pmin(TIMESTRK, days_10yr),

    #retained for descriptive summaries of stroke time truncated at 10 years
    TIMESTRK_10yr = ifelse(TIMESTRK > days_10yr, days_10yr, TIMESTRK)
  )

#number of strokes after 10 years that were re-coded as censored
sum(analytic$STROKE == 1 & analytic$TIMESTRK > days_10yr)

#check that no strokes should be recorded at day 0 after excluding PREVSTRK == 1
#use TIMESTRK (the original event time) rather than FU_10yr so the check is independent of 10-year truncation
strk_t0 <- analytic %>% filter(STROKE == 1 & TIMESTRK == 0)
nrow(strk_t0)
```

#Section 2: Data Quality and Missingness Checks
```{r}
#summary statistics for continuous variables
analytic %>%
  select(AGE, SYSBP, TOTCHOL, BMI) %>%
  summary()

#outlier flags looped into one table- review for investigator to confirm if any values are biologically implausible and should be excluded
cont_checks <- list(
  AGE     = list(lo = 32,  hi = 81),
  SYSBP   = list(lo = 80,  hi = 250),
  TOTCHOL = list(lo = 100, hi = 600),
  BMI     = list(lo = 15,  hi = 80)
)

map_dfr(names(cont_checks), function(v) {
  tibble(
    Variable   = v,
    Low_bound  = cont_checks[[v]]$lo,
    N_below    = sum(analytic[[v]] < cont_checks[[v]]$lo, na.rm = TRUE),
    High_bound = cont_checks[[v]]$hi,
    N_above    = sum(analytic[[v]] > cont_checks[[v]]$hi, na.rm = TRUE)
  )
})

#binary variable checks — confirm only 0/1 values exist
binary_vars <- c("DIABETES", "BPMEDS", "CURSMOKE", "PREVCHD")

map_dfr(binary_vars, function(v) {
  tibble(
    Variable = v,
    Unique_values = paste(sort(unique(analytic[[v]])), collapse = ", ")
  )
})

#check missingness on analytic sample
analytic %>%
  select(AGE, SYSBP, TOTCHOL, BMI, DIABETES, BPMEDS, CURSMOKE,
         PREVCHD, STROKE_10yr, TIMESTRK_10yr, DEATH, TIMEDTH) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N_Missing") %>%
  mutate(Pct_Missing = round(N_Missing / nrow(analytic) * 100, 2)) %>%
  arrange(desc(N_Missing))

#check if missingness in BPMEDS and TOTCHOL is related to stroke outcome
#stroke rates compared between missing and non-missing groups
analytic %>%
  mutate(miss_BPMEDS = is.na(BPMEDS)) %>%
  group_by(miss_BPMEDS) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#fisher's exact test - BPMEDS missingness vs stroke outcome (small cell sizes)
fisher.test(table(is.na(analytic$BPMEDS), analytic$STROKE_10yr))

analytic %>%
  mutate(miss_TOTCHOL = is.na(TOTCHOL)) %>%
  group_by(miss_TOTCHOL) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#fisher's exact test - TOTCHOL missingness vs stroke outcome
fisher.test(table(is.na(analytic$TOTCHOL), analytic$STROKE_10yr))

analytic %>%
  mutate(miss_BMI = is.na(BMI)) %>%
  group_by(miss_BMI) %>%
  summarise(N = n(), Stroke_pct = round(mean(STROKE_10yr) * 100, 2))

#summarize possible exclusions- biologically impossible values, data errors, and core missing
#reviewed with investigator: flagged extreme SYSBP and TOTCHOL values will be retained
cat("Flagged (SYSBP > 250):", sum(analytic$SYSBP > 250, na.rm = TRUE), "\n")
cat("Flagged (TOTCHOL > 600):", sum(analytic$TOTCHOL > 600, na.rm = TRUE), "\n")
cat("Remaining:", nrow(analytic), "\n")
```

#Section 3: Cohort Summary and Descriptive Characteristics
```{r}
#core covariates (AGE, SYSBP, DIABETES) are complete — no further exclusions needed yet

#Exclusion summary
data.frame(
  Step = c(
    "Period 1 baseline",
    "Exclude prevalent stroke (PREVSTRK = 1)",
    "Analytic cohort for exploratory and survival analyses"
  ),
  N_remaining = c(
    nrow(baseline),
    nrow(baseline %>% filter(PREVSTRK == 0)),
    nrow(analytic)
  ),
  N_excluded = c(
    NA,
    nrow(baseline) - nrow(baseline %>% filter(PREVSTRK == 0)),
    NA
  )
)

#cohort overview table for interim presentation
table_a <- tibble(
  item = c(
    "period 1 baseline observations",
    "excluded prevalent stroke",
    "flagged sysbp > 250 (retained)",
    "flagged totchol > 600 (retained)",
    "exploratory analysis sample",
    "men: n (10-year strokes, %)",
    "women: n (10-year strokes, %)"
  ),
  value = c(
    nrow(baseline),
    nrow(baseline) - nrow(baseline %>% filter(PREVSTRK == 0)),
    sum((baseline %>% filter(PREVSTRK == 0))$SYSBP > 250, na.rm = TRUE),
    sum((baseline %>% filter(PREVSTRK == 0))$TOTCHOL > 600, na.rm = TRUE),
    nrow(analytic),
    analytic %>%
      filter(SEX == 1) %>%
      summarise(txt = paste0(n(), " (", sum(STROKE_10yr), ", ",
                             round(mean(STROKE_10yr) * 100, 2), "%)")) %>%
      pull(txt),
    analytic %>%
      filter(SEX == 2) %>%
      summarise(txt = paste0(n(), " (", sum(STROKE_10yr), ", ",
                             round(mean(STROKE_10yr) * 100, 2), "%)")) %>%
      pull(txt)
  )
)

table_a
knitr::kable(table_a, col.names = c("Cohort Summary", "Value"))

# make table 1 style descriptive summary by sex

library(gtsummary)
library(dplyr)

table1_data <- analytic %>%
  mutate(
    sex = factor(SEX, levels = c(1, 2), labels = c("Men", "Women")),
    diabetes = factor(DIABETES, levels = c(0, 1), labels = c("No", "Yes")),
    current_smoker = factor(CURSMOKE, levels = c(0, 1), labels = c("No", "Yes")),
    antihyp_med = factor(BPMEDS, levels = c(0, 1), labels = c("No", "Yes")),
    prev_chd = factor(PREVCHD, levels = c(0, 1), labels = c("No", "Yes")),
    stroke_10yr = factor(STROKE_10yr, levels = c(0, 1), labels = c("No", "Yes"))
  ) %>%
  select(
    sex,
    AGE,
    SYSBP,
    TOTCHOL,
    BMI,
    diabetes,
    current_smoker,
    antihyp_med,
    prev_chd,
    stroke_10yr
  )

table1 <- table1_data %>%
  tbl_summary(
    by = sex,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_dichotomous() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 1,
      all_dichotomous() ~ c(0, 1)
    ),
    type = list(
      diabetes ~ "dichotomous",
      current_smoker ~ "dichotomous",
      antihyp_med ~ "dichotomous",
      prev_chd ~ "dichotomous",
      stroke_10yr ~ "dichotomous"
    ),
    value = list(
      diabetes ~ "Yes",
      current_smoker ~ "Yes",
      antihyp_med ~ "Yes",
      prev_chd ~ "Yes",
      stroke_10yr ~ "Yes"
    ),
    label = list(
      AGE ~ "Age (years)",
      SYSBP ~ "Systolic BP (mmHg)",
      TOTCHOL ~ "Total cholesterol (mg/dL)",
      BMI ~ "BMI (kg/m^2)",
      diabetes ~ "Diabetes (yes)",
      current_smoker ~ "Current smoker",
      antihyp_med ~ "Blood Pressure Medication (yes)",
      prev_chd ~ "Prevalent CHD (yes)",
      stroke_10yr ~ "10-year stroke event"
    ),
    missing = "no"
  ) %>%
  add_overall(last = FALSE) %>%
  modify_header(label ~ "**Characteristic**") %>%
  bold_labels()

table1

#added for survival ready follow-up time and sex stratified datasets

#follow-up and event summary by sex
analytic %>%
  mutate(Sex = ifelse(SEX == 1, "Men", "Women")) %>%
  group_by(Sex) %>%
  summarise(
    N = n(),
    N_stroke_10yr = sum(STROKE_10yr, na.rm = TRUE),
    N_censored_before_10yr = sum(STROKE_10yr == 0 & FU_10yr < days_10yr, na.rm = TRUE),
    N_censored_at_10yr = sum(STROKE_10yr == 0 & FU_10yr >= days_10yr, na.rm = TRUE),
    Mean_followup_days = round(mean(FU_10yr, na.rm = TRUE), 1),
    Median_followup_days = round(median(FU_10yr, na.rm = TRUE), 1)
  )

#create sex stratified datasets for later Kaplan-Meier and Cox models
analytic_men <- analytic %>% filter(SEX == 1)
analytic_women <- analytic %>% filter(SEX == 2)

#check event counts in sex stratified datasets
nrow(analytic_men)
sum(analytic_men$STROKE_10yr, na.rm = TRUE)

nrow(analytic_women)
sum(analytic_women$STROKE_10yr, na.rm = TRUE)
```

# NOTE: TOTCHOL (52 missing), BMI (17 missing), and BPMEDS (60 missing) are
# retained in the analytic dataset at this stage. Flagged extreme SYSBP and
# TOTCHOL values are also retained per investigator guidance. If any of these
# variables are included in the final model, additional participants will be
# excluded only for missingness and the exclusion table above will be updated
# accordingly.

## Section 4: Descriptive Variable Assessment and Model Preparation

```{r}
library(ggplot2)

#summarize change over time in sysbp and diabetes

change_over_time <- data %>%
  mutate(
    sex = ifelse(SEX == 1, "men", "women"),
    period = factor(PERIOD, levels = c(1, 2, 3))
  ) %>%
  group_by(sex, period) %>%
  summarise(
    n_obs = n(),
    sysbp_mean = round(mean(SYSBP, na.rm = TRUE), 1),
    sysbp_sd = round(sd(SYSBP, na.rm = TRUE), 1),
    diabetes_n = sum(DIABETES == 1, na.rm = TRUE),
    diabetes_pct = round(mean(DIABETES == 1, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

change_over_time

change_over_time %>%
  ggplot(aes(x = period, y = sysbp_mean, group = sex, linetype = sex, shape = sex)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "mean systolic blood pressure across examination periods",
    x = "examination period",
    y = "mean systolic blood pressure (mmHg)"
  ) +
  theme_minimal()

change_over_time %>%
  ggplot(aes(x = period, y = diabetes_pct, group = sex, linetype = sex, shape = sex)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "diabetes prevalence across examination periods",
    x = "examination period",
    y = "diabetes prevalence (%)"
  ) +
  theme_minimal()

#define variable sets used below

cont_vars <- c("AGE", "SYSBP", "TOTCHOL", "BMI")
bin_vars <- c("CURSMOKE", "DIABETES", "BPMEDS", "PREVCHD")

#added for categorized systolic blood pressure and model completeness checks

#categorize systolic blood pressure for descriptive summaries and later Kaplan-Meier curves
analytic <- analytic %>%
  mutate(
    SYSBP_CAT = case_when(
      SYSBP < 120 ~ "<120",
      SYSBP >= 120 & SYSBP < 140 ~ "120-139",
      SYSBP >= 140 & SYSBP < 160 ~ "140-159",
      SYSBP >= 160 ~ ">=160",
      TRUE ~ NA_character_
    ),
    SYSBP_CAT = factor(SYSBP_CAT, levels = c("<120", "120-139", "140-159", ">=160"))
  )

#check distribution of categorized systolic blood pressure by sex
analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex, SYSBP_CAT) %>%
  summarise(n = n(), .groups = "drop")

#check completeness for core and candidate model variables
core_vars <- c("AGE", "SYSBP", "DIABETES")
candidate_vars <- c("CURSMOKE", "BPMEDS", "PREVCHD", "TOTCHOL", "BMI")
full_model_vars <- c(core_vars, candidate_vars)

model_completeness <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex) %>%
  summarise(
    N_total = n(),
    N_core_complete = sum(complete.cases(across(all_of(core_vars)))),
    N_full_candidate_complete = sum(complete.cases(across(all_of(full_model_vars)))),
    N_add_cursmoke = sum(complete.cases(across(all_of(c(core_vars, "CURSMOKE"))))),
    N_add_bpmeds = sum(complete.cases(across(all_of(c(core_vars, "BPMEDS"))))),
    N_add_prevchd = sum(complete.cases(across(all_of(c(core_vars, "PREVCHD"))))),
    N_add_totchol = sum(complete.cases(across(all_of(c(core_vars, "TOTCHOL"))))),
    N_add_bmi = sum(complete.cases(across(all_of(c(core_vars, "BMI"))))),
    .groups = "drop"
  )

model_completeness

#more detailed summaries of continuous model variables by sex
cont_distribution_summary <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  select(sex, all_of(cont_vars)) %>%
  pivot_longer(cols = all_of(cont_vars), names_to = "variable", values_to = "value") %>%
  group_by(sex, variable) %>%
  summarise(
    n_nonmissing = sum(!is.na(value)),
    mean = round(mean(value, na.rm = TRUE), 1),
    sd = round(sd(value, na.rm = TRUE), 1),
    median = round(median(value, na.rm = TRUE), 1),
    q1 = round(quantile(value, 0.25, na.rm = TRUE), 1),
    q3 = round(quantile(value, 0.75, na.rm = TRUE), 1),
    min = round(min(value, na.rm = TRUE), 1),
    max = round(max(value, na.rm = TRUE), 1),
    .groups = "drop"
  )

cont_distribution_summary

#look at univariate distributions by sex for continuous variables going into the models
#this is the main visual check for whether the continuous variables appear skewed

analytic %>%
  select(all_of(cont_vars), SEX) %>%
  pivot_longer(all_of(cont_vars), names_to = "variable", values_to = "value") %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  ggplot(aes(x = value, fill = sex)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  facet_wrap(~ variable, scales = "free") +
  labs(
    title = "distribution of continuous variables by sex",
    x = NULL,
    y = "count",
    fill = "sex"
  ) +
  theme_minimal()

#check sysbp and bpmeds relationship

sysbp_bpmeds_summary <- analytic %>%
  mutate(
    sex = ifelse(SEX == 1, "men", "women"),
    bpmeds_cat = case_when(
      BPMEDS == 0 ~ "no",
      BPMEDS == 1 ~ "yes",
      TRUE ~ "missing"
    ),
    bpmeds_cat = factor(bpmeds_cat, levels = c("no", "yes", "missing"))
  ) %>%
  group_by(sex, bpmeds_cat) %>%
  summarise(
    n = n(),
    sysbp_mean = round(mean(SYSBP, na.rm = TRUE), 1),
    sysbp_sd = round(sd(SYSBP, na.rm = TRUE), 1),
    .groups = "drop"
  )

sysbp_bpmeds_summary

sysbp_bpmeds_corr <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex) %>%
  summarise(
    corr_sysbp_bpmeds = round(cor(SYSBP, BPMEDS, use = "complete.obs"), 3),
    .groups = "drop"
  )

sysbp_bpmeds_corr

sysbp_bpmeds_test <- analytic %>%
  filter(!is.na(BPMEDS)) %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex) %>%
  summarise(
    p_value = wilcox.test(SYSBP ~ BPMEDS)$p.value,
    .groups = "drop"
  )

sysbp_bpmeds_test

analytic %>%
  mutate(
    sex = ifelse(SEX == 1, "men", "women"),
    bpmeds_cat = case_when(
      BPMEDS == 0 ~ "no",
      BPMEDS == 1 ~ "yes",
      TRUE ~ "missing"
    ),
    bpmeds_cat = factor(bpmeds_cat, levels = c("no", "yes", "missing"))
  ) %>%
  ggplot(aes(x = bpmeds_cat, y = SYSBP, fill = sex)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~ sex) +
  labs(
    title = "systolic blood pressure by blood pressure medication status",
    x = "blood pressure medication status",
    y = "systolic blood pressure (mmHg)",
    fill = "sex"
  ) +
  theme_minimal()

#check correlations among continuous candidate predictors

cont_corr <- analytic %>%
  select(all_of(cont_vars)) %>%
  cor(use = "pairwise.complete.obs")

round(cont_corr, 3)

#check associations between continuous and binary candidate predictors

cont_bin_assoc <- expand_grid(cont_var = cont_vars, bin_var = bin_vars) %>%
  rowwise() %>%
  mutate(
    n_complete = sum(!is.na(analytic[[cont_var]]) & !is.na(analytic[[bin_var]])),
    corr = round(cor(analytic[[cont_var]], analytic[[bin_var]], use = "complete.obs"), 3),
    p_value = {
      df_use <- analytic %>%
        select(all_of(c(cont_var, bin_var))) %>%
        drop_na()
      wilcox.test(df_use[[cont_var]] ~ df_use[[bin_var]])$p.value
    },
    mean_if_no = {
      df_use <- analytic %>%
        select(all_of(c(cont_var, bin_var))) %>%
        drop_na()
      round(mean(df_use[[cont_var]][df_use[[bin_var]] == 0], na.rm = TRUE), 1)
    },
    mean_if_yes = {
      df_use <- analytic %>%
        select(all_of(c(cont_var, bin_var))) %>%
        drop_na()
      round(mean(df_use[[cont_var]][df_use[[bin_var]] == 1], na.rm = TRUE), 1)
    }
  ) %>%
  ungroup()

cont_bin_assoc

#check associations among binary candidate predictors

binary_assoc <- expand_grid(var1 = bin_vars, var2 = bin_vars) %>%
  filter(var1 < var2) %>%
  rowwise() %>%
  mutate(
    n_complete = sum(!is.na(analytic[[var1]]) & !is.na(analytic[[var2]])),
    p_value = {
      keep <- !is.na(analytic[[var1]]) & !is.na(analytic[[var2]])
      tbl <- table(analytic[[var1]][keep], analytic[[var2]][keep])
      chisq_obj <- suppressWarnings(chisq.test(tbl, correct = FALSE))
      
      if (any(chisq_obj$expected < 5)) {
        fisher.test(tbl)$p.value
      } else {
        chisq_obj$p.value
      }
    },
    test_used = {
      keep <- !is.na(analytic[[var1]]) & !is.na(analytic[[var2]])
      tbl <- table(analytic[[var1]][keep], analytic[[var2]][keep])
      chisq_obj <- suppressWarnings(chisq.test(tbl, correct = FALSE))
      
      if (any(chisq_obj$expected < 5)) {
        "fisher"
      } else {
        "chi-square"
      }
    }
  ) %>%
  ungroup()

binary_assoc

#do bivariate analysis by sex for continuous variables

#added for more detailed bivariate summaries of continuous variables by sex

cont_results_detailed <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_split(sex) %>%
  map_dfr(function(df_sex) {
    
    sex_label <- unique(df_sex$sex)
    
    map_dfr(cont_vars, function(v) {
      df_use <- df_sex %>% filter(!is.na(.data[[v]]))
      
      tibble(
        sex = sex_label,
        variable = v,
        stroke_mean_sd = df_use %>%
          filter(STROKE_10yr == 1) %>%
          summarise(txt = sprintf("%.1f (%.1f)",
                                  mean(.data[[v]], na.rm = TRUE),
                                  sd(.data[[v]], na.rm = TRUE))) %>%
          pull(txt),
        no_stroke_mean_sd = df_use %>%
          filter(STROKE_10yr == 0) %>%
          summarise(txt = sprintf("%.1f (%.1f)",
                                  mean(.data[[v]], na.rm = TRUE),
                                  sd(.data[[v]], na.rm = TRUE))) %>%
          pull(txt),
        stroke_median_iqr = df_use %>%
          filter(STROKE_10yr == 1) %>%
          summarise(txt = sprintf("%.1f (%.1f, %.1f)",
                                  median(.data[[v]], na.rm = TRUE),
                                  quantile(.data[[v]], 0.25, na.rm = TRUE),
                                  quantile(.data[[v]], 0.75, na.rm = TRUE))) %>%
          pull(txt),
        no_stroke_median_iqr = df_use %>%
          filter(STROKE_10yr == 0) %>%
          summarise(txt = sprintf("%.1f (%.1f, %.1f)",
                                  median(.data[[v]], na.rm = TRUE),
                                  quantile(.data[[v]], 0.25, na.rm = TRUE),
                                  quantile(.data[[v]], 0.75, na.rm = TRUE))) %>%
          pull(txt),
        p_value = round(wilcox.test(df_use[[v]] ~ df_use$STROKE_10yr)$p.value, 4)
      )
    })
  })

cont_results_detailed

#do bivariate analysis by sex for binary variables
bin_results <- analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_split(sex) %>%
  map_dfr(function(df_sex) {
    
    sex_label <- unique(df_sex$sex)
    
    map_dfr(bin_vars, function(v) {
      df_use <- df_sex[!is.na(df_sex[[v]]), ]
      tbl <- table(df_use[[v]], df_use$STROKE_10yr)
      
      test <- if (any(tbl < 5)) {
        fisher.test(tbl)
      } else {
        suppressWarnings(chisq.test(tbl))
      }
      
      stroke_pct_no <- df_use %>%
        filter(.data[[v]] == 0) %>%
        summarise(stroke_pct = round(mean(STROKE_10yr) * 100, 1)) %>%
        pull(stroke_pct)
      
      stroke_pct_yes <- df_use %>%
        filter(.data[[v]] == 1) %>%
        summarise(stroke_pct = round(mean(STROKE_10yr) * 100, 1)) %>%
        pull(stroke_pct)
      
      tibble(
        sex = sex_label,
        variable = v,
        stroke_pct_no = stroke_pct_no,
        stroke_pct_yes = stroke_pct_yes,
        p_value = round(test$p.value, 4),
        test_used = if (any(tbl < 5)) "fisher" else "chi-square"
      )
    })
  })

bin_results
```



## Section 5: Kaplan-Meier Survival Analysis

```{r}
library(survival)
library(survminer)

#added for Kaplan-Meier survival analysis using 10-year observed follow-up

#re-create sex stratified datasets here so they include SYSBP_CAT created in Section 4
analytic_men <- analytic %>% filter(SEX == 1)
analytic_women <- analytic %>% filter(SEX == 2)

#create cleaner labeled datasets for Kaplan-Meier plots
analytic_km <- analytic %>%
  mutate(
    sex_label = factor(SEX, levels = c(1, 2), labels = c("Men", "Women"))
  )

analytic_men_km <- analytic_men %>%
  mutate(
    diabetes_label = factor(DIABETES, levels = c(0, 1), labels = c("No diabetes", "Diabetes")),
    smoke_label = factor(CURSMOKE, levels = c(0, 1), labels = c("Not current", "Current")),
    prevchd_label = factor(PREVCHD, levels = c(0, 1), labels = c("No CHD", "CHD"))
  )

analytic_women_km <- analytic_women %>%
  mutate(
    diabetes_label = factor(DIABETES, levels = c(0, 1), labels = c("No diabetes", "Diabetes")),
    smoke_label = factor(CURSMOKE, levels = c(0, 1), labels = c("Not current", "Current")),
    prevchd_label = factor(PREVCHD, levels = c(0, 1), labels = c("No CHD", "CHD"))
  )

#BPMEDS has missing values, so create complete-case datasets for those plots/tests
analytic_men_bpmeds <- analytic_men %>%
  filter(!is.na(BPMEDS)) %>%
  mutate(bpmeds_label = factor(BPMEDS, levels = c(0, 1), labels = c("No BP meds", "BP meds")))

analytic_women_bpmeds <- analytic_women %>%
  filter(!is.na(BPMEDS)) %>%
  mutate(bpmeds_label = factor(BPMEDS, levels = c(0, 1), labels = c("No BP meds", "BP meds")))

#summarize follow-up time by sex for survival section
analytic %>%
  mutate(sex = ifelse(SEX == 1, "men", "women")) %>%
  group_by(sex) %>%
  summarise(
    n = n(),
    n_events = sum(STROKE_10yr, na.rm = TRUE),
    median_followup_days = round(median(FU_10yr, na.rm = TRUE), 1),
    q1_followup_days = round(quantile(FU_10yr, 0.25, na.rm = TRUE), 1),
    q3_followup_days = round(quantile(FU_10yr, 0.75, na.rm = TRUE), 1),
    .groups = "drop"
  )

#helper function for cleaner Kaplan-Meier plots
plot_km_event <- function(fit, data, title) {
  ggsurvplot(
    fit,
    data = data,
    fun = "event",
    conf.int = FALSE,
    censor = FALSE,
    risk.table = TRUE,
    risk.table.height = 0.22,
    risk.table.fontsize = 3.2,
    break.time.by = 730,
    xlim = c(0, days_10yr),
    xlab = "Follow-up time (days)",
    ylab = "Cumulative stroke probability",
    title = title,
    ggtheme = theme_minimal(base_size = 12),
    tables.theme = theme_cleantable(base_size = 10)
  )
}

#helper function for log-rank test summary
logrank_summary <- function(formula, data, group_var) {
  fit <- survdiff(formula, data = data)
  tibble(
    variable = group_var,
    chisq = round(fit$chisq, 3),
    df = length(fit$n) - 1,
    p_value = round(1 - pchisq(fit$chisq, df = length(fit$n) - 1), 4),
    n_total = sum(fit$n)
  )
}

#overall Kaplan-Meier curve by sex
km_sex <- survfit(Surv(FU_10yr, STROKE_10yr) ~ sex_label, data = analytic_km)
plot_km_event(km_sex, analytic_km, "Kaplan-Meier cumulative stroke probability by sex")

logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ sex_label, analytic_km, "Sex")

#Kaplan-Meier curves within men

#diabetes
km_men_diabetes <- survfit(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, data = analytic_men_km)
plot_km_event(km_men_diabetes, analytic_men_km, "Men: cumulative stroke probability by diabetes")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_men_km, "Men: diabetes")

#current smoking
km_men_smoke <- survfit(Surv(FU_10yr, STROKE_10yr) ~ smoke_label, data = analytic_men_km)
plot_km_event(km_men_smoke, analytic_men_km, "Men: cumulative stroke probability by current smoking")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label, analytic_men_km, "Men: current smoking")

#blood pressure medication use
km_men_bpmeds <- survfit(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label, data = analytic_men_bpmeds)
plot_km_event(km_men_bpmeds, analytic_men_bpmeds, "Men: cumulative stroke probability by BP medication use")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label, analytic_men_bpmeds, "Men: BP medication")

#prevalent chd
km_men_prevchd <- survfit(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label, data = analytic_men_km)
plot_km_event(km_men_prevchd, analytic_men_km, "Men: cumulative stroke probability by prevalent CHD")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label, analytic_men_km, "Men: prevalent CHD")

#categorized systolic blood pressure
km_men_sysbp <- survfit(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT, data = analytic_men)
plot_km_event(km_men_sysbp, analytic_men, "Men: cumulative stroke probability by systolic BP category")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT, analytic_men, "Men: systolic BP category")

#Kaplan-Meier curves within women

#diabetes
km_women_diabetes <- survfit(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, data = analytic_women_km)
plot_km_event(km_women_diabetes, analytic_women_km, "Women: cumulative stroke probability by diabetes")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_women_km, "Women: diabetes")

#current smoking
km_women_smoke <- survfit(Surv(FU_10yr, STROKE_10yr) ~ smoke_label, data = analytic_women_km)
plot_km_event(km_women_smoke, analytic_women_km, "Women: cumulative stroke probability by current smoking")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label, analytic_women_km, "Women: current smoking")

#blood pressure medication use
km_women_bpmeds <- survfit(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label, data = analytic_women_bpmeds)
plot_km_event(km_women_bpmeds, analytic_women_bpmeds, "Women: cumulative stroke probability by BP medication use")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label, analytic_women_bpmeds, "Women: BP medication")

#prevalent chd
km_women_prevchd <- survfit(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label, data = analytic_women_km)
plot_km_event(km_women_prevchd, analytic_women_km, "Women: cumulative stroke probability by prevalent CHD")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label, analytic_women_km, "Women: prevalent CHD")

#categorized systolic blood pressure
km_women_sysbp <- survfit(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT, data = analytic_women)
plot_km_event(km_women_sysbp, analytic_women, "Women: cumulative stroke probability by systolic BP category")
logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT, analytic_women, "Women: systolic BP category")


km_logrank_results <- bind_rows(
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ sex_label, analytic_km, "Sex"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_men_km, "Men: diabetes"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label, analytic_men_km, "Men: current smoking"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label, analytic_men_bpmeds, "Men: BP medication"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label, analytic_men_km, "Men: prevalent CHD"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT, analytic_men, "Men: systolic BP category"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_women_km, "Women: diabetes"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label, analytic_women_km, "Women: current smoking"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label, analytic_women_bpmeds, "Women: BP medication"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label, analytic_women_km, "Women: prevalent CHD"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT, analytic_women, "Women: systolic BP category")
)

km_logrank_results

#added to save all Kaplan-Meier plots into one PDF and print one combined log-rank summary table

#combine all log-rank results into one table
km_logrank_results <- bind_rows(
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ sex_label, analytic_km, "Sex"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_men_km, "Men: diabetes"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label, analytic_men_km, "Men: current smoking"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label, analytic_men_bpmeds, "Men: BP medication"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label, analytic_men_km, "Men: prevalent CHD"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT, analytic_men, "Men: systolic BP category"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_women_km, "Women: diabetes"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label, analytic_women_km, "Women: current smoking"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label, analytic_women_bpmeds, "Women: BP medication"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label, analytic_women_km, "Women: prevalent CHD"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT, analytic_women, "Women: systolic BP category")
)

km_logrank_results
knitr::kable(km_logrank_results, digits = 4)

#create plot objects first
p_km_sex <- plot_km_event(
  km_sex, analytic_km,
  "Kaplan-Meier cumulative stroke probability by sex"
)

p_km_men_diabetes <- plot_km_event(
  km_men_diabetes, analytic_men_km,
  "Men: cumulative stroke probability by diabetes"
)

p_km_men_smoke <- plot_km_event(
  km_men_smoke, analytic_men_km,
  "Men: cumulative stroke probability by current smoking"
)

p_km_men_bpmeds <- plot_km_event(
  km_men_bpmeds, analytic_men_bpmeds,
  "Men: cumulative stroke probability by BP medication use"
)

p_km_men_prevchd <- plot_km_event(
  km_men_prevchd, analytic_men_km,
  "Men: cumulative stroke probability by prevalent CHD"
)

p_km_men_sysbp <- plot_km_event(
  km_men_sysbp, analytic_men,
  "Men: cumulative stroke probability by systolic BP category"
)

p_km_women_diabetes <- plot_km_event(
  km_women_diabetes, analytic_women_km,
  "Women: cumulative stroke probability by diabetes"
)

p_km_women_smoke <- plot_km_event(
  km_women_smoke, analytic_women_km,
  "Women: cumulative stroke probability by current smoking"
)

p_km_women_bpmeds <- plot_km_event(
  km_women_bpmeds, analytic_women_bpmeds,
  "Women: cumulative stroke probability by BP medication use"
)

p_km_women_prevchd <- plot_km_event(
  km_women_prevchd, analytic_women_km,
  "Women: cumulative stroke probability by prevalent CHD"
)

p_km_women_sysbp <- plot_km_event(
  km_women_sysbp, analytic_women,
  "Women: cumulative stroke probability by systolic BP category"
)

#save all KM plots into one pdf
pdf(
  file = "C:/Users/kokeshem/Spring 2026/Advanced Statistical Methods/Project 3/km_plots_project3.pdf",
  width = 11,
  height = 8.5
)

print(p_km_sex)
print(p_km_men_diabetes)
print(p_km_men_smoke)
print(p_km_men_bpmeds)
print(p_km_men_prevchd)
print(p_km_men_sysbp)
print(p_km_women_diabetes)
print(p_km_women_smoke)
print(p_km_women_bpmeds)
print(p_km_women_prevchd)
print(p_km_women_sysbp)

dev.off()
```

```{r}
## Section 6: Cox Proportional Hazards Models with LASSO Variable Selection
library(survival)
library(glmnet)
library(broom)

#added for sex-stratified Cox PH models with LASSO-based variable selection

#define core and candidate predictors for Cox modeling
core_vars <- c("AGE", "SYSBP", "DIABETES")
candidate_vars <- c("CURSMOKE", "BPMEDS", "PREVCHD", "TOTCHOL", "BMI")
all_model_vars <- c(core_vars, candidate_vars)

#restrict to complete cases for variables considered in the LASSO step
cox_men_data <- analytic_men %>%
  select(FU_10yr, STROKE_10yr, all_of(all_model_vars)) %>%
  drop_na()

cox_women_data <- analytic_women %>%
  select(FU_10yr, STROKE_10yr, all_of(all_model_vars)) %>%
  drop_na()

#check sample sizes after complete-case restriction for LASSO modeling
tibble(
  sex = c("men", "women"),
  n_complete_case = c(nrow(cox_men_data), nrow(cox_women_data)),
  n_events = c(sum(cox_men_data$STROKE_10yr), sum(cox_women_data$STROKE_10yr))
)

#create survival responses for glmnet
y_men <- with(cox_men_data, Surv(FU_10yr, STROKE_10yr))
y_women <- with(cox_women_data, Surv(FU_10yr, STROKE_10yr))

#create design matrices for glmnet (no intercept column)
x_men <- model.matrix(
  ~ AGE + SYSBP + DIABETES + CURSMOKE + BPMEDS + PREVCHD + TOTCHOL + BMI,
  data = cox_men_data
)[, -1]

x_women <- model.matrix(
  ~ AGE + SYSBP + DIABETES + CURSMOKE + BPMEDS + PREVCHD + TOTCHOL + BMI,
  data = cox_women_data
)[, -1]

#penalty factors:
#force AGE, SYSBP, and DIABETES into the model by setting penalty = 0
penalty_men <- ifelse(colnames(x_men) %in% core_vars, 0, 1)
penalty_women <- ifelse(colnames(x_women) %in% core_vars, 0, 1)

#cross-validated LASSO Cox models by sex
set.seed(6624)

cv_lasso_men <- cv.glmnet(
  x = x_men,
  y = y_men,
  family = "cox",
  alpha = 1,
  penalty.factor = penalty_men,
  nfolds = 10,
  standardize = TRUE
)

cv_lasso_women <- cv.glmnet(
  x = x_women,
  y = y_women,
  family = "cox",
  alpha = 1,
  penalty.factor = penalty_women,
  nfolds = 10,
  standardize = TRUE
)

#plot cross-validation curves
plot(cv_lasso_men)
plot(cv_lasso_women)

#extract selected coefficients at lambda.1se for a more parsimonious model
coef_men_1se <- coef(cv_lasso_men, s = "lambda.1se")
coef_women_1se <- coef(cv_lasso_women, s = "lambda.1se")

selected_men <- rownames(coef_men_1se)[as.vector(coef_men_1se != 0)]
selected_women <- rownames(coef_women_1se)[as.vector(coef_women_1se != 0)]

selected_men
selected_women

#clean selected variable names for use in coxph refit
#(model.matrix names are the same as the original variables here because all predictors are numeric/binary)
final_vars_men <- selected_men
final_vars_women <- selected_women

#summarize selected predictors
lasso_selected_summary <- tibble(
  sex = c("men", "women"),
  lambda_min = c(cv_lasso_men$lambda.min, cv_lasso_women$lambda.min),
  lambda_1se = c(cv_lasso_men$lambda.1se, cv_lasso_women$lambda.1se),
  selected_predictors_1se = c(
    paste(final_vars_men, collapse = ", "),
    paste(final_vars_women, collapse = ", ")
  )
)

lasso_selected_summary

#refit standard coxph models with predictors selected by LASSO
formula_men <- as.formula(
  paste("Surv(FU_10yr, STROKE_10yr) ~", paste(final_vars_men, collapse = " + "))
)

formula_women <- as.formula(
  paste("Surv(FU_10yr, STROKE_10yr) ~", paste(final_vars_women, collapse = " + "))
)

cox_final_men <- coxph(formula_men, data = cox_men_data)
cox_final_women <- coxph(formula_women, data = cox_women_data)

summary(cox_final_men)
summary(cox_final_women)

#tidy final model results with hazard ratios and 95% confidence intervals
cox_final_men_tidy <- tidy(cox_final_men, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(sex = "men")

cox_final_women_tidy <- tidy(cox_final_women, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(sex = "women")

cox_final_results <- bind_rows(cox_final_men_tidy, cox_final_women_tidy) %>%
  select(sex, term, estimate, conf.low, conf.high, p.value)

cox_final_results

knitr::kable(cox_final_results, digits = 3)

#optional: univariable Cox models by sex for descriptive comparison with LASSO-selected final models
run_univ_cox <- function(data, vars, sex_label) {
  map_dfr(vars, function(v) {
    fit <- coxph(as.formula(paste("Surv(FU_10yr, STROKE_10yr) ~", v)), data = data)
    tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
      mutate(sex = sex_label, predictor = v) %>%
      select(sex, predictor, estimate, conf.low, conf.high, p.value)
  })
}

cox_univ_results <- bind_rows(
  run_univ_cox(cox_men_data, all_model_vars, "men"),
  run_univ_cox(cox_women_data, all_model_vars, "women")
)

cox_univ_results

#save selected variable lists for later prediction section
final_vars_men
final_vars_women


#added for sensitivity analyses of the LASSO variable selection approach

#extract selected coefficients at lambda.min
coef_men_min <- coef(cv_lasso_men, s = "lambda.min")
coef_women_min <- coef(cv_lasso_women, s = "lambda.min")

selected_men_min <- rownames(coef_men_min)[as.vector(coef_men_min != 0)]
selected_women_min <- rownames(coef_women_min)[as.vector(coef_women_min != 0)]

selected_men_min
selected_women_min

#elastic net sensitivity analysis (alpha = 0.5)
set.seed(6624)

cv_enet_men <- cv.glmnet(
  x = x_men,
  y = y_men,
  family = "cox",
  alpha = 0.5,
  penalty.factor = penalty_men,
  nfolds = 10,
  standardize = TRUE
)

cv_enet_women <- cv.glmnet(
  x = x_women,
  y = y_women,
  family = "cox",
  alpha = 0.5,
  penalty.factor = penalty_women,
  nfolds = 10,
  standardize = TRUE
)

plot(cv_enet_men)
plot(cv_enet_women)

coef_men_enet <- coef(cv_enet_men, s = "lambda.1se")
coef_women_enet <- coef(cv_enet_women, s = "lambda.1se")

selected_men_enet <- rownames(coef_men_enet)[as.vector(coef_men_enet != 0)]
selected_women_enet <- rownames(coef_women_enet)[as.vector(coef_women_enet != 0)]

selected_men_enet
selected_women_enet

lasso_sensitivity_summary <- tibble(
  sex = c("men", "women"),
  lasso_1se = c(paste(selected_men, collapse = ", "),
                paste(selected_women, collapse = ", ")),
  lasso_min = c(paste(selected_men_min, collapse = ", "),
                paste(selected_women_min, collapse = ", ")),
  enet_1se = c(paste(selected_men_enet, collapse = ", "),
               paste(selected_women_enet, collapse = ", "))
)

lasso_sensitivity_summary
knitr::kable(lasso_sensitivity_summary)
```


```{r}
## Section 7: Proportional Hazards Diagnostics for Final Cox Models

library(survival)
library(survminer)
library(ggplot2)
library(dplyr)
library(tibble)
library(knitr)

#added for proportional hazards diagnostics of the final refit Cox models

#Schoenfeld residual tests for final models
ph_men <- cox.zph(cox_final_men)
ph_women <- cox.zph(cox_final_women)

ph_men
ph_women

#tidy Schoenfeld test results
ph_men_table <- tibble(
  sex = "Men",
  term = rownames(ph_men$table),
  chisq = ph_men$table[, "chisq"],
  df = ph_men$table[, "df"],
  p_value = ph_men$table[, "p"]
)

ph_women_table <- tibble(
  sex = "Women",
  term = rownames(ph_women$table),
  chisq = ph_women$table[, "chisq"],
  df = ph_women$table[, "df"],
  p_value = ph_women$table[, "p"]
)

ph_results <- bind_rows(ph_men_table, ph_women_table) %>%
  mutate(
    term = recode(
      term,
      AGE = "Age",
      SYSBP = "Systolic blood pressure",
      DIABETES = "Diabetes",
      GLOBAL = "Global test"
    )
  )

ph_results
kable(
  ph_results,
  digits = 4,
  col.names = c("Sex", "Term", "Chi-square", "DF", "P-value")
)

#men schoenfeld plots

p_men_age <- ggcoxzph(ph_men, var = "AGE", se = FALSE) +
  labs(
    title = "Men: Schoenfeld residual plot for age",
    subtitle = "Assessing proportional hazards for age in the final Cox model",
    x = "Follow-up time (days)",
    y = "Scaled Schoenfeld residuals for age"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

p_men_age

p_men_sysbp <- ggcoxzph(ph_men, var = "SYSBP", se = FALSE) +
  labs(
    title = "Men: Schoenfeld residual plot for systolic blood pressure",
    subtitle = "Assessing proportional hazards for systolic blood pressure in the final Cox model",
    x = "Follow-up time (days)",
    y = "Scaled Schoenfeld residuals for systolic blood pressure"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

p_men_sysbp

p_men_diabetes <- ggcoxzph(ph_men, var = "DIABETES", se = FALSE) +
  labs(
    title = "Men: Schoenfeld residual plot for diabetes",
    subtitle = "Assessing proportional hazards for diabetes in the final Cox model",
    x = "Follow-up time (days)",
    y = "Scaled Schoenfeld residuals for diabetes"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

p_men_diabetes

#women schoenfeld residual plots
p_women_age <- ggcoxzph(ph_women, var = "AGE", se = FALSE) +
  labs(
    title = "Women: Schoenfeld residual plot for age",
    subtitle = "Assessing proportional hazards for age in the final Cox model",
    x = "Follow-up time (days)",
    y = "Scaled Schoenfeld residuals for age"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

p_women_age

p_women_sysbp <- ggcoxzph(ph_women, var = "SYSBP", se = FALSE) +
  labs(
    title = "Women: Schoenfeld residual plot for systolic blood pressure",
    subtitle = "Assessing proportional hazards for systolic blood pressure in the final Cox model",
    x = "Follow-up time (days)",
    y = "Scaled Schoenfeld residuals for systolic blood pressure"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

p_women_sysbp

p_women_diabetes <- ggcoxzph(ph_women, var = "DIABETES", se = FALSE) +
  labs(
    title = "Women: Schoenfeld residual plot for diabetes",
    subtitle = "Assessing proportional hazards for diabetes in the final Cox model",
    x = "Follow-up time (days)",
    y = "Scaled Schoenfeld residuals for diabetes"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

p_women_diabetes

#log-log survival plots for diabetes

surv_men_diabetes_ll <- survfit(Surv(FU_10yr, STROKE_10yr) ~ DIABETES, data = cox_men_data)

p_ll_men <- ggsurvplot(
  surv_men_diabetes_ll,
  data = cox_men_data,
  conf.int = FALSE,
  censor = FALSE,
  fun = "cloglog",
  legend.labs = c("No diabetes", "Diabetes"),
  legend.title = "Diabetes status",
  xlab = "Follow-up time (days)",
  ylab = "log(-log(Survival probability))",
  title = "Men: log-minus-log survival plot for diabetes",
  subtitle = "Approximately parallel curves support the proportional hazards assumption",
  ggtheme = theme_minimal(base_size = 13),
  break.time.by = 730
)

print(p_ll_men)

surv_women_diabetes_ll <- survfit(Surv(FU_10yr, STROKE_10yr) ~ DIABETES, data = cox_women_data)

p_ll_women <- ggsurvplot(
  surv_women_diabetes_ll,
  data = cox_women_data,
  conf.int = FALSE,
  censor = FALSE,
  fun = "cloglog",
  legend.labs = c("No diabetes", "Diabetes"),
  legend.title = "Diabetes status",
  xlab = "Follow-up time (days)",
  ylab = "log(-log(Survival probability))",
  title = "Women: log-minus-log survival plot for diabetes",
  subtitle = "Approximately parallel curves support the proportional hazards assumption",
  ggtheme = theme_minimal(base_size = 13),
  break.time.by = 730
)

print(p_ll_women)


```

