---
title: "Baseline Risk Factors and 10-Year Probability of Incident Stroke: A Sex-Stratified Cox Proportional Hazards Analysis of the Framingham Heart Study"
author: "Emily Kokesh"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
---

<style type="text/css">
/* Screen: comfortable reading */
body { font-family: "Source Sans Pro", "Helvetica Neue", Arial, sans-serif; }

/* Print / save-as-PDF: conform to project guidelines
   (12-pt font, double-spaced body text, 1-inch margins).
   Tables, figures, and their captions are excluded from the
   double-spacing rule since the 6-page limit excludes them. */
@media print {
  @page { size: letter; margin: 1in; }
  body {
    font-size: 12pt;
    line-height: 2.0;
  }
  h1, h2, h3, h4 { line-height: 1.2; }
  .gt_table, .gt_table *,
  figure, figcaption,
  .figure, .figure * { line-height: 1.2 !important; }
  /* Don't break inside tables and figures */
  table, figure, .figure { page-break-inside: avoid; }
  /* Hide the floating TOC when printing */
  #TOC, .tocify { display: none !important; }
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 150
)

library(tidyverse)
library(survival)
library(survminer)
library(glmnet)
library(broom)
library(gt)
library(knitr)
library(DiagrammeR)
library(glue)
library(tibble)
library(scales)

#data paths
data_path <- "C:/Users/kokeshem/Spring 2026/Advanced Statistical Methods/Project 3/frmgham2.csv"

#shared gt theme
style_gt <- function(gt_tbl, caption_md = NULL) {
  out <- gt_tbl %>%
    opt_table_font(font = list(google_font("Source Sans Pro"), default_fonts())) %>%
    tab_options(
      table.font.size = px(13),
      table.border.top.style    = "solid",
      table.border.top.width    = px(2),
      table.border.bottom.style = "solid",
      table.border.bottom.width = px(2),
      heading.border.bottom.style = "solid",
      heading.border.bottom.width = px(1.5),
      column_labels.border.top.style    = "solid",
      column_labels.border.top.width    = px(1.5),
      column_labels.border.bottom.style = "solid",
      column_labels.border.bottom.width = px(1),
      column_labels.font.weight = "bold",
      row_group.font.weight     = "bold",
      heading.align             = "left",
      table.align               = "center",
      source_notes.font.size    = px(11),
      source_notes.padding      = px(6),
      data_row.padding          = px(5)
    ) %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_column_labels()
    )
  if (!is.null(caption_md)) {
    out <- out %>% tab_header(title = md(caption_md))
  }
  out
}
```

```{r run_analysis, include=FALSE, cache=FALSE}
#Analysis Pipeline

#cohort and outcome
data     <- read.csv(data_path)
baseline <- data %>% filter(PERIOD == 1)
analytic <- baseline %>% filter(PREVSTRK == 0)

days_10yr <- 365.25 * 10

analytic <- analytic %>%
  mutate(
    STROKE_10yr   = ifelse(STROKE == 1 & TIMESTRK <= days_10yr, 1, 0),
    FU_10yr       = pmin(TIMESTRK, days_10yr),
    TIMESTRK_10yr = ifelse(TIMESTRK > days_10yr, days_10yr, TIMESTRK),
    SYSBP_CAT = case_when(
      SYSBP <  160 ~ "<160",
      SYSBP >= 160 ~ ">=160",
      TRUE         ~ NA_character_
    ) %>% factor(levels = c("<160", ">=160"))
  )

analytic_men   <- analytic %>% filter(SEX == 1)
analytic_women <- analytic %>% filter(SEX == 2)

#lasso, coxph
core_vars      <- c("AGE", "SYSBP", "DIABETES")
candidate_vars <- c("CURSMOKE", "BPMEDS", "PREVCHD", "TOTCHOL", "BMI")
all_model_vars <- c(core_vars, candidate_vars)

cox_men_data <- analytic_men %>%
  select(FU_10yr, STROKE_10yr, all_of(all_model_vars)) %>%
  drop_na()

cox_women_data <- analytic_women %>%
  select(FU_10yr, STROKE_10yr, all_of(all_model_vars)) %>%
  drop_na()

y_men   <- with(cox_men_data,   Surv(FU_10yr, STROKE_10yr))
y_women <- with(cox_women_data, Surv(FU_10yr, STROKE_10yr))

x_men <- model.matrix(
  ~ AGE + SYSBP + DIABETES + CURSMOKE + BPMEDS + PREVCHD + TOTCHOL + BMI,
  data = cox_men_data)[, -1]
x_women <- model.matrix(
  ~ AGE + SYSBP + DIABETES + CURSMOKE + BPMEDS + PREVCHD + TOTCHOL + BMI,
  data = cox_women_data)[, -1]

penalty_men   <- ifelse(colnames(x_men)   %in% core_vars, 0, 1)
penalty_women <- ifelse(colnames(x_women) %in% core_vars, 0, 1)

set.seed(6624)
cv_lasso_men <- cv.glmnet(
  x_men, y_men, family = "cox", alpha = 1,
  penalty.factor = penalty_men, nfolds = 5, standardize = TRUE
)
cv_lasso_women <- cv.glmnet(
  x_women, y_women, family = "cox", alpha = 1,
  penalty.factor = penalty_women, nfolds = 5, standardize = TRUE
)

coef_men_min   <- coef(cv_lasso_men,   s = "lambda.min")
coef_women_min <- coef(cv_lasso_women, s = "lambda.min")

selected_men_min   <- rownames(coef_men_min)[as.vector(coef_men_min != 0)]
selected_women_min <- rownames(coef_women_min)[as.vector(coef_women_min != 0)]

final_vars_union <- sort(unique(c(selected_men_min, selected_women_min)))
final_vars_men   <- final_vars_union
final_vars_women <- final_vars_union

refit_men_data <- analytic_men %>%
  select(FU_10yr, STROKE_10yr, all_of(final_vars_men)) %>% drop_na()
refit_women_data <- analytic_women %>%
  select(FU_10yr, STROKE_10yr, all_of(final_vars_women)) %>% drop_na()

formula_men   <- as.formula(paste("Surv(FU_10yr, STROKE_10yr) ~",
                                  paste(final_vars_men, collapse = " + ")))
formula_women <- as.formula(paste("Surv(FU_10yr, STROKE_10yr) ~",
                                  paste(final_vars_women, collapse = " + ")))

cox_final_men   <- coxph(formula_men,   data = refit_men_data)
cox_final_women <- coxph(formula_women, data = refit_women_data)

#table 3 labels
term_labels <- c(
  "AGE (per 1 year)"       = "Age (per 1 year)",
  "SYSBP (per 10 mmHg)"    = "Systolic BP (per 10 mmHg)",
  "DIABETES"               = "Diabetes",
  "CURSMOKE"               = "Current smoker",
  "BPMEDS"                 = "Antihypertensive medication",
  "PREVCHD"                = "Prevalent CHD",
  "TOTCHOL (per 10 mg/dL)" = "Total cholesterol (per 10 mg/dL)",
  "BMI (per 1 kg/m^2)"     = "BMI (per 1 kg/m\u00B2)"
)

rescale_hr <- function(tidy_df) {
  tidy_df %>%
    mutate(
      scale_factor = case_when(
        term == "SYSBP" ~ 10,
        term == "TOTCHOL" ~ 10,
        TRUE ~ 1
      ),
      estimate  = exp(estimate  * scale_factor),
      conf.low  = exp(conf.low  * scale_factor),
      conf.high = exp(conf.high * scale_factor),
      term = case_when(
        term == "SYSBP"   ~ "SYSBP (per 10 mmHg)",
        term == "TOTCHOL" ~ "TOTCHOL (per 10 mg/dL)",
        term == "AGE"     ~ "AGE (per 1 year)",
        term == "BMI"     ~ "BMI (per 1 kg/m^2)",
        TRUE ~ term
      )
    )
}

cox_final_men_tidy <- tidy(cox_final_men, conf.int = TRUE) %>%
  rescale_hr() %>% mutate(sex = "men")
cox_final_women_tidy <- tidy(cox_final_women, conf.int = TRUE) %>%
  rescale_hr() %>% mutate(sex = "women")

cox_final_results <- bind_rows(cox_final_men_tidy, cox_final_women_tidy) %>%
  select(sex, term, estimate, conf.low, conf.high, p.value)

#cohort flow diagram
n_period1             <- nrow(baseline)
n_prevstroke_excluded <- sum(baseline$PREVSTRK == 1, na.rm = TRUE)
n_analytic            <- nrow(analytic)
n_men                 <- nrow(analytic_men)
n_women               <- nrow(analytic_women)
n_lasso_men           <- nrow(cox_men_data)
n_lasso_women         <- nrow(cox_women_data)
n_refit_men           <- nrow(refit_men_data)
n_refit_women         <- nrow(refit_women_data)

cohort_flow_code <- glue("
digraph cohort_flow {{
  graph [layout = dot, rankdir = TB, bgcolor = 'transparent']
  node [shape = box, style = 'rounded,filled', fontname = 'Source Sans Pro', fontsize = 11,
        fillcolor = '#F4F7FA', color = '#36526B', penwidth = 1.2]
  edge [fontname = 'Source Sans Pro', fontsize = 10, color = '#36526B']

  A [label = 'Period 1 baseline sample\\nN = {n_period1}']
  B [label = 'Excluded: prevalent stroke\\nN = {n_prevstroke_excluded}',
     fillcolor = '#FDECEA', color = '#B23A48']
  C [label = 'Baseline-eligible analytic cohort\\nN = {n_analytic}']

  D [label = 'Men\\nAnalytic sample: N = {n_men}\\nLASSO sample: N = {n_lasso_men}\\nFinal refit sample: N = {n_refit_men}']
  E [label = 'Women\\nAnalytic sample: N = {n_women}\\nLASSO sample: N = {n_lasso_women}\\nFinal refit sample: N = {n_refit_women}']

  A -> C
  A -> B [style = dashed]
  C -> D
  C -> E
}}
")

#TABLE 1: baseline + follow-up by sex
fmt_mean_sd <- function(x, d = 1) {
  x <- x[!is.na(x)]; if (!length(x)) return("\u2014")
  sprintf(paste0("%.", d, "f (%.", d, "f)"), mean(x), sd(x))
}
fmt_n_pct <- function(x) {
  nm <- x[!is.na(x)]; if (!length(nm)) return("\u2014")
  n <- sum(nm == 1); sprintf("%s (%.1f%%)", format(n, big.mark = ","), 100 * n / length(nm))
}
fmt_missing_n <- function(x) {
  nm <- sum(is.na(x)); if (!nm) NA_character_ else format(nm, big.mark = ",")
}
fmt_time_ms <- function(time, ind, v = 1, d = 1) {
  x <- time[ind == v & !is.na(ind) & !is.na(time)]; if (!length(x)) return("\u2014")
  sprintf(paste0("%.", d, "f (%.", d, "f)"), mean(x), sd(x))
}
fmt_km <- function(time, event, t = days_10yr, d = 1) {
  k <- !is.na(time) & !is.na(event); if (!sum(k)) return("\u2014")
  s <- summary(survfit(Surv(time[k], event[k]) ~ 1), times = t, extend = TRUE)$surv
  if (!length(s) || is.na(s)) return("\u2014")
  sprintf(paste0("%.", d, "f%%"), 100 * s)
}

t1 <- analytic %>%
  mutate(
    sex = factor(SEX, c(1, 2), c("Men", "Women")),
    diabetes = DIABETES, smoker = CURSMOKE, bpmed = BPMEDS,
    chd = PREVCHD, stroke10 = STROKE_10yr
  ) %>%
  select(sex, AGE, SYSBP, TOTCHOL, BMI, diabetes, smoker, bpmed, chd, stroke10, FU_10yr)

n_overall_t1 <- nrow(t1); n_men_t1 <- sum(t1$sex == "Men"); n_women_t1 <- sum(t1$sex == "Women")

row_cont <- function(v, lab) tibble(
  Characteristic = lab,
  Overall = fmt_mean_sd(t1[[v]]),
  Men     = fmt_mean_sd(t1 %>% filter(sex == "Men")   %>% pull(all_of(v))),
  Women   = fmt_mean_sd(t1 %>% filter(sex == "Women") %>% pull(all_of(v)))
)
row_miss <- function(v) {
  m <- fmt_missing_n(t1[[v]]); if (is.na(m)) return(NULL)
  tibble(Characteristic = "  Missing, n", Overall = m,
         Men   = fmt_missing_n(t1 %>% filter(sex == "Men")   %>% pull(all_of(v))),
         Women = fmt_missing_n(t1 %>% filter(sex == "Women") %>% pull(all_of(v))))
}
row_bin <- function(v, lab) tibble(
  Characteristic = lab,
  Overall = fmt_n_pct(t1[[v]]),
  Men     = fmt_n_pct(t1 %>% filter(sex == "Men")   %>% pull(all_of(v))),
  Women   = fmt_n_pct(t1 %>% filter(sex == "Women") %>% pull(all_of(v)))
)

table1_body <- bind_rows(
  row_cont("AGE",     "Age (years)"),
  row_cont("SYSBP",   "Systolic BP (mmHg)"),
  row_cont("TOTCHOL", "Total cholesterol (mg/dL)"),
  row_miss("TOTCHOL"),
  row_cont("BMI",     "BMI (kg/m\u00B2)"),
  row_miss("BMI"),
  row_bin("diabetes", "Diabetes mellitus"),
  row_bin("smoker",   "Current smoker"),
  row_bin("bpmed",    "Antihypertensive medication"),
  row_miss("bpmed"),
  row_bin("chd",      "Prevalent CHD"),
  row_bin("stroke10", "10-year stroke event")
) %>% mutate(section = "Baseline characteristics")

followup_body <- bind_rows(
  tibble(Characteristic = "Time to event (days)",
         Overall = fmt_time_ms(t1$FU_10yr, t1$stroke10, 1),
         Men   = fmt_time_ms(t1$FU_10yr[t1$sex=="Men"],   t1$stroke10[t1$sex=="Men"],   1),
         Women = fmt_time_ms(t1$FU_10yr[t1$sex=="Women"], t1$stroke10[t1$sex=="Women"], 1)),
  tibble(Characteristic = "Time to censoring (days)",
         Overall = fmt_time_ms(t1$FU_10yr, t1$stroke10, 0),
         Men   = fmt_time_ms(t1$FU_10yr[t1$sex=="Men"],   t1$stroke10[t1$sex=="Men"],   0),
         Women = fmt_time_ms(t1$FU_10yr[t1$sex=="Women"], t1$stroke10[t1$sex=="Women"], 0)),
  tibble(Characteristic = "10-year stroke-free survival (KM)",
         Overall = fmt_km(t1$FU_10yr, t1$stroke10),
         Men   = fmt_km(t1$FU_10yr[t1$sex=="Men"],   t1$stroke10[t1$sex=="Men"]),
         Women = fmt_km(t1$FU_10yr[t1$sex=="Women"], t1$stroke10[t1$sex=="Women"]))
) %>% mutate(section = "Follow-up and survival")

table1_data <- bind_rows(table1_body, followup_body)

table1_gt <- table1_data %>%
  gt(groupname_col = "section") %>%
  cols_label(
    Characteristic = "Characteristic",
    Overall = md(paste0("**Overall**<br>N = ", format(n_overall_t1, big.mark = ","))),
    Men     = md(paste0("**Men**<br>N = ",     format(n_men_t1,     big.mark = ","))),
    Women   = md(paste0("**Women**<br>N = ",   format(n_women_t1,   big.mark = ",")))
  ) %>%
  cols_align("left", Characteristic) %>%
  cols_align("center", c(Overall, Men, Women)) %>%
  tab_source_note(md("Continuous variables: mean (SD). Binary variables: n (%) using non-missing denominator. Missing counts shown only when > 0. The KM row reports estimated stroke-free survival at 10 years.")) %>%
  style_gt(caption_md = "**Table 1.** Baseline characteristics and 10-year follow-up experience, overall and by sex.")

#TABLE 2: risk factors across examination periods
analytic_ids <- analytic %>% distinct(RANDID)
table2_src   <- data %>% semi_join(analytic_ids, by = "RANDID")

cell_cont <- function(x) {
  if (!sum(!is.na(x))) return("\u2014")
  sprintf("%.1f (%.1f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
}
cell_bin  <- function(x, lv) {
  n <- sum(!is.na(x)); if (!n) return("\u2014")
  k <- sum(x == lv, na.rm = TRUE)
  sprintf("%d (%.1f%%)", k, 100 * k / n)
}
n_nm <- function(x) sum(!is.na(x))
n_lv <- function(x, lv) sum(x == lv, na.rm = TRUE)

build_cont <- function(v, lab) {
  g <- function(p) table2_src[[v]][table2_src$PERIOD == p]
  tibble(Variable = lab,
         p1_n = as.character(n_nm(g(1))), p1_v = cell_cont(g(1)),
         p2_n = as.character(n_nm(g(2))), p2_v = cell_cont(g(2)),
         p3_n = as.character(n_nm(g(3))), p3_v = cell_cont(g(3)),
         section = "Continuous")
}
build_bin <- function(v, lab) {
  g <- function(p) table2_src[[v]][table2_src$PERIOD == p]
  tibble(Variable = lab,
         p1_n = as.character(n_lv(g(1), 1)), p1_v = cell_bin(g(1), 1),
         p2_n = as.character(n_lv(g(2), 1)), p2_v = cell_bin(g(2), 1),
         p3_n = as.character(n_lv(g(3), 1)), p3_v = cell_bin(g(3), 1),
         section = "Binary (Yes)")
}

table2_data <- bind_rows(
  build_cont("AGE",     "Age (years)"),
  build_cont("SYSBP",   "Systolic BP (mmHg)"),
  build_cont("TOTCHOL", "Total cholesterol (mg/dL)"),
  build_cont("BMI",     "BMI (kg/m\u00B2)"),
  build_bin("DIABETES", "Diabetes"),
  build_bin("CURSMOKE", "Current smoker"),
  build_bin("BPMEDS",   "Antihypertensive medication"),
  build_bin("PREVCHD",  "Prevalent CHD")
)

n_p1 <- sum(table2_src$PERIOD == 1, na.rm = TRUE)
n_p2 <- sum(table2_src$PERIOD == 2, na.rm = TRUE)
n_p3 <- sum(table2_src$PERIOD == 3, na.rm = TRUE)

table2_gt <- table2_data %>%
  gt(groupname_col = "section") %>%
  tab_spanner(md(paste0("**Period 1**<br>N = ", format(n_p1, big.mark = ","))), c(p1_n, p1_v)) %>%
  tab_spanner(md(paste0("**Period 2**<br>N = ", format(n_p2, big.mark = ","))), c(p2_n, p2_v)) %>%
  tab_spanner(md(paste0("**Period 3**<br>N = ", format(n_p3, big.mark = ","))), c(p3_n, p3_v)) %>%
  cols_label(Variable = "Variable",
             p1_n = "n", p1_v = "Value",
             p2_n = "n", p2_v = "Value",
             p3_n = "n", p3_v = "Value") %>%
  cols_align("left", Variable) %>%
  cols_align("center", c(p1_n, p1_v, p2_n, p2_v, p3_n, p3_v)) %>%
  tab_source_note(md("Same baseline-eligible cohort as Table 1. Continuous: mean (SD). Binary: n (%) at the Yes level, using the non-missing denominator within period.")) %>%
  style_gt(caption_md = "**Table 2.** Distribution of selected risk factors across examination periods.")

#KM log-rank table-Table S1
analytic <- analytic %>%
  mutate(
    sex_label      = factor(SEX,      c(1,2), c("Men","Women")),
    diabetes_label = factor(DIABETES, c(0,1), c("No diabetes","Diabetes")),
    smoke_label    = factor(CURSMOKE, c(0,1), c("Not current","Current")),
    prevchd_label  = factor(PREVCHD,  c(0,1), c("No CHD","CHD")),
    bpmeds_label   = factor(BPMEDS,   c(0,1), c("No BP meds","BP meds"))
  )

analytic_men          <- analytic %>% filter(SEX == 1)
analytic_women        <- analytic %>% filter(SEX == 2)
analytic_men_bpmeds   <- analytic_men   %>% filter(!is.na(BPMEDS))
analytic_women_bpmeds <- analytic_women %>% filter(!is.na(BPMEDS))

logrank_summary <- function(formula, data, group_var, stratum = NA_character_) {
  fit <- survdiff(formula, data = data)
  tibble(
    stratum  = stratum,
    variable = group_var,
    chisq    = fit$chisq,
    df       = length(fit$n) - 1,
    p_value  = 1 - pchisq(fit$chisq, df = length(fit$n) - 1),
    n_total  = sum(fit$n)
  )
}

km_logrank_results <- bind_rows(
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ sex_label,      analytic,              "Sex",                         "Overall"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_men,          "Diabetes",                    "Men"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label,    analytic_men,          "Current smoking",             "Men"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label,   analytic_men_bpmeds,   "Antihypertensive med.",       "Men"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label,  analytic_men,          "Prevalent CHD",               "Men"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT,      analytic_men,          "Systolic BP \u2265 160 mmHg", "Men"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_women,        "Diabetes",                    "Women"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label,    analytic_women,        "Current smoking",             "Women"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label,   analytic_women_bpmeds, "Antihypertensive med.",       "Women"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label,  analytic_women,        "Prevalent CHD",               "Women"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT,      analytic_women,        "Systolic BP \u2265 160 mmHg", "Women")
)

logrank_gt <- km_logrank_results %>%
  gt(groupname_col = "stratum") %>%
  cols_label(
    variable = "Comparison",
    chisq    = html("&chi;<sup>2</sup>"),
    df       = "df",
    p_value  = "P-value",
    n_total  = "N"
  ) %>%
  fmt_number(columns = chisq, decimals = 2) %>%
  fmt(columns = p_value,
      fns = function(x) ifelse(x < 0.001, "<0.001", sprintf("%.3f", x))) %>%
  fmt_number(columns = n_total, decimals = 0, use_seps = TRUE) %>%
  cols_align("left",   variable) %>%
  cols_align("center", c(chisq, df, p_value, n_total)) %>%
  tab_source_note(md("Log-rank test of equality of 10-year stroke-free survival between groups, within each stratum.")) %>%
  style_gt(caption_md = "**Table S1.** Log-rank tests for unadjusted 10-year stroke comparisons.")

#TABLE 3: final Cox models
format_hr_ci <- function(hr, lo, hi) sprintf("%.2f (%.2f, %.2f)", hr, lo, hi)
format_p     <- function(p) ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))

table3_data <- cox_final_results %>%
  mutate(
    term_lab = unname(term_labels[term]),
    hr_ci    = format_hr_ci(estimate, conf.low, conf.high),
    p_fmt    = format_p(p.value)
  ) %>%
  select(sex, term_lab, hr_ci, p_fmt) %>%
  pivot_wider(
    names_from  = sex,
    values_from = c(hr_ci, p_fmt),
    names_glue  = "{sex}_{.value}"
  ) %>%
  mutate(term_lab = factor(term_lab, levels = c(
    "Age (per 1 year)", "Systolic BP (per 10 mmHg)", "Diabetes",
    "Current smoker", "Antihypertensive medication",
    "Prevalent CHD", "Total cholesterol (per 10 mg/dL)",
    "BMI (per 1 kg/m\u00B2)"
  ))) %>%
  arrange(term_lab) %>%
  select(term_lab, men_hr_ci, men_p_fmt, women_hr_ci, women_p_fmt)

table3_gt <- table3_data %>%
  gt() %>%
  cols_label(
    term_lab    = "Variable",
    men_hr_ci   = "HR (95% CI)",
    men_p_fmt   = "P-value",
    women_hr_ci = "HR (95% CI)",
    women_p_fmt = "P-value"
  ) %>%
  tab_spanner(label = md("**Men**"),   columns = c(men_hr_ci,   men_p_fmt)) %>%
  tab_spanner(label = md("**Women**"), columns = c(women_hr_ci, women_p_fmt)) %>%
  cols_align("left",   term_lab) %>%
  cols_align("center", c(men_hr_ci, men_p_fmt, women_hr_ci, women_p_fmt)) %>%
  tab_source_note(md("Hazard ratios from sex-specific Cox proportional hazards models using the union of LASSO-selected predictors (\u03BB.min). Age, systolic BP, and diabetes were forced into the model.")) %>%
  style_gt(caption_md = "**Table 3.** Adjusted hazard ratios for 10-year incident stroke, by sex.")


#PH diagnostics
ph_men   <- cox.zph(cox_final_men)
ph_women <- cox.zph(cox_final_women)
ph_results <- bind_rows(
  as.data.frame(ph_men$table)   %>% rownames_to_column("term") %>% mutate(sex = "men"),
  as.data.frame(ph_women$table) %>% rownames_to_column("term") %>% mutate(sex = "women")
) %>% select(sex, term, chisq, df, p) %>% rename(p_value = p)

#Profile predictions + Table 4

sysbp_normal_men <- refit_men_data %>%
  filter(DIABETES == 0, CURSMOKE == 0) %>% summarise(m = mean(SYSBP)) %>% pull(m)
sysbp_normal_women <- refit_women_data %>%
  filter(DIABETES == 0, CURSMOKE == 0) %>% summarise(m = mean(SYSBP)) %>% pull(m)

build_profiles <- function(sbp_norm) {
  ages <- c(40, 50, 60)
  defs <- tribble(
    ~profile, ~profile_label,                    ~SYSBP,   ~DIABETES, ~CURSMOKE,
    "P1",     "No major risk factors",           sbp_norm, 0,         0,
    "P2",     "High SBP only",                   160,      0,         0,
    "P3",     "Diabetes only",                   sbp_norm, 1,         0,
    "P4",     "High SBP + diabetes",             160,      1,         0,
    "P5",     "High SBP + diabetes + smoking",   160,      1,         1
  )
  expand_grid(AGE = ages, defs) %>%
    mutate(BPMEDS = 0) %>%
    select(profile, profile_label, AGE, SYSBP, DIABETES, CURSMOKE, BPMEDS)
}

profiles_men   <- build_profiles(sysbp_normal_men)
profiles_women <- build_profiles(sysbp_normal_women)

predict_10yr_stroke <- function(fit, prof) {
  sf <- survfit(fit, newdata = prof)
  s  <- summary(sf, times = days_10yr)
  tibble(surv = as.numeric(s$surv),
         surv_lower = as.numeric(s$lower),
         surv_upper = as.numeric(s$upper)) %>%
    mutate(stroke_prob       = 1 - surv,
           stroke_prob_lower = 1 - surv_upper,
           stroke_prob_upper = 1 - surv_lower) %>%
    bind_cols(prof, .)
}

men_predictions   <- predict_10yr_stroke(cox_final_men,   profiles_men)
women_predictions <- predict_10yr_stroke(cox_final_women, profiles_women)

all_predictions <- bind_rows(
  men_predictions   %>% mutate(sex = "men"),
  women_predictions %>% mutate(sex = "women")
) %>%
  mutate(
    stroke_prob_pct       = round(stroke_prob       * 100, 2),
    stroke_prob_lower_pct = round(stroke_prob_lower * 100, 2),
    stroke_prob_upper_pct = round(stroke_prob_upper * 100, 2)
  ) %>%
  select(sex, profile, profile_label, AGE, SYSBP, DIABETES, CURSMOKE, BPMEDS,
         stroke_prob_pct, stroke_prob_lower_pct, stroke_prob_upper_pct)

format_pct_ci <- function(est, lo, hi) sprintf("%.1f%% (%.1f, %.1f)", est, lo, hi)

profile_info <- profiles_men %>%
  distinct(profile, profile_label, DIABETES, CURSMOKE, BPMEDS, SYSBP) %>%
  rename(men_sbp = SYSBP) %>%
  left_join(profiles_women %>% distinct(profile, SYSBP) %>% rename(women_sbp = SYSBP),
            by = "profile") %>%
  mutate(
    Diabetes = ifelse(DIABETES == 1, "Yes", "No"),
    Smoking  = ifelse(CURSMOKE == 1, "Yes", "No"),
    bp_med   = ifelse(BPMEDS   == 1, "Yes", "No"),
    men_sbp   = sprintf("%.0f", men_sbp),
    women_sbp = sprintf("%.0f", women_sbp)
  ) %>%
  select(profile, profile_label, Diabetes, Smoking, bp_med, men_sbp, women_sbp)

predictions_wide <- all_predictions %>%
  mutate(
    sex  = ifelse(sex == "men", "men", "women"),
    risk = format_pct_ci(stroke_prob_pct, stroke_prob_lower_pct, stroke_prob_upper_pct),
    col  = paste0(sex, "_age", AGE)
  ) %>%
  select(profile, col, risk) %>%
  pivot_wider(names_from = col, values_from = risk)

table4_data <- profile_info %>%
  left_join(predictions_wide, by = "profile") %>%
  select(profile, profile_label, Diabetes, Smoking, bp_med,
         men_sbp,   men_age40,   men_age50,   men_age60,
         women_sbp, women_age40, women_age50, women_age60)

table4_gt <- table4_data %>%
  gt() %>%
  cols_label(
    profile       = "Profile",
    profile_label = "Description",
    Diabetes      = "Diabetes",
    Smoking       = "Smoking",
    bp_med        = "BP med",
    men_sbp       = "SBP",
    men_age40     = "Age 40",
    men_age50     = "Age 50",
    men_age60     = "Age 60",
    women_sbp     = "SBP",
    women_age40   = "Age 40",
    women_age50   = "Age 50",
    women_age60   = "Age 60"
  ) %>%
  tab_spanner(label = md("**Profile definition**"),
              columns = c(profile_label, Diabetes, Smoking, bp_med)) %>%
  tab_spanner(label = md("**Men — predicted 10-year stroke probability**"),
              columns = c(men_sbp, men_age40, men_age50, men_age60)) %>%
  tab_spanner(label = md("**Women — predicted 10-year stroke probability**"),
              columns = c(women_sbp, women_age40, women_age50, women_age60)) %>%
  cols_align("left",   c(profile, profile_label)) %>%
  cols_align("center", c(Diabetes, Smoking, bp_med,
                         men_sbp,   men_age40,   men_age50,   men_age60,
                         women_sbp, women_age40, women_age50, women_age60)) %>%
  tab_source_note(md("Predicted risks are percent (95% CI) from the final sex-specific Cox models. Antihypertensive medication fixed at No for all profiles. SBP shown in mmHg; 'normal' SBP values use the sex-specific mean among model-sample participants without diabetes who did not currently smoke.")) %>%
  style_gt(caption_md = "**Table 4.** Predicted 10-year stroke probability by sex, age, and prespecified baseline risk profile.")


#Figure 2 data
figure2_data <- all_predictions %>%
  mutate(
    sex     = factor(ifelse(sex == "men", "Men", "Women"), levels = c("Men", "Women")),
    profile = factor(profile,
                     levels = c("P1", "P2", "P3", "P4", "P5"),
                     labels = c("P1: Low risk",
                                "P2: High SBP",
                                "P3: Diabetes",
                                "P4: SBP + DM",
                                "P5: SBP + DM + Smoke"))
  )

#counts used inline

N_baseline          <- nrow(baseline)
N_excluded_prevstrk <- N_baseline - nrow(analytic)
N_analytic          <- nrow(analytic)
N_men               <- nrow(analytic_men)
N_women             <- nrow(analytic_women)
N_modeling_men      <- nrow(refit_men_data)
N_modeling_women    <- nrow(refit_women_data)
n_events_men        <- sum(refit_men_data$STROKE_10yr)
n_events_women      <- sum(refit_women_data$STROKE_10yr)

get_hr_row <- function(df, sex_in, term_in) df %>% filter(sex == sex_in, term == term_in) %>% slice(1)
```

# Introduction

The dataset for this project was a longitudinal subset of the Framingham Heart Study containing clinical, laboratory, and event data on 4,434 participants observed across as many as three examination periods. The main goals of the analysis were to identify which baseline risk factors were associated with incident stroke and to estimate 10-year stroke probability for clinically meaningful baseline risk profiles. In accordance with the investigartor's specifications, the analysis used only Period 1 covariates, restricted follow-up to 10 years, excluded participants with prevalent stroke at baseline, treated death as a censoring event, and was conducted separately in men and women. The primary scientific hypothesis was that older age, higher systolic blood pressure, and baseline diabetes would each be associated with greater risk of 10-year incident stroke within each sex, while five additional variables were evaluated as candidate covariates. These primary associations were assessed through two-sided hypothesis tests of whether the corresponding Cox regression coefficients were equal to zero in the sex-specific models. A secondary objective was to use the final male and female models to estimate 10-year stroke probabilities for five prespecified baseline profiles at ages 40, 50, and 60. This secondary objective was descriptive and emphasized point estimates with 95% confidence intervals rather than formal hypothesis testing.

# Methods

## Data management

Continuous baseline variables (age, systolic blood pressure, total cholesterol, body mass index) were screened for implausible values. A small number of extreme systolic blood pressure and total cholesterol values were flagged (2) and, per investigator review, retained. Binary variables (diabetes, current smoking, blood pressure medication use, prevalent coronary heart disease) were checked to confirm 0/1 coding. Baseline missingness was confined to blood pressure medication use, total cholesterol, and body mass index, while three primary predictors were complete. A categorical systolic blood pressure variable (<160 vs. ≥160 mmHg) was derived for unadjusted survival displays to align with the ≥160 mmHg threshold used in the prespecified risk profiles.

Time to first incident stroke was measured from the Framingham stroke indicator and time-to-stroke variable, with follow-up administratively censored at 3,652.5 days (10 years). Strokes at or before that time were treated as events. Stroke free participants through 10 years were censored at that time, and those who died or were lost to follow-up earlier were censored at their last observed time.

## Descriptive analyses

Descriptive statistics summarized baseline covariates and follow-up overall and by sex (Table 1) and tracked risk factor distributions across Periods 1–3 within the baseline-eligible cohort (Table 2) to gauge covariate change over time. Kaplan–Meier methods, which are appropriate for right-censored time-to-event data, estimated unadjusted 10-year stroke free survival by sex and, within each sex, by diabetes, current smoking, blood pressure medication use, prevalent coronary heart disease, and the categorical systolic blood pressure variable. Because 10-year stroke incidence was low, curves were displayed as cumulative incidence (one minus survival) for visual clarity. Log-rank tests accompanied each comparison (Table S1) and were interpreted descriptively. Adjusted inference came from the Cox models below.

## Cox proportional hazards modeling and variable selection

A sex-specific Cox proportional hazards model was fit for time to 10-year incident stroke. The Cox model was appropriate because the outcome was right-censored time-to-event data with a non-competing censoring mechanism, and the sex-specific fit addressed the investigator's requirement for separate male and female analyses without requiring proportional hazards to hold across sex. Age, systolic blood pressure, and diabetes were included in each model as primary predictors, while current smoking, blood pressure medication use, prevalent coronary heart disease, total cholesterol, and body mass index were the five candidate covariates, treated as potential confounders or precision variables.

Because there was no a priori reason to include every candidate and the number of stroke events was small relative to the number of candidates, variable selection used a LASSO-penalized Cox regression rather than stepwise selection, since LASSO handles correlated predictors more stably and reduces overfitting in settings with limited events. The LASSO was fit within each sex with 5-fold cross-validation and penalty factors of 0 for the primary predictors and 1 for the candidates, so that the primary predictors were always retained. The penalty value minimizing cross-validated partial likelihood deviance (`lambda.min`) determined the selected set. The union of predictors selected across sexes was then adopted as the final covariate set for both models, so that the two sex-specific hazard ratios would be directly comparable. Standard unpenalized Cox models were refit separately in men and women using complete cases on this final set.

Hazard ratios, 95% confidence intervals, and two-sided Wald p-values were obtained from the refit models, with age reported per 1-year increase, systolic blood pressure per 10 mmHg, and total cholesterol per 10 mg/dL when retained. The Wald tests of the age, systolic blood pressure, and diabetes coefficients in each model addressed the primary statistical hypotheses, evaluated at two-sided $\alpha = 0.05$. No multiple comparison adjustment was applied because the three tests corresponded to prespecified primary hypotheses rather than exploratory comparisons. The proportional hazards assumption was assessed using Schoenfeld residual tests (per coefficient and global) supplemented by scaled Schoenfeld and log-minus-log plots.

## Profile-specific 10-year stroke probability estimation

Ten-year stroke probabilities and 95% confidence intervals were estimated for five prespecified profiles at ages 40, 50, and 60: (P1) no major risk factors, (P2) high systolic blood pressure only, (P3) diabetes only, (P4) high systolic blood pressure plus diabetes, and (P5) high systolic blood pressure, diabetes, and current smoking. High systolic blood pressure was fixed at 160 mmHg and antihypertensive medication use at No in all profiles. Low risk systolic blood pressure was set to the sex specific mean among model sample participants without diabetes who were not current smokers. Predicted probabilities were obtained from the sex specific Cox survival estimates at 10 years.

# Results

## Cohort summary and descriptive analyses

The cohort-construction steps are shown in Figure 1. The Period 1 baseline cohort contained `r N_baseline` participants; after excluding `r N_excluded_prevstrk` with prevalent stroke, `r N_analytic` remained (`r N_men` men, `r N_women` women). Complete-case refit Cox model samples contained `r N_modeling_men` men and `r N_modeling_women` women, with `r n_events_men` and `r n_events_women` 10-year stroke events, respectively.

<p style="text-align:center; font-weight:bold; margin-top:1.5em; margin-bottom:0.4em;">
Figure 1. Cohort flow diagram.
</p>

```{r cohort_flow, out.width='80%'}
grViz(cohort_flow_code)
```

<p style="text-align:center; font-size:0.9em; color:#444; max-width:720px; margin:0.4em auto 1.5em auto;">
Cohort flow for construction of the analytic cohort and sex-specific modeling samples.
</p>

Baseline characteristics and 10-year follow-up experience by sex are summarized in Table 1. Men and women were similar in age but differed on other risk factors, with men more often current smokers and women more often on antihypertensive medication. Ten-year incident stroke was rare and comparable between sexes. More than 97% of each sex remained stroke free at 10 years.

```{r table1}
table1_gt
```

Table 2 tracks the same risk factors across Periods 1–3. Systolic blood pressure, diabetes prevalence, and antihypertensive medication use increased across periods while current smoking declined, indicating that baseline values became less representative of later exposure status.

```{r table2}
table2_gt
```

## Unadjusted survival patterns

Log-rank tests for unadjusted 10-year stroke comparisons are shown in Table S1. A significant log-rank test indicates that 10-year stroke-free survival differed between the compared groups, without adjustment for other risk factors. The overall sex comparison was not significant (p = `r sprintf("%.3f", km_logrank_results$p_value[km_logrank_results$stratum == "Overall" & km_logrank_results$variable == "Sex"])`). Categorical systolic blood pressure showed the strongest unadjusted separation in both sexes (p < 0.001). In men, diabetes and antihypertensive medication use were also strongly associated with 10-year stroke incidence (both p < 0.001), while prevalent CHD was not statistically significant (p = 0.084). In women, antihypertensive medication use and prevalent CHD were also strongly associated (both p < 0.001), with a weaker but still significant association for diabetes (p = 0.013). Current smoking showed no significant unadjusted association in either sex.

```{r km_logrank_table}
logrank_gt
```

## Cox proportional hazards model results

LASSO selection at `lambda.min` retained two of the five candidate covariates in men (current smoking and antihypertensive medication use) and none in women. Under the union rule, both were added to the three forced primary predictors, yielding a common final covariate set for both sexes: `r paste(final_vars_men, collapse = ", ")`. Adjusted hazard ratios from the refit sex-specific Cox models are shown in Table 3.

In men, each 1-year increase in age (HR `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "AGE (per 1 year)")$estimate)`, 95% CI `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "AGE (per 1 year)")$conf.low)`–`r sprintf("%.2f", get_hr_row(cox_final_results, "men", "AGE (per 1 year)")$conf.high)`; p `r format_p(get_hr_row(cox_final_results, "men", "AGE (per 1 year)")$p.value)`), each 10 mmHg increase in systolic blood pressure (HR `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "SYSBP (per 10 mmHg)")$estimate)`, `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "SYSBP (per 10 mmHg)")$conf.low)`–`r sprintf("%.2f", get_hr_row(cox_final_results, "men", "SYSBP (per 10 mmHg)")$conf.high)`; p `r format_p(get_hr_row(cox_final_results, "men", "SYSBP (per 10 mmHg)")$p.value)`), and diabetes (HR `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "DIABETES")$estimate)`, `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "DIABETES")$conf.low)`–`r sprintf("%.2f", get_hr_row(cox_final_results, "men", "DIABETES")$conf.high)`; p `r format_p(get_hr_row(cox_final_results, "men", "DIABETES")$p.value)`) were each associated with higher 10-year stroke hazard — for example, the age HR indicates a `r sprintf("%.0f", (get_hr_row(cox_final_results, "men", "AGE (per 1 year)")$estimate - 1) * 100)`% higher hazard per additional year, holding other predictors fixed — and all three prespecified null hypotheses were rejected. Current smoking was also significant (HR `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "CURSMOKE")$estimate)`, `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "CURSMOKE")$conf.low)`–`r sprintf("%.2f", get_hr_row(cox_final_results, "men", "CURSMOKE")$conf.high)`; p `r format_p(get_hr_row(cox_final_results, "men", "CURSMOKE")$p.value)`); antihypertensive medication use was not (HR `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "BPMEDS")$estimate)`; p `r format_p(get_hr_row(cox_final_results, "men", "BPMEDS")$p.value)`).

In women, age (HR `r sprintf("%.2f", get_hr_row(cox_final_results, "women", "AGE (per 1 year)")$estimate)`, `r sprintf("%.2f", get_hr_row(cox_final_results, "women", "AGE (per 1 year)")$conf.low)`–`r sprintf("%.2f", get_hr_row(cox_final_results, "women", "AGE (per 1 year)")$conf.high)`; p `r format_p(get_hr_row(cox_final_results, "women", "AGE (per 1 year)")$p.value)`) and systolic blood pressure (HR `r sprintf("%.2f", get_hr_row(cox_final_results, "women", "SYSBP (per 10 mmHg)")$estimate)` per 10 mmHg, `r sprintf("%.2f", get_hr_row(cox_final_results, "women", "SYSBP (per 10 mmHg)")$conf.low)`–`r sprintf("%.2f", get_hr_row(cox_final_results, "women", "SYSBP (per 10 mmHg)")$conf.high)`; p `r format_p(get_hr_row(cox_final_results, "women", "SYSBP (per 10 mmHg)")$p.value)`) showed similarly elevated adjusted hazards, while the diabetes coefficient was attenuated and imprecise (HR `r sprintf("%.2f", get_hr_row(cox_final_results, "women", "DIABETES")$estimate)`, `r sprintf("%.2f", get_hr_row(cox_final_results, "women", "DIABETES")$conf.low)`–`r sprintf("%.2f", get_hr_row(cox_final_results, "women", "DIABETES")$conf.high)`; p `r format_p(get_hr_row(cox_final_results, "women", "DIABETES")$p.value)`); the null was rejected for age and systolic blood pressure but not diabetes. Neither current smoking nor antihypertensive medication use was significantly associated with stroke hazard in women (both p > 0.1).

Schoenfeld residual tests did not indicate meaningful departure from proportional hazards in either model (global p = `r sprintf("%.3f", ph_results$p_value[ph_results$sex == "men" & ph_results$term == "GLOBAL"])` in men, `r sprintf("%.3f", ph_results$p_value[ph_results$sex == "women" & ph_results$term == "GLOBAL"])` in women). Log-minus-log plots showed some non-parallelism in sparsely populated subgroups, likely reflecting small event counts rather than true PH violation, so the models were retained without modification.

```{r table3}
table3_gt
```

## Predicted 10-year stroke probability by risk profile

Predicted 10-year stroke probabilities for the five prespecified profiles at ages 40, 50, and 60 are shown in Table 4 and Figure 2. Based on point estimates, predicted risk increased with age within every profile, and within each sex. In men this ordering was monotonic, but in women P2 (high SBP only) produced a slightly higher point estimate than P3 (diabetes only), possibly reflecting the weak and non significant diabetes coefficient in the female model combined with the SBP increase from around 136 to 160 mmHg. At age 60, point estimates in men ranged from `r sprintf("%.1f%%", all_predictions %>% filter(sex == "men", profile == "P1", AGE == 60) %>% pull(stroke_prob_pct))` for profile P1 (no major risk factors) to `r sprintf("%.1f%%", all_predictions %>% filter(sex == "men", profile == "P5", AGE == 60) %>% pull(stroke_prob_pct))` for profile P5 (high systolic blood pressure, diabetes, and current smoking). The corresponding range in women was `r sprintf("%.1f%%", all_predictions %>% filter(sex == "women", profile == "P1", AGE == 60) %>% pull(stroke_prob_pct))` to `r sprintf("%.1f%%", all_predictions %>% filter(sex == "women", profile == "P5", AGE == 60) %>% pull(stroke_prob_pct))`. 

However, the ordering across profiles and ages was not strongly supported by the data. Confidence intervals were wide for the higher risk male profiles, and for women, all profiles that included diabetes had lower bounds at or near 0%. The same was true for the youngest high risk male profiles, indicating that the data could not exclude a very low underlying risk in these subgroups. As a result, confidence intervals for most high risk profiles overlapped considerably with those for lower risk profiles at the same age and with younger ages within the same profile. Therefore, the differences in point estimates are best interpreted as suggesting an overall pattern of increasing risk, rather than providing precise estimates of the magnitude of those differences. This lack of precision likely reflects the small number of stroke events in these subgroups, especially among women with diabetes.

```{r table4}
table4_gt
```

<p style="text-align:center; font-weight:bold; margin-top:1.5em; margin-bottom:0.4em;">
Figure 2. Predicted 10-year stroke probability by age, sex, and baseline risk profile.
</p>

```{r figure2, fig.width=8.5, fig.height=4.8, out.width='95%'}
profile_colors <- c(
  "P1: Low risk"          = "#117733",
  "P2: High SBP"          = "#88CCEE",
  "P3: Diabetes"          = "#DDCC77",
  "P4: SBP + DM"          = "#CC6677",
  "P5: SBP + DM + Smoke"  = "#882255"
)

ggplot(figure2_data,
       aes(x = AGE, y = stroke_prob_pct,
           group = profile, color = profile, shape = profile)) +
  geom_ribbon(aes(ymin = stroke_prob_lower_pct, ymax = stroke_prob_upper_pct,
                  fill = profile),
              alpha = 0.10, color = NA, show.legend = FALSE) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.6, fill = "white", stroke = 0.9) +
  facet_wrap(~ sex) +
  scale_x_continuous(breaks = c(40, 50, 60)) +
  scale_y_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0.02, 0.08))) +
  scale_color_manual(values = profile_colors) +
  scale_fill_manual(values  = profile_colors) +
  scale_shape_manual(values = c(16, 17, 15, 18, 21)) +
  labs(
    x = "Age (years)",
    y = "Predicted 10-year stroke probability",
    color = "Risk profile", shape = "Risk profile"
  ) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE),
         shape = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal(base_family = "sans", base_size = 12) +
  theme(
    legend.position    = "bottom",
    legend.title       = element_text(face = "bold"),
    legend.key.width   = unit(1.2, "cm"),
    strip.text         = element_text(face = "bold", size = 12),
    strip.background   = element_rect(fill = "#F4F7FA", color = NA),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.spacing      = unit(1.5, "lines"),
    plot.margin        = margin(10, 10, 10, 10),
    axis.title.x       = element_text(margin = margin(t = 8)),
    axis.title.y       = element_text(margin = margin(r = 8))
  )
```

<p style="text-align:center; font-size:0.9em; color:#444; max-width:760px; margin:0.2em auto 1.5em auto;">
Predicted 10-year stroke probability by age, sex, and prespecified baseline risk profile, from the final sex-specific Cox proportional hazards models. Shaded bands indicate 95% confidence intervals.
</p>

# Discussion

Age and systolic blood pressure emerged as the most consistent baseline predictors of 10-year incident stroke in both sexes, with statistically significant adjusted hazards in each model. Diabetes was a strong, significant predictor in men and contributed to the male risk profile predictions, but its adjusted association in women was attenuated and imprecise, likely reflecting the small number of stroke events among women with diabetes. One consequence visible in the profile based predictions is that in women, the high SBP profile produced slightly higher predicted 10-year stroke risk than the diabetes only profile at every age, possibly driven by the near-null female diabetes coefficient rather than any protective effect of diabetes. Predicted stroke risk rose with age and with the number of risk factors in both sexes, but the confidence intervals for higher risk profiles were wide and overlapped across profiles and ages. These patterns show the general direction of increasing risk but not precise estimates.

Several limitations should be noted. First, the number of 10-year stroke events was small relative to the number of candidate predictors, which reduced precision and led to wide confidence intervals. This was most apparent for the diabetes effect in women. All female risk profiles that included diabetes had lower confidence bounds reaching 0%. Similar imprecision was also seen in the women’s current smoking and antihypertensive medication coefficients, both of which had confidence intervals that crossed the null.

Second, the analysis used Period 1 values only, although Table 2 showed meaningful drift in systolic blood pressure, diabetes prevalence, and medication use across follow-up. This indicates that a time-varying covariate analysis could yield different effect estimates if later exposure status diverged substantially from baseline. Third, per the investigator's specification, death was treated as right-censoring rather than as a competing risk, which is appropriate for a cause specific Cox analysis but may overstate absolute stroke probability in profiles with high competing mortality risk. 

Within these constraints, the analysis answered the investigator's questions: age and systolic blood pressure were significant adjusted predictors of 10-year incident stroke in both sexes, diabetes was a significant predictor in men, and the sex-specific models produced 10-year stroke probability point estimates for each prespecified baseline profile. Precision of those probability estimates was limited in women and in higher risk male profiles, and the investigator may wish to interpret absolute risks cautiously in those subgroups. 

# Reproducible Research Information

- **GitHub Repository**: `https://github.com/emkokesh/BIOS6624`

- **Raw dataset**: `C:/Users/kokeshem/Spring 2026/Advanced Statistical Methods/Project 3/frmgham2.csv`

- **Primary analysis script**: `Code/Primary_Analysis_Project3.Rmd`

- **Report source**: `Reports/Project3_Report_Final.Rmd`
