---
title: "Baseline Risk Factors and 10-Year Probability of Incident Stroke: A Sex-Stratified Cox Proportional Hazards Analysis of the Framingham Heart Study"
author: "Emily Kokesh"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 150
)

library(tidyverse)
library(survival)
library(survminer)
library(glmnet)
library(broom)
library(gt)
library(knitr)
library(DiagrammeR)
library(glue)
library(tibble)
library(scales)

#repository, data paths
repo_url  <- "[insert GitHub repository link here]"
data_path <- "C:/Users/kokeshem/Spring 2026/Advanced Statistical Methods/Project 3/frmgham2.csv"

#shared gt theme
style_gt <- function(gt_tbl, caption_md = NULL) {
  out <- gt_tbl %>%
    opt_table_font(font = list(google_font("Source Sans Pro"), default_fonts())) %>%
    tab_options(
      table.font.size = px(13),
      table.border.top.style    = "solid",
      table.border.top.width    = px(2),
      table.border.bottom.style = "solid",
      table.border.bottom.width = px(2),
      heading.border.bottom.style = "solid",
      heading.border.bottom.width = px(1.5),
      column_labels.border.top.style    = "solid",
      column_labels.border.top.width    = px(1.5),
      column_labels.border.bottom.style = "solid",
      column_labels.border.bottom.width = px(1),
      column_labels.font.weight = "bold",
      row_group.font.weight     = "bold",
      heading.align             = "left",
      table.align               = "center",
      source_notes.font.size    = px(11),
      source_notes.padding      = px(6),
      data_row.padding          = px(5)
    ) %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_column_labels()
    )
  if (!is.null(caption_md)) {
    out <- out %>% tab_header(title = md(caption_md))
  }
  out
}
```

```{r run_analysis, include=FALSE, cache=FALSE}
#analysis pipeline

#cohort and outcome
data     <- read.csv(data_path)
baseline <- data %>% filter(PERIOD == 1)
analytic <- baseline %>% filter(PREVSTRK == 0)

days_10yr <- 365.25 * 10

analytic <- analytic %>%
  mutate(
    STROKE_10yr   = ifelse(STROKE == 1 & TIMESTRK <= days_10yr, 1, 0),
    FU_10yr       = pmin(TIMESTRK, days_10yr),
    TIMESTRK_10yr = ifelse(TIMESTRK > days_10yr, days_10yr, TIMESTRK),
    SYSBP_CAT = case_when(
      SYSBP <  160 ~ "<160",
      SYSBP >= 160 ~ ">=160",
      TRUE         ~ NA_character_
    ) %>% factor(levels = c("<160", ">=160"))
  )

analytic_men   <- analytic %>% filter(SEX == 1)
analytic_women <- analytic %>% filter(SEX == 2)

#lasso and cox ph
core_vars      <- c("AGE", "SYSBP", "DIABETES")
candidate_vars <- c("CURSMOKE", "BPMEDS", "PREVCHD", "TOTCHOL", "BMI")
all_model_vars <- c(core_vars, candidate_vars)

cox_men_data <- analytic_men %>%
  select(FU_10yr, STROKE_10yr, all_of(all_model_vars)) %>%
  drop_na()

cox_women_data <- analytic_women %>%
  select(FU_10yr, STROKE_10yr, all_of(all_model_vars)) %>%
  drop_na()

y_men   <- with(cox_men_data,   Surv(FU_10yr, STROKE_10yr))
y_women <- with(cox_women_data, Surv(FU_10yr, STROKE_10yr))

x_men <- model.matrix(
  ~ AGE + SYSBP + DIABETES + CURSMOKE + BPMEDS + PREVCHD + TOTCHOL + BMI,
  data = cox_men_data)[, -1]
x_women <- model.matrix(
  ~ AGE + SYSBP + DIABETES + CURSMOKE + BPMEDS + PREVCHD + TOTCHOL + BMI,
  data = cox_women_data)[, -1]

penalty_men   <- ifelse(colnames(x_men)   %in% core_vars, 0, 1)
penalty_women <- ifelse(colnames(x_women) %in% core_vars, 0, 1)

set.seed(6624)
cv_lasso_men <- cv.glmnet(
  x_men, y_men, family = "cox", alpha = 1,
  penalty.factor = penalty_men, nfolds = 5, standardize = TRUE
)
cv_lasso_women <- cv.glmnet(
  x_women, y_women, family = "cox", alpha = 1,
  penalty.factor = penalty_women, nfolds = 5, standardize = TRUE
)

coef_men_min   <- coef(cv_lasso_men,   s = "lambda.min")
coef_women_min <- coef(cv_lasso_women, s = "lambda.min")

selected_men_min   <- rownames(coef_men_min)[as.vector(coef_men_min != 0)]
selected_women_min <- rownames(coef_women_min)[as.vector(coef_women_min != 0)]

final_vars_union <- sort(unique(c(selected_men_min, selected_women_min)))
final_vars_men   <- final_vars_union
final_vars_women <- final_vars_union

refit_men_data <- analytic_men %>%
  select(FU_10yr, STROKE_10yr, all_of(final_vars_men)) %>% drop_na()
refit_women_data <- analytic_women %>%
  select(FU_10yr, STROKE_10yr, all_of(final_vars_women)) %>% drop_na()

formula_men   <- as.formula(paste("Surv(FU_10yr, STROKE_10yr) ~",
                                  paste(final_vars_men, collapse = " + ")))
formula_women <- as.formula(paste("Surv(FU_10yr, STROKE_10yr) ~",
                                  paste(final_vars_women, collapse = " + ")))

cox_final_men   <- coxph(formula_men,   data = refit_men_data)
cox_final_women <- coxph(formula_women, data = refit_women_data)

#labels for table 3
term_labels <- c(
  "AGE (per 1 year)"       = "Age (per 1 year)",
  "SYSBP (per 10 mmHg)"    = "Systolic BP (per 10 mmHg)",
  "DIABETES"               = "Diabetes",
  "CURSMOKE"               = "Current smoker",
  "BPMEDS"                 = "Antihypertensive medication",
  "PREVCHD"                = "Prevalent CHD",
  "TOTCHOL (per 10 mg/dL)" = "Total cholesterol (per 10 mg/dL)",
  "BMI (per 1 kg/m^2)"     = "BMI (per 1 kg/m\u00B2)"
)

rescale_hr <- function(tidy_df) {
  tidy_df %>%
    mutate(
      scale_factor = case_when(
        term == "SYSBP" ~ 10,
        term == "TOTCHOL" ~ 10,
        TRUE ~ 1
      ),
      estimate  = exp(estimate  * scale_factor),
      conf.low  = exp(conf.low  * scale_factor),
      conf.high = exp(conf.high * scale_factor),
      term = case_when(
        term == "SYSBP"   ~ "SYSBP (per 10 mmHg)",
        term == "TOTCHOL" ~ "TOTCHOL (per 10 mg/dL)",
        term == "AGE"     ~ "AGE (per 1 year)",
        term == "BMI"     ~ "BMI (per 1 kg/m^2)",
        TRUE ~ term
      )
    )
}

cox_final_men_tidy <- tidy(cox_final_men, conf.int = TRUE) %>%
  rescale_hr() %>% mutate(sex = "men")
cox_final_women_tidy <- tidy(cox_final_women, conf.int = TRUE) %>%
  rescale_hr() %>% mutate(sex = "women")

cox_final_results <- bind_rows(cox_final_men_tidy, cox_final_women_tidy) %>%
  select(sex, term, estimate, conf.low, conf.high, p.value)

#cohort flow diagram
n_period1             <- nrow(baseline)
n_prevstroke_excluded <- sum(baseline$PREVSTRK == 1, na.rm = TRUE)
n_analytic            <- nrow(analytic)
n_men                 <- nrow(analytic_men)
n_women               <- nrow(analytic_women)
n_lasso_men           <- nrow(cox_men_data)
n_lasso_women         <- nrow(cox_women_data)
n_refit_men           <- nrow(refit_men_data)
n_refit_women         <- nrow(refit_women_data)

cohort_flow_code <- glue("
digraph cohort_flow {{
  graph [layout = dot, rankdir = TB, bgcolor = 'transparent']
  node [shape = box, style = 'rounded,filled', fontname = 'Source Sans Pro', fontsize = 11,
        fillcolor = '#F4F7FA', color = '#36526B', penwidth = 1.2]
  edge [fontname = 'Source Sans Pro', fontsize = 10, color = '#36526B']

  A [label = 'Period 1 baseline sample\\nN = {n_period1}']
  B [label = 'Excluded: prevalent stroke\\nN = {n_prevstroke_excluded}',
     fillcolor = '#FDECEA', color = '#B23A48']
  C [label = 'Baseline-eligible analytic cohort\\nN = {n_analytic}']

  D [label = 'Men\\nAnalytic sample: N = {n_men}\\nLASSO sample: N = {n_lasso_men}\\nFinal refit sample: N = {n_refit_men}']
  E [label = 'Women\\nAnalytic sample: N = {n_women}\\nLASSO sample: N = {n_lasso_women}\\nFinal refit sample: N = {n_refit_women}']

  A -> C
  A -> B [style = dashed]
  C -> D
  C -> E
}}
")

#TABLE 1: baseline + follow-up by sex

fmt_mean_sd <- function(x, d = 1) {
  x <- x[!is.na(x)]; if (!length(x)) return("\u2014")
  sprintf(paste0("%.", d, "f (%.", d, "f)"), mean(x), sd(x))
}
fmt_n_pct <- function(x) {
  nm <- x[!is.na(x)]; if (!length(nm)) return("\u2014")
  n <- sum(nm == 1); sprintf("%s (%.1f%%)", format(n, big.mark = ","), 100 * n / length(nm))
}
fmt_missing_n <- function(x) {
  nm <- sum(is.na(x)); if (!nm) NA_character_ else format(nm, big.mark = ",")
}
fmt_time_ms <- function(time, ind, v = 1, d = 1) {
  x <- time[ind == v & !is.na(ind) & !is.na(time)]; if (!length(x)) return("\u2014")
  sprintf(paste0("%.", d, "f (%.", d, "f)"), mean(x), sd(x))
}
fmt_km <- function(time, event, t = days_10yr, d = 1) {
  k <- !is.na(time) & !is.na(event); if (!sum(k)) return("\u2014")
  s <- summary(survfit(Surv(time[k], event[k]) ~ 1), times = t, extend = TRUE)$surv
  if (!length(s) || is.na(s)) return("\u2014")
  sprintf(paste0("%.", d, "f%%"), 100 * s)
}

t1 <- analytic %>%
  mutate(
    sex = factor(SEX, c(1, 2), c("Men", "Women")),
    diabetes = DIABETES, smoker = CURSMOKE, bpmed = BPMEDS,
    chd = PREVCHD, stroke10 = STROKE_10yr
  ) %>%
  select(sex, AGE, SYSBP, TOTCHOL, BMI, diabetes, smoker, bpmed, chd, stroke10, FU_10yr)

n_overall_t1 <- nrow(t1); n_men_t1 <- sum(t1$sex == "Men"); n_women_t1 <- sum(t1$sex == "Women")

row_cont <- function(v, lab) tibble(
  Characteristic = lab,
  Overall = fmt_mean_sd(t1[[v]]),
  Men     = fmt_mean_sd(t1 %>% filter(sex == "Men")   %>% pull(all_of(v))),
  Women   = fmt_mean_sd(t1 %>% filter(sex == "Women") %>% pull(all_of(v)))
)
row_miss <- function(v) {
  m <- fmt_missing_n(t1[[v]]); if (is.na(m)) return(NULL)
  tibble(Characteristic = "  Missing, n", Overall = m,
         Men   = fmt_missing_n(t1 %>% filter(sex == "Men")   %>% pull(all_of(v))),
         Women = fmt_missing_n(t1 %>% filter(sex == "Women") %>% pull(all_of(v))))
}
row_bin <- function(v, lab) tibble(
  Characteristic = lab,
  Overall = fmt_n_pct(t1[[v]]),
  Men     = fmt_n_pct(t1 %>% filter(sex == "Men")   %>% pull(all_of(v))),
  Women   = fmt_n_pct(t1 %>% filter(sex == "Women") %>% pull(all_of(v)))
)

table1_body <- bind_rows(
  row_cont("AGE",     "Age (years)"),
  row_cont("SYSBP",   "Systolic BP (mmHg)"),
  row_cont("TOTCHOL", "Total cholesterol (mg/dL)"),
  row_miss("TOTCHOL"),
  row_cont("BMI",     "BMI (kg/m\u00B2)"),
  row_miss("BMI"),
  row_bin("diabetes", "Diabetes mellitus"),
  row_bin("smoker",   "Current smoker"),
  row_bin("bpmed",    "Antihypertensive medication"),
  row_miss("bpmed"),
  row_bin("chd",      "Prevalent CHD"),
  row_bin("stroke10", "10-year stroke event")
) %>% mutate(section = "Baseline characteristics")

followup_body <- bind_rows(
  tibble(Characteristic = "Time to event (days)",
         Overall = fmt_time_ms(t1$FU_10yr, t1$stroke10, 1),
         Men   = fmt_time_ms(t1$FU_10yr[t1$sex=="Men"],   t1$stroke10[t1$sex=="Men"],   1),
         Women = fmt_time_ms(t1$FU_10yr[t1$sex=="Women"], t1$stroke10[t1$sex=="Women"], 1)),
  tibble(Characteristic = "Time to censoring (days)",
         Overall = fmt_time_ms(t1$FU_10yr, t1$stroke10, 0),
         Men   = fmt_time_ms(t1$FU_10yr[t1$sex=="Men"],   t1$stroke10[t1$sex=="Men"],   0),
         Women = fmt_time_ms(t1$FU_10yr[t1$sex=="Women"], t1$stroke10[t1$sex=="Women"], 0)),
  tibble(Characteristic = "10-year stroke-free survival (KM)",
         Overall = fmt_km(t1$FU_10yr, t1$stroke10),
         Men   = fmt_km(t1$FU_10yr[t1$sex=="Men"],   t1$stroke10[t1$sex=="Men"]),
         Women = fmt_km(t1$FU_10yr[t1$sex=="Women"], t1$stroke10[t1$sex=="Women"]))
) %>% mutate(section = "Follow-up and survival")

table1_data <- bind_rows(table1_body, followup_body)

table1_gt <- table1_data %>%
  gt(groupname_col = "section") %>%
  cols_label(
    Characteristic = "Characteristic",
    Overall = md(paste0("**Overall**<br>N = ", format(n_overall_t1, big.mark = ","))),
    Men     = md(paste0("**Men**<br>N = ",     format(n_men_t1,     big.mark = ","))),
    Women   = md(paste0("**Women**<br>N = ",   format(n_women_t1,   big.mark = ",")))
  ) %>%
  cols_align("left", Characteristic) %>%
  cols_align("center", c(Overall, Men, Women)) %>%
  tab_source_note(md("Continuous variables: mean (SD). Binary variables: n (%) using non-missing denominator. Missing counts shown only when > 0. The KM row reports estimated stroke-free survival at 10 years.")) %>%
  style_gt(caption_md = "**Table 1.** Baseline characteristics and 10-year follow-up experience, overall and by sex.")

#TABLE 2: risk factors across examination periods
analytic_ids <- analytic %>% distinct(RANDID)
table2_src   <- data %>% semi_join(analytic_ids, by = "RANDID")

cell_cont <- function(x) {
  if (!sum(!is.na(x))) return("\u2014")
  sprintf("%.1f (%.1f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
}
cell_bin  <- function(x, lv) {
  n <- sum(!is.na(x)); if (!n) return("\u2014")
  k <- sum(x == lv, na.rm = TRUE)
  sprintf("%d (%.1f%%)", k, 100 * k / n)
}
n_nm <- function(x) sum(!is.na(x))
n_lv <- function(x, lv) sum(x == lv, na.rm = TRUE)

build_cont <- function(v, lab) {
  g <- function(p) table2_src[[v]][table2_src$PERIOD == p]
  tibble(Variable = lab,
         p1_n = as.character(n_nm(g(1))), p1_v = cell_cont(g(1)),
         p2_n = as.character(n_nm(g(2))), p2_v = cell_cont(g(2)),
         p3_n = as.character(n_nm(g(3))), p3_v = cell_cont(g(3)),
         section = "Continuous")
}
build_bin <- function(v, lab) {
  g <- function(p) table2_src[[v]][table2_src$PERIOD == p]
  tibble(Variable = lab,
         p1_n = as.character(n_lv(g(1), 1)), p1_v = cell_bin(g(1), 1),
         p2_n = as.character(n_lv(g(2), 1)), p2_v = cell_bin(g(2), 1),
         p3_n = as.character(n_lv(g(3), 1)), p3_v = cell_bin(g(3), 1),
         section = "Binary (Yes)")
}

table2_data <- bind_rows(
  build_cont("AGE",     "Age (years)"),
  build_cont("SYSBP",   "Systolic BP (mmHg)"),
  build_cont("TOTCHOL", "Total cholesterol (mg/dL)"),
  build_cont("BMI",     "BMI (kg/m\u00B2)"),
  build_bin("DIABETES", "Diabetes"),
  build_bin("CURSMOKE", "Current smoker"),
  build_bin("BPMEDS",   "Antihypertensive medication"),
  build_bin("PREVCHD",  "Prevalent CHD")
)

n_p1 <- sum(table2_src$PERIOD == 1, na.rm = TRUE)
n_p2 <- sum(table2_src$PERIOD == 2, na.rm = TRUE)
n_p3 <- sum(table2_src$PERIOD == 3, na.rm = TRUE)

table2_gt <- table2_data %>%
  gt(groupname_col = "section") %>%
  tab_spanner(md(paste0("**Period 1**<br>N = ", format(n_p1, big.mark = ","))), c(p1_n, p1_v)) %>%
  tab_spanner(md(paste0("**Period 2**<br>N = ", format(n_p2, big.mark = ","))), c(p2_n, p2_v)) %>%
  tab_spanner(md(paste0("**Period 3**<br>N = ", format(n_p3, big.mark = ","))), c(p3_n, p3_v)) %>%
  cols_label(Variable = "Variable",
             p1_n = "n", p1_v = "Value",
             p2_n = "n", p2_v = "Value",
             p3_n = "n", p3_v = "Value") %>%
  cols_align("left", Variable) %>%
  cols_align("center", c(p1_n, p1_v, p2_n, p2_v, p3_n, p3_v)) %>%
  tab_source_note(md("Same baseline-eligible cohort as Table 1. Continuous: mean (SD). Binary: n (%) at the Yes level, using the non-missing denominator within period.")) %>%
  style_gt(caption_md = "**Table 2.** Distribution of selected risk factors across examination periods.")

#KM log-rank table

analytic <- analytic %>%
  mutate(
    sex_label      = factor(SEX,      c(1,2), c("Men","Women")),
    diabetes_label = factor(DIABETES, c(0,1), c("No diabetes","Diabetes")),
    smoke_label    = factor(CURSMOKE, c(0,1), c("Not current","Current")),
    prevchd_label  = factor(PREVCHD,  c(0,1), c("No CHD","CHD")),
    bpmeds_label   = factor(BPMEDS,   c(0,1), c("No BP meds","BP meds"))
  )

analytic_men          <- analytic %>% filter(SEX == 1)
analytic_women        <- analytic %>% filter(SEX == 2)
analytic_men_bpmeds   <- analytic_men   %>% filter(!is.na(BPMEDS))
analytic_women_bpmeds <- analytic_women %>% filter(!is.na(BPMEDS))

logrank_summary <- function(formula, data, group_var, stratum = NA_character_) {
  fit <- survdiff(formula, data = data)
  tibble(
    stratum  = stratum,
    variable = group_var,
    chisq    = fit$chisq,
    df       = length(fit$n) - 1,
    p_value  = 1 - pchisq(fit$chisq, df = length(fit$n) - 1),
    n_total  = sum(fit$n)
  )
}

km_logrank_results <- bind_rows(
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ sex_label,      analytic,              "Sex",                         "Overall"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_men,          "Diabetes",                    "Men"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label,    analytic_men,          "Current smoking",             "Men"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label,   analytic_men_bpmeds,   "Antihypertensive med.",       "Men"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label,  analytic_men,          "Prevalent CHD",               "Men"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT,      analytic_men,          "Systolic BP \u2265 160 mmHg", "Men"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ diabetes_label, analytic_women,        "Diabetes",                    "Women"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ smoke_label,    analytic_women,        "Current smoking",             "Women"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ bpmeds_label,   analytic_women_bpmeds, "Antihypertensive med.",       "Women"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ prevchd_label,  analytic_women,        "Prevalent CHD",               "Women"),
  logrank_summary(Surv(FU_10yr, STROKE_10yr) ~ SYSBP_CAT,      analytic_women,        "Systolic BP \u2265 160 mmHg", "Women")
)

logrank_gt <- km_logrank_results %>%
  gt(groupname_col = "stratum") %>%
  cols_label(
    variable = "Comparison",
    chisq    = html("&chi;<sup>2</sup>"),
    df       = "df",
    p_value  = "P-value",
    n_total  = "N"
  ) %>%
  fmt_number(columns = chisq, decimals = 2) %>%
  fmt(columns = p_value,
      fns = function(x) ifelse(x < 0.001, "<0.001", sprintf("%.3f", x))) %>%
  fmt_number(columns = n_total, decimals = 0, use_seps = TRUE) %>%
  cols_align("left",   variable) %>%
  cols_align("center", c(chisq, df, p_value, n_total)) %>%
  tab_source_note(md("Log-rank test of equality of 10-year stroke-free survival between groups, within each stratum.")) %>%
  style_gt(caption_md = "**Table S1.** Log-rank tests for unadjusted 10-year stroke comparisons.")


#TABLE 3: final Cox models

format_hr_ci <- function(hr, lo, hi) sprintf("%.2f (%.2f, %.2f)", hr, lo, hi)
format_p     <- function(p) ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))

table3_data <- cox_final_results %>%
  mutate(
    term_lab = unname(term_labels[term]),
    hr_ci    = format_hr_ci(estimate, conf.low, conf.high),
    p_fmt    = format_p(p.value)
  ) %>%
  select(sex, term_lab, hr_ci, p_fmt) %>%
  pivot_wider(
    names_from  = sex,
    values_from = c(hr_ci, p_fmt),
    names_glue  = "{sex}_{.value}"
  ) %>%
  mutate(term_lab = factor(term_lab, levels = c(
    "Age (per 1 year)", "Systolic BP (per 10 mmHg)", "Diabetes",
    "Current smoker", "Antihypertensive medication",
    "Prevalent CHD", "Total cholesterol (per 10 mg/dL)",
    "BMI (per 1 kg/m\u00B2)"
  ))) %>%
  arrange(term_lab) %>%
  select(term_lab, men_hr_ci, men_p_fmt, women_hr_ci, women_p_fmt)

table3_gt <- table3_data %>%
  gt() %>%
  cols_label(
    term_lab    = "Variable",
    men_hr_ci   = "HR (95% CI)",
    men_p_fmt   = "P-value",
    women_hr_ci = "HR (95% CI)",
    women_p_fmt = "P-value"
  ) %>%
  tab_spanner(label = md("**Men**"),   columns = c(men_hr_ci,   men_p_fmt)) %>%
  tab_spanner(label = md("**Women**"), columns = c(women_hr_ci, women_p_fmt)) %>%
  cols_align("left",   term_lab) %>%
  cols_align("center", c(men_hr_ci, men_p_fmt, women_hr_ci, women_p_fmt)) %>%
  tab_source_note(md("Hazard ratios from sex-specific Cox proportional hazards models using the union of LASSO-selected predictors (\u03BB.min). Age, systolic BP, and diabetes were forced into the model.")) %>%
  style_gt(caption_md = "**Table 3.** Adjusted hazard ratios for 10-year incident stroke, by sex.")


#PH diagnostics

ph_men   <- cox.zph(cox_final_men)
ph_women <- cox.zph(cox_final_women)
ph_results <- bind_rows(
  as.data.frame(ph_men$table)   %>% rownames_to_column("term") %>% mutate(sex = "men"),
  as.data.frame(ph_women$table) %>% rownames_to_column("term") %>% mutate(sex = "women")
) %>% select(sex, term, chisq, df, p) %>% rename(p_value = p)

#Profile predictions + Table 4

sysbp_normal_men <- refit_men_data %>%
  filter(DIABETES == 0, CURSMOKE == 0) %>% summarise(m = mean(SYSBP)) %>% pull(m)
sysbp_normal_women <- refit_women_data %>%
  filter(DIABETES == 0, CURSMOKE == 0) %>% summarise(m = mean(SYSBP)) %>% pull(m)

build_profiles <- function(sbp_norm) {
  ages <- c(40, 50, 60)
  defs <- tribble(
    ~profile, ~profile_label,                    ~SYSBP,   ~DIABETES, ~CURSMOKE,
    "P1",     "No major risk factors",           sbp_norm, 0,         0,
    "P2",     "High SBP only",                   160,      0,         0,
    "P3",     "Diabetes only",                   sbp_norm, 1,         0,
    "P4",     "High SBP + diabetes",             160,      1,         0,
    "P5",     "High SBP + diabetes + smoking",   160,      1,         1
  )
  expand_grid(AGE = ages, defs) %>%
    mutate(BPMEDS = 0) %>%
    select(profile, profile_label, AGE, SYSBP, DIABETES, CURSMOKE, BPMEDS)
}

profiles_men   <- build_profiles(sysbp_normal_men)
profiles_women <- build_profiles(sysbp_normal_women)

predict_10yr_stroke <- function(fit, prof) {
  sf <- survfit(fit, newdata = prof)
  s  <- summary(sf, times = days_10yr)
  tibble(surv = as.numeric(s$surv),
         surv_lower = as.numeric(s$lower),
         surv_upper = as.numeric(s$upper)) %>%
    mutate(stroke_prob       = 1 - surv,
           stroke_prob_lower = 1 - surv_upper,
           stroke_prob_upper = 1 - surv_lower) %>%
    bind_cols(prof, .)
}

men_predictions   <- predict_10yr_stroke(cox_final_men,   profiles_men)
women_predictions <- predict_10yr_stroke(cox_final_women, profiles_women)

all_predictions <- bind_rows(
  men_predictions   %>% mutate(sex = "men"),
  women_predictions %>% mutate(sex = "women")
) %>%
  mutate(
    stroke_prob_pct       = round(stroke_prob       * 100, 2),
    stroke_prob_lower_pct = round(stroke_prob_lower * 100, 2),
    stroke_prob_upper_pct = round(stroke_prob_upper * 100, 2)
  ) %>%
  select(sex, profile, profile_label, AGE, SYSBP, DIABETES, CURSMOKE, BPMEDS,
         stroke_prob_pct, stroke_prob_lower_pct, stroke_prob_upper_pct)

format_pct_ci <- function(est, lo, hi) sprintf("%.1f%% (%.1f, %.1f)", est, lo, hi)

profile_info <- profiles_men %>%
  distinct(profile, profile_label, DIABETES, CURSMOKE, BPMEDS, SYSBP) %>%
  rename(men_sbp = SYSBP) %>%
  left_join(profiles_women %>% distinct(profile, SYSBP) %>% rename(women_sbp = SYSBP),
            by = "profile") %>%
  mutate(
    Diabetes = ifelse(DIABETES == 1, "Yes", "No"),
    Smoking  = ifelse(CURSMOKE == 1, "Yes", "No"),
    bp_med   = ifelse(BPMEDS   == 1, "Yes", "No"),
    men_sbp   = sprintf("%.0f", men_sbp),
    women_sbp = sprintf("%.0f", women_sbp)
  ) %>%
  select(profile, profile_label, Diabetes, Smoking, bp_med, men_sbp, women_sbp)

predictions_wide <- all_predictions %>%
  mutate(
    sex  = ifelse(sex == "men", "men", "women"),
    risk = format_pct_ci(stroke_prob_pct, stroke_prob_lower_pct, stroke_prob_upper_pct),
    col  = paste0(sex, "_age", AGE)
  ) %>%
  select(profile, col, risk) %>%
  pivot_wider(names_from = col, values_from = risk)

table4_data <- profile_info %>%
  left_join(predictions_wide, by = "profile") %>%
  select(profile, profile_label, Diabetes, Smoking, bp_med,
         men_sbp,   men_age40,   men_age50,   men_age60,
         women_sbp, women_age40, women_age50, women_age60)

table4_gt <- table4_data %>%
  gt() %>%
  cols_label(
    profile       = "Profile",
    profile_label = "Description",
    Diabetes      = "Diabetes",
    Smoking       = "Smoking",
    bp_med        = "BP med",
    men_sbp       = "SBP",
    men_age40     = "Age 40",
    men_age50     = "Age 50",
    men_age60     = "Age 60",
    women_sbp     = "SBP",
    women_age40   = "Age 40",
    women_age50   = "Age 50",
    women_age60   = "Age 60"
  ) %>%
  tab_spanner(label = md("**Profile definition**"),
              columns = c(profile_label, Diabetes, Smoking, bp_med)) %>%
  tab_spanner(label = md("**Men — predicted 10-year stroke probability**"),
              columns = c(men_sbp, men_age40, men_age50, men_age60)) %>%
  tab_spanner(label = md("**Women — predicted 10-year stroke probability**"),
              columns = c(women_sbp, women_age40, women_age50, women_age60)) %>%
  cols_align("left",   c(profile, profile_label)) %>%
  cols_align("center", c(Diabetes, Smoking, bp_med,
                         men_sbp,   men_age40,   men_age50,   men_age60,
                         women_sbp, women_age40, women_age50, women_age60)) %>%
  tab_source_note(md("Predicted risks are percent (95% CI) from the final sex-specific Cox models. Antihypertensive medication fixed at No for all profiles. SBP shown in mmHg; 'normal' SBP values use the sex-specific mean among model-sample participants without diabetes who did not currently smoke.")) %>%
  style_gt(caption_md = "**Table 4.** Predicted 10-year stroke probability by sex, age, and prespecified baseline risk profile.")


#Figure 2 data

figure2_data <- all_predictions %>%
  mutate(
    sex     = factor(ifelse(sex == "men", "Men", "Women"), levels = c("Men", "Women")),
    profile = factor(profile,
                     levels = c("P1", "P2", "P3", "P4", "P5"),
                     labels = c("P1: Low risk",
                                "P2: High SBP",
                                "P3: Diabetes",
                                "P4: SBP + DM",
                                "P5: SBP + DM + Smoke"))
  )


#counts used inline

N_baseline          <- nrow(baseline)
N_excluded_prevstrk <- N_baseline - nrow(analytic)
N_analytic          <- nrow(analytic)
N_men               <- nrow(analytic_men)
N_women             <- nrow(analytic_women)
N_modeling_men      <- nrow(refit_men_data)
N_modeling_women    <- nrow(refit_women_data)
n_events_men        <- sum(refit_men_data$STROKE_10yr)
n_events_women      <- sum(refit_women_data$STROKE_10yr)

get_hr_row <- function(df, sex_in, term_in) df %>% filter(sex == sex_in, term == term_in) %>% slice(1)
```

# Introduction

The dataset provided for this project was a longitudinal subset of the Framingham Heart Study containing clinical examination, laboratory, questionnaire, and event data on 4,434 participants observed in up to three examination periods and followed for cardiovascular outcomes. The investigator's goal was to identify baseline risk factors for incident stroke and to estimate 10-year stroke probability for clinically meaningful baseline risk profiles. The investigator specified that analyses should use only Period 1 covariates, restrict follow-up to the first 10 years, exclude participants with prevalent stroke at baseline, treat death as censoring rather than as a competing risk, and be conducted separately in men and women.

The primary scientific question was which baseline factors were associated with 10-year incident stroke within each sex. The primary predictors of interest were age, systolic blood pressure, and diabetes. Additional candidate predictors were current smoking, antihypertensive medication use, prevalent coronary heart disease, total cholesterol, and body mass index. A secondary objective was to translate the final sex-specific models into 10-year stroke probabilities for five prespecified profiles evaluated at ages 40, 50, and 60.

The main statistical hypotheses were that, within each sex, older age, higher systolic blood pressure, and diabetes were associated with higher hazard of incident stroke during the first 10 years after baseline. In the final Cox proportional hazards models, these hypotheses were tested as two-sided null hypotheses that the corresponding regression coefficients were zero. Additional candidate predictors were assessed through penalized variable selection and retained if they contributed predictive information beyond the three core variables. The profile-prediction objective was descriptive and focused on estimated 10-year stroke probability and 95% confidence intervals rather than formal hypothesis testing.

# Methods

## Data source and cohort definition

The Framingham teaching dataset is stored in longitudinal format, with one to three examination records per participant and event times recorded from the baseline examination. Variables used in this analysis included sex, examination period, age, systolic blood pressure, total cholesterol, body mass index, diabetes, current smoking, antihypertensive medication use, prevalent coronary heart disease, prevalent stroke, incident stroke, time to stroke, and time to death. Period 1 records were used to define baseline. Participants with prevalent stroke at baseline were excluded so that only incident stroke events were analyzed. No additional participants were excluded at cohort construction.

## Outcome definition and follow-up

The primary outcome was time to first incident stroke during the first 10 years after baseline. A 10-year event indicator was defined using the Framingham stroke indicator and time-to-stroke variable. Follow-up was administratively censored at 10 years using 3,652.5 days. Participants with stroke on or before that time were treated as events at the observed stroke time. Participants who remained stroke-free through 10 years, including those with stroke recorded after 10 years, were administratively censored at 10 years. Participants who died or were otherwise lost to follow-up before 10 years without prior stroke were censored at that earlier time. Death was treated as right-censoring, consistent with the project instructions.

## Data management and descriptive analyses

Continuous baseline variables considered for modeling were age, systolic blood pressure, total cholesterol, and body mass index. Binary baseline variables were diabetes, current smoking, antihypertensive medication use, and prevalent coronary heart disease. Continuous variables were screened for implausible values, and binary variables were checked to confirm valid coding. Extreme systolic blood pressure and total cholesterol values were flagged for review in the code and retained. Baseline missingness was summarized for all analysis variables. Missingness was limited to antihypertensive medication use, total cholesterol, and body mass index; age, systolic blood pressure, diabetes, current smoking, and prevalent coronary heart disease were complete at baseline.

Preliminary descriptive analyses were used to summarize the analytic cohort, evaluate the amount of covariate change across examination periods, and assess potential collinearity and unadjusted associations. Table 1 summarized baseline characteristics and 10-year follow-up by sex, including missingness for incomplete baseline variables, the number of stroke events, mean time to event among those with stroke, mean time to censoring among those without stroke, and estimated 10-year stroke-free survival from Kaplan–Meier methods. Table 2 summarized selected risk factors across Periods 1, 2, and 3 within the same baseline-eligible cohort to assess whether a future time-varying covariate analysis might be informative. Additional exploratory summaries examined distributions of continuous variables, the relationship between systolic blood pressure and antihypertensive medication use, correlations among continuous predictors, and bivariate associations between candidate baseline variables and 10-year stroke status.

## Unadjusted survival analysis

Kaplan–Meier methods were used to summarize unadjusted 10-year stroke-free survival. Curves were examined overall by sex and, within each sex, across diabetes status, current smoking, antihypertensive medication use, prevalent coronary heart disease, and categorized systolic blood pressure. Systolic blood pressure was categorized as less than 160 mmHg versus at least 160 mmHg to align with the prespecified profile definitions. Because 10-year stroke incidence was low, Kaplan–Meier curves were displayed as cumulative stroke probability, defined as 1 minus the estimated survival function. Log-rank tests were computed for each unadjusted comparison and were used descriptively rather than as the primary inferential analysis.

## Cox proportional hazards modeling and variable selection

The primary adjusted analyses used sex-specific Cox proportional hazards regression for time to incident stroke during the first 10 years of follow-up. Age, systolic blood pressure, and diabetes were retained in all models a priori. Candidate additional predictors were current smoking, antihypertensive medication use, prevalent coronary heart disease, total cholesterol, and body mass index.

The final analysis used penalized Cox regression with LASSO rather than the earlier backward-selection plan. Within each sex, a Cox model with LASSO penalty was fit using 5-fold cross-validation and `lambda.min`. Penalty factors were set to 0 for age, systolic blood pressure, and diabetes so that these three variables were forced into the model, and to 1 for the five candidate predictors. The predictors selected at `lambda.min` were identified separately in men and women. To support comparability of the final sex-specific models, the union of predictors selected in either sex was then used as the final covariate set for both sexes.

After variable selection, standard unpenalized Cox proportional hazards models were refit separately in men and women using complete cases for the retained final-model variables only. This refit step allowed observations excluded in the initial LASSO step because of missingness in nonselected candidate variables to be recovered. Hazard ratios, 95% confidence intervals, and two-sided Wald p-values were obtained from the refit Cox models. Systolic blood pressure was reported per 10 mmHg increase; age was reported per 1-year increase; total cholesterol, if retained, was reported per 10 mg/dL; and body mass index, if retained, was reported per 1 kg/m². Statistical significance was assessed at $\alpha = 0.05$. No multiple-comparison adjustment was applied because the primary inferential focus was the final sex-specific Cox models and profile-based predictions, whereas the Kaplan–Meier and other exploratory analyses were descriptive.

## Proportional hazards assessment

The proportional hazards assumption was assessed for the final sex-specific Cox models using both formal and graphical methods. Schoenfeld residual tests were computed for each model coefficient and globally within each model. Scaled Schoenfeld residual plots were examined for retained predictors. Log-minus-log survival plots were also reviewed for binary predictors in the final models.

## Profile-specific 10-year stroke probability estimation

The final sex-specific Cox models were used to estimate 10-year stroke probability for five prespecified baseline profiles at ages 40, 50, and 60. The profiles were: no major risk factors; high systolic blood pressure only; diabetes only; high systolic blood pressure plus diabetes; and high systolic blood pressure plus diabetes plus current smoking. For profiles with elevated blood pressure, systolic blood pressure was fixed at 160 mmHg. Antihypertensive medication use was fixed at 0 for all profiles. For the lower-risk profiles, systolic blood pressure was set to the sex-specific mean among model-sample participants without diabetes who were not current smokers. Predicted 10-year stroke probability and 95% confidence intervals were obtained from the sex-specific Cox survival estimates evaluated at 10 years.

# Results

## Cohort summary and descriptive analyses

Figure 1 summarizes construction of the analytic cohort and the sex-specific samples used for survival modeling. The Period 1 baseline cohort contained `r N_baseline` observations. After exclusion of `r N_excluded_prevstrk` participants with prevalent stroke, the analytic cohort contained `r N_analytic` participants, including `r N_men` men and `r N_women` women. The final refit Cox model samples contained `r N_modeling_men` men and `r N_modeling_women` women, with `r n_events_men` and `r n_events_women` 10-year stroke events, respectively.

<p style="text-align:center; font-weight:bold; margin-top:1.5em; margin-bottom:0.4em;">
Figure 1. Cohort flow diagram.
</p>

```{r cohort_flow, out.width='80%'}
grViz(cohort_flow_code)
```

<p style="text-align:center; font-size:0.9em; color:#444; max-width:720px; margin:0.4em auto 1.5em auto;">
Cohort flow for construction of the analytic cohort and sex-specific modeling samples.
</p>

Table 1 summarizes baseline characteristics and 10-year follow-up experience by sex in the analytic cohort.

```{r table1}
table1_gt
```

Across the first 10 years of follow-up, stroke incidence was low in both sexes. Missing baseline data were limited to antihypertensive medication use, total cholesterol, and body mass index. Table 2 summarizes selected risk factors across examination periods within the same baseline-eligible cohort. Systolic blood pressure and diabetes prevalence increased across periods, whereas current smoking declined, supporting the investigator's interest in whether a future time-varying covariate analysis might yield different results.

```{r table2}
table2_gt
```

## Unadjusted survival patterns

Kaplan–Meier methods were used to summarize unadjusted 10-year cumulative stroke incidence overall by sex and, within each sex, by diabetes, current smoking, antihypertensive medication use, prevalent coronary heart disease, and categorized systolic blood pressure. Because 10-year stroke incidence was low, curves were displayed as cumulative incidence, defined as 1 minus the Kaplan–Meier stroke-free survival estimate.

The overall log-rank test comparing men and women did not suggest a large unadjusted difference in 10-year stroke incidence. Within both sexes, the clearest unadjusted separation was seen for categorized systolic blood pressure and diabetes. Participants with baseline systolic blood pressure of at least 160 mmHg had higher cumulative stroke incidence than those with lower systolic blood pressure, and participants with diabetes also showed higher cumulative incidence than those without diabetes. In contrast, curve separation for current smoking, antihypertensive medication use, and prevalent coronary heart disease was less consistent and should be interpreted cautiously given the small number of stroke events.

Table S1 summarizes the log-rank tests for all prespecified unadjusted comparisons. These Kaplan–Meier analyses were used descriptively; the primary adjusted inference came from the sex-specific Cox proportional hazards models.

```{r km_logrank_table}
logrank_gt
```

## Cox proportional hazards model results

The final Cox models used the union of LASSO-selected predictors across the two sex-specific models. The final covariate set was `r paste(final_vars_men, collapse = ", ")`. Table 3 reports adjusted hazard ratios, 95% confidence intervals, and p-values from the refit sex-specific Cox models.

In men, age, systolic blood pressure, and diabetes were the strongest predictors of 10-year incident stroke. Each additional year of age was associated with an adjusted hazard ratio of `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "AGE (per 1 year)")$estimate)` (95% CI `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "AGE (per 1 year)")$conf.low)` to `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "AGE (per 1 year)")$conf.high)`). Each 10 mmHg increase in systolic blood pressure was associated with an adjusted hazard ratio of `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "SYSBP (per 10 mmHg)")$estimate)` (95% CI `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "SYSBP (per 10 mmHg)")$conf.low)` to `r sprintf("%.2f", get_hr_row(cox_final_results, "men", "SYSBP (per 10 mmHg)")$conf.high)`). Baseline diabetes was also associated with substantially higher adjusted hazard in men.

In women, age and systolic blood pressure showed similarly elevated adjusted hazards, whereas the diabetes effect was attenuated and imprecise relative to the men's model. Current smoking and antihypertensive medication use had wide confidence intervals in both sexes.

```{r table3}
table3_gt
```

## Proportional hazards diagnostics

Schoenfeld residual tests did not provide strong evidence against the proportional hazards assumption in either sex-specific final model. The global Schoenfeld test p-values were `r sprintf("%.3f", ph_results$p_value[ph_results$sex == "men" & ph_results$term == "GLOBAL"])` in men and `r sprintf("%.3f", ph_results$p_value[ph_results$sex == "women" & ph_results$term == "GLOBAL"])` in women. Visual review of the residual plots and log-minus-log plots suggested that any departures from proportional hazards were modest and concentrated in sparsely populated exposure groups late in follow-up.

## Predicted 10-year stroke probability by risk profile

Table 4 and Figure 2 summarize predicted 10-year stroke probability for the five prespecified profiles at ages 40, 50, and 60, separately in men and women. Predicted risk increased with age within every profile and increased across progressively higher-risk profiles within each sex. At age 60, predicted 10-year stroke probability in men ranged from `r sprintf("%.1f%%", all_predictions %>% filter(sex == "men", profile == "P1", AGE == 60) %>% pull(stroke_prob_pct))` for the no-major-risk-factor profile to `r sprintf("%.1f%%", all_predictions %>% filter(sex == "men", profile == "P5", AGE == 60) %>% pull(stroke_prob_pct))` for the profile with elevated systolic blood pressure, diabetes, and current smoking. The corresponding range in women was `r sprintf("%.1f%%", all_predictions %>% filter(sex == "women", profile == "P1", AGE == 60) %>% pull(stroke_prob_pct))` to `r sprintf("%.1f%%", all_predictions %>% filter(sex == "women", profile == "P5", AGE == 60) %>% pull(stroke_prob_pct))`.

```{r table4}
table4_gt
```

<p style="text-align:center; font-weight:bold; margin-top:1.5em; margin-bottom:0.4em;">
Figure 2. Predicted 10-year stroke probability by age, sex, and baseline risk profile.
</p>

```{r figure2, fig.width=8.5, fig.height=4.8, out.width='95%'}
profile_colors <- c(
  "P1: Low risk"          = "#117733",
  "P2: High SBP"          = "#88CCEE",
  "P3: Diabetes"          = "#DDCC77",
  "P4: SBP + DM"          = "#CC6677",
  "P5: SBP + DM + Smoke"  = "#882255"
)

ggplot(figure2_data,
       aes(x = AGE, y = stroke_prob_pct,
           group = profile, color = profile, shape = profile)) +
  geom_ribbon(aes(ymin = stroke_prob_lower_pct, ymax = stroke_prob_upper_pct,
                  fill = profile),
              alpha = 0.10, color = NA, show.legend = FALSE) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.6, fill = "white", stroke = 0.9) +
  facet_wrap(~ sex) +
  scale_x_continuous(breaks = c(40, 50, 60)) +
  scale_y_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0.02, 0.08))) +
  scale_color_manual(values = profile_colors) +
  scale_fill_manual(values  = profile_colors) +
  scale_shape_manual(values = c(16, 17, 15, 18, 21)) +
  labs(
    x = "Age (years)",
    y = "Predicted 10-year stroke probability",
    color = "Risk profile", shape = "Risk profile"
  ) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE),
         shape = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_minimal(base_family = "sans", base_size = 12) +
  theme(
    legend.position    = "bottom",
    legend.title       = element_text(face = "bold"),
    legend.key.width   = unit(1.2, "cm"),
    strip.text         = element_text(face = "bold", size = 12),
    strip.background   = element_rect(fill = "#F4F7FA", color = NA),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.spacing      = unit(1.5, "lines"),
    plot.margin        = margin(10, 10, 10, 10),
    axis.title.x       = element_text(margin = margin(t = 8)),
    axis.title.y       = element_text(margin = margin(r = 8))
  )
```

<p style="text-align:center; font-size:0.9em; color:#444; max-width:760px; margin:0.2em auto 1.5em auto;">
Predicted 10-year stroke probability by age, sex, and prespecified baseline risk profile, from the final sex-specific Cox proportional hazards models. Shaded bands indicate 95% confidence intervals.
</p>

# Discussion

This sex-stratified analysis identified age and systolic blood pressure as the most consistent baseline predictors of 10-year incident stroke in both sexes. Diabetes was also strongly associated with stroke risk in men and contributed materially to the male risk-profile predictions, whereas its adjusted association was weaker and less precise in women. These findings were reflected directly in the predicted 10-year stroke probabilities, which increased with age and with accumulation of baseline risk factors in both sexes.

Several limitations should be considered. First, the number of 10-year stroke events was small relative to the number of candidate predictors, which limited precision and likely contributed to instability in candidate-variable selection. Second, the main analysis used baseline covariate values only, although Table 2 showed that systolic blood pressure and diabetes prevalence changed across follow-up; a time-varying covariate analysis could therefore produce different effect estimates, particularly if baseline measurements became less representative of later exposure status. Third, death was treated as censoring rather than as a competing risk, as specified by the investigator. This choice is appropriate for a cause-specific Cox analysis but may overstate absolute stroke probability in participants at high competing mortality risk. Fourth, antihypertensive medication use in this observational setting should be interpreted as a marker of treated hypertension rather than as a randomized intervention.

Despite these limitations, the analysis addressed the investigator's primary questions. The Cox models identified clinically interpretable baseline predictors of incident stroke within each sex, and the profile-based predictions translated those model results into estimated 10-year stroke probabilities that are directly interpretable for common baseline risk scenarios.


# Reproducible Research Information

- **GitHub Repository**: `https://github.com/emkokesh/BIOS6624`

- **Raw dataset**: `C:/Users/kokeshem/Spring 2026/Advanced Statistical Methods/Project 3/frmgham2.csv`

- **Primary analysis script**: `Code/Primary_Analysis_Project3.Rmd`

- **Report source**: `Reports/Project3_Report.Rmd`